---
title: "05-WGCNA-edgeR"
author: "Zach Bengtsson"
date: "2023-07-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## WGCNA with edgeR normalization

```{r}
# Load necessary packages
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("edgeR")
BiocManager::install("GO.db")
BiocManager::install("BiocGenerics")
BiocManager::install("WGCNA")

install.packages("fastcluster")
conflicts(detail = TRUE)
```
```{r}
library(edgeR)
library(GO.db)
library(fastcluster)
library(BiocGenerics)
library(WGCNA)
library(tidyverse)
```


```{r}
# Assume 'counts' is a matrix where rows are genes, and columns are samples
# Assume 'samples' is a data frame where rows are samples, and columns are sample traits

# Create samples data frame
samples <- data.frame(
  sample = c("S12M", "S13M", "S16F", "S19F", "S22F", "S23M", "S29F", "S31M", "S35F", "S36F", "S39F", "S3F", "S41F", "S44F", "S48M", "S50F", "S52F", "S53F", "S54F", "S59M", "S64M", "S6M", "S76F", "S77F", "S7M", "S9M"),
  treatment = c("Exposed", "Control", "Control", "Control", "Exposed", "Exposed", "Exposed", "Exposed", "Exposed", "Exposed", "Control", "Exposed", "Exposed", "Control", "Exposed", "Exposed", "Control", "Control", "Control", "Exposed", "Control", "Control", "Control", "Exposed", "Control", "Exposed"),
  sex = c("male", "male", "female", "female", "female", "male", "female", "male", "female", "female", "female", "female", "female", "female", "male", "female", "female", "female", "female", "male", "male", "male", "female", "female", "male", "male")
)

# Import count matrix
counts <- read.table("~/github/oyster-lnc/output/02-kallisto-merge/merged_counts.txt", header = TRUE, row.names = 1, check.names = FALSE)
```
```{r}
print(samples)
```

```{r}
# Convert counts to DGEList object
dge <- DGEList(counts = counts)

# Normalization using TMM method
dge <- calcNormFactors(dge, method = "TMM")

# Compute log2 CPM values, with prior.count=1
log2_cpm <- cpm(dge, log = TRUE, prior.count = 1)

# Filter out lowly expressed genes
keep <- rowSums(cpm(dge) > 1) >= 2
log2_cpm <- log2_cpm[keep,]

# Prepare the trait data
trait_data <- as.data.frame(t(samples))

# Run WGCNA analysis
# Allow multi-threading
enableWGCNAThreads()

# Choose a soft-thresholding power
powers <- c(1:10)
sft <- pickSoftThreshold(t(log2_cpm), powerVector = powers, verbose = 5)
# Choose a power based on the results of this analysis

# Run one-step network construction and module detection
net = blockwiseModules(t(log2_cpm), power = 6, # Replace 6 with your chosen power
                       TOMType = "unsigned", minModuleSize = 30,
                       reassignThreshold = 0, mergeCutHeight = 0.25,
                       numericLabels = TRUE, pamRespectsDendro = FALSE,
                       saveTOMs = TRUE, saveTOMFileBase = "TOM",
                       verbose = 3)

# Convert labels to colors for easier plotting
mergedColors = labels2colors(net$colors)

# Plot the dendrogram
plotDendroAndColors(net$dendrograms[[1]], mergedColors[net$blockGenes[[1]]],
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)

```

```{r}
# Module Eigengenes
MEs <- moduleEigengenes(t(log2_cpm), colors = mergedColors)$eigengenes
```

```{r}
# Module-Trait Relationships
MEs_trait_correlation <- cor(MEs, t(trait_data))
MEs_trait_pvalue <- corPvalueStudent(as.matrix(MEs_trait_correlation), ncol(trait_data))

# Gene Significance and Module Membership
GS <- apply(log2_cpm, 1, function(row) cor(row, t(trait_data)[,1], use = "p"))  # for the first trait
MM <- moduleEigengenes(t(log2_cpm), mergedColors)$eigengenes  # Module Membership
```


```{r}
# Count the number of genes in each module
moduleSizes <- table(net$colors)

# Plot the module sizes
barplot(moduleSizes, main = "Number of genes in each module",
        xlab = "Module colors", ylab = "Number of genes",
        col = names(moduleSizes))
```

# Subsetting for males and females to complete 2 separate WGCNAs

## Subset dataset

```{r}
# Subset for males
male_samples <- samples[samples$sex == "male", ]
male_log2_cpm <- log2_cpm[, colnames(log2_cpm) %in% male_samples$sample]

# Subset for females
female_samples <- samples[samples$sex == "female", ]
female_log2_cpm <- log2_cpm[, colnames(log2_cpm) %in% female_samples$sample]

```

## Male Runs

## Unsigned

```{r}
# Subset the data for males
male_samples <- samples[samples$sex == "male", ]
male_counts <- counts[, male_samples$sample]

# Convert male counts to DGEList object
dge_male <- DGEList(counts = male_counts)

# Normalization using TMM method
dge_male <- calcNormFactors(dge_male, method = "TMM")

# Compute log2 CPM values, with prior.count=1
log2_cpm_male <- cpm(dge_male, log = TRUE, prior.count = 1)

# Filter out lowly expressed genes
keep_male <- rowSums(cpm(dge_male) > 1) >= 2
log2_cpm_male <- log2_cpm_male[keep_male,]

# Prepare the trait data for males
trait_data_male <- as.data.frame(t(male_samples))

# Run WGCNA analysis for males
enableWGCNAThreads()
```

```{r}
powers <- c(1:10)
pickSoftThreshold(t(log2_cpm_male), powerVector = powers, verbose = 5)
```

## M - Unsigned Figures

Pick soft threshold identified power of 7 as the best choice to achieve a scale-free topology for your male dataset...
```{r}
# Module detection for males
net_male = blockwiseModules(t(log2_cpm_male), power = 7,
                       TOMType = "unsigned", minModuleSize = 30,
                       reassignThreshold = 0, mergeCutHeight = 0.25,
                       numericLabels = TRUE, pamRespectsDendro = FALSE,
                       saveTOMs = TRUE, saveTOMFileBase = "TOM_male",
                       verbose = 3)

# Convert labels to colors for easier plotting for males
mergedColors_male = labels2colors(net_male$colors)

# Plotting for males
plotDendroAndColors(net_male$dendrograms[[1]], mergedColors_male[net_male$blockGenes[[1]]],
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)

# Count the number of genes in each male module
moduleSizes_male <- table(net_male$colors)

# Plot the male module sizes
barplot(moduleSizes_male, main = "Number of genes in each male module",
        xlab = "Module colors", ylab = "Number of genes",
        col = names(moduleSizes_male))

```

```{r}
# Assuming 'net_male' is your WGCNA network object for males and 'log2_cpm_male' is your original expression matrix

# 1. Module Eigengenes
MEs_male = net_male$MEs

# 2. Module Membership
moduleColors_male = net_male$colors  # these are the module assignments for each gene
membershipMatrix_male = as.data.frame(cor(t(log2_cpm_male), MEs_male, method = "pearson"))  # Compute module membership for each gene
colnames(membershipMatrix_male) = paste0("ME", names(MEs_male))  # This will name columns like MEblue, MEbrown, etc.

# Let's say you're interested in the blue module
moduleColor = "1"  # Replace with the numeric label of the module you're interested in

# You can subset these to get information on a specific module
specificME_male = MEs_male[, moduleColor]
specificMM_male = membershipMatrix_male[, paste0("ME", moduleColor)]

# 3. Original Expression values
moduleGenes_male = names(moduleColors_male[moduleColors_male == moduleColor])
moduleExpression_male = t(log2_cpm_male)[moduleGenes_male, ]

# 4. Heatmaps (You may want to standardize or z-score the expression values before plotting)
heatmap(moduleExpression_male, Rowv=NA, scale="row", margins=c(5,10))  # This is a basic heatmap; you may want to adjust parameters for better visualization

```

```{r}
colnames(MEs_male)

```

```{r}
print(moduleColors_male)
```


```{r}
# Assuming 'net_male' is your WGCNA network object for males and 'log2_cpm_male' is your original expression matrix

# 1. Module Eigengenes
MEs_male = net_male$MEs

# 2. Module Membership
moduleColors_male = net_male$colors  # these are the module assignments for each gene
membershipMatrix_male = as.data.frame(cor(t(log2_cpm_male), MEs_male, method = "pearson"))  # Compute module membership for each gene
colnames(membershipMatrix_male) = paste0("ME", names(MEs_male))  # This will name columns like MEblue, MEbrown, etc.

# Let's say you're interested in the blue module
moduleColor = "1"  # Replace with the numeric label of the module you're interested in

# You can subset these to get information on a specific module
specificME_male = MEs_male[, moduleColor]
specificMM_male = membershipMatrix_male[, paste0("ME", moduleColor)]

# 3. Original Expression values
moduleGenes_male = names(moduleColors_male[moduleColors_male == moduleColor])
moduleExpression_male = t(log2_cpm_male)[moduleGenes_male, ]

# 4. Heatmaps (You may want to standardize or z-score the expression values before plotting)
heatmap(moduleExpression_male, Rowv=NA, scale="row", margins=c(5,10))  # This is a basic heatmap; you may want to adjust parameters for better visualization


```


```{r}
# Convert moduleColors to a tibble
moduleColors_tibble <- as_tibble(moduleColors_male, rownames = "Gene")

# Filter genes that belong to a specific module
moduleGenes <- moduleColors_tibble %>% 
                  filter(value == "value") %>% 
                  pull(Gene)

# Get the expression data for these genes
moduleExpression_tibble <- geneData_tibble %>% 
                              filter(Gene %in% moduleGenes)
```



## Signed

## M - Signed Figures

```{r}
# Module detection for males
net_male = blockwiseModules(t(log2_cpm_male), power = 7,
                       TOMType = "signed", minModuleSize = 30,
                       reassignThreshold = 0, mergeCutHeight = 0.25,
                       numericLabels = TRUE, pamRespectsDendro = FALSE,
                       saveTOMs = TRUE, saveTOMFileBase = "TOM_male",
                       verbose = 3)

# Convert labels to colors for easier plotting for males
mergedColors_male = labels2colors(net_male$colors)

# Plotting for males
plotDendroAndColors(net_male$dendrograms[[1]], mergedColors_male[net_male$blockGenes[[1]]],
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)

# Count the number of genes in each male module
moduleSizes_male <- table(net_male$colors)

# Plot the male module sizes
barplot(moduleSizes_male, main = "Number of genes in each male module",
        xlab = "Module colors", ylab = "Number of genes",
        col = names(moduleSizes_male))

```

## Female Runs

## Unsigned

```{r}
# Subset the data for females
female_samples <- samples[samples$sex == "female", ]
female_counts <- counts[, female_samples$sample]

# Convert female counts to DGEList object
dge_female <- DGEList(counts = female_counts)

# Normalization using TMM method
dge_female <- calcNormFactors(dge_female, method = "TMM")

# Compute log2 CPM values, with prior.count=1
log2_cpm_female <- cpm(dge_female, log = TRUE, prior.count = 1)

# Filter out lowly expressed genes
keep_female <- rowSums(cpm(dge_female) > 1) >= 2
log2_cpm_female <- log2_cpm_female[keep_female,]

# Prepare the trait data for females
trait_data_female <- as.data.frame(t(female_samples))

# Run WGCNA analysis for females
enableWGCNAThreads()
```

Select power to achieve scale-free topology...
```{r}
powers <- c(1:10)
pickSoftThreshold(t(log2_cpm_female), powerVector = powers, verbose = 5)
```

## F - Unsigned Figures

Power of 3 identified by pickSoftThreshold...
```{r}
# Module detection for females
net_female = blockwiseModules(t(log2_cpm_female), power = 3,
                       TOMType = "unsigned", minModuleSize = 30,
                       reassignThreshold = 0, mergeCutHeight = 0.25,
                       numericLabels = TRUE, pamRespectsDendro = FALSE,
                       saveTOMs = TRUE, saveTOMFileBase = "TOM_female",
                       verbose = 3)

# Convert labels to colors for easier plotting for females
mergedColors_female = labels2colors(net_female$colors)

# Plotting for females
plotDendroAndColors(net_female$dendrograms[[1]], mergedColors_female[net_female$blockGenes[[1]]],
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)

# Count the number of genes in each female module
moduleSizes_female <- table(net_female$colors)

# Plot the female module sizes
barplot(moduleSizes_female, main = "Number of genes in each female module",
        xlab = "Module colors", ylab = "Number of genes",
        col = names(moduleSizes_female))

```

## Signed

```{r}
# Subset the data for females
female_samples <- samples[samples$sex == "female", ]
female_counts <- counts[, female_samples$sample]

# Convert female counts to DGEList object
dge_female <- DGEList(counts = female_counts)

# Normalization using TMM method
dge_female <- calcNormFactors(dge_female, method = "TMM")

# Compute log2 CPM values, with prior.count=1
log2_cpm_female <- cpm(dge_female, log = TRUE, prior.count = 1)

# Filter out lowly expressed genes
keep_female <- rowSums(cpm(dge_female) > 1) >= 2
log2_cpm_female <- log2_cpm_female[keep_female,]

# Prepare the trait data for females
trait_data_female <- as.data.frame(t(female_samples))

# Run WGCNA analysis for females
enableWGCNAThreads()
```

Select power to achieve scale-free topology...
```{r}
powers <- c(1:10)
pickSoftThreshold(t(log2_cpm_female), powerVector = powers, verbose = 5)
```

## F - Signed Figures

Power of 3 identified by pickSoftThreshold...
```{r}
# Module detection for females
net_female = blockwiseModules(t(log2_cpm_female), power = 3,
                       TOMType = "signed", minModuleSize = 30,
                       reassignThreshold = 0, mergeCutHeight = 0.25,
                       numericLabels = TRUE, pamRespectsDendro = FALSE,
                       saveTOMs = TRUE, saveTOMFileBase = "TOM_female",
                       verbose = 3)

# Convert labels to colors for easier plotting for females
mergedColors_female = labels2colors(net_female$colors)

# Plotting for females
plotDendroAndColors(net_female$dendrograms[[1]], mergedColors_female[net_female$blockGenes[[1]]],
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)

# Count the number of genes in each female module
moduleSizes_female <- table(net_female$colors)

# Plot the female module sizes
barplot(moduleSizes_female, main = "Number of genes in each female module",
        xlab = "Module colors", ylab = "Number of genes",
        col = names(moduleSizes_female))

```

```{r}
# Calculate module eigengenes
MEs_female = moduleEigengenes(t(log2_cpm_female), colors = net_female$colors)$eigengenes

# Organize the module eigengenes into a data frame for easier handling later
MEs_female = orderMEs(MEs_female)

```

```{r}
# Hierarchical clustering of module eigengenes
mediss = 1 - cor(MEs_female)
METree = hclust(as.dist(mediss), method = "average")

```

```{r}
# Plot the dendrogram
plot(METree, main = "Clustering of module eigengenes", xlab = "", sub = "")

```

```{r}
# Calculate the dissimilarity threshold from the correlation threshold
dissimilarityThreshold = 1 - 0.60

# Merge module colors based on the cut height (dissimilarity threshold)
# Note: This step actually performs the merging based on the determined cut height
mergedModules_female = mergeCloseModules(t(log2_cpm_female), colors = net_female$colors, cutHeight = dissimilarityThreshold)

# Extract the 'colors' vector from the list returned by mergeCloseModules
mergedColors_female = mergedModules_female$colors

# New module eigengenes after merging
newMEs_female = moduleEigengenes(t(log2_cpm_female), colors = mergedColors_female)$eigengenes
newMEs_female = orderMEs(newMEs_female)

```

```{r}
# Recalculate Module Eigengenes with the new merged colors
MEs_female_merged = moduleEigengenes(t(log2_cpm_female), colors = mergedColors_female)$eigengenes

# Reorder the module eigengenes
MEs_female_merged = orderMEs(MEs_female_merged)

# Recompute the hierarchical clustering dendrogram using the merged module eigengenes
medissMerged = 1 - cor(MEs_female_merged)
METreeMerged = hclust(as.dist(medissMerged), method = "average")

# Plot the dendrogram and add the module colors underneath
plot(METreeMerged, main = "Clustering of merged module eigengenes", xlab = "", sub = "")
abline(h = dissimilarityThreshold, col = "red")

# Add module colors underneath the dendrogram
mergedModuleColors = mergedColors_female[order(MEs_female_merged)]
colorLabels = as.character(unique(mergedModuleColors))
moduleColors = labels2colors(mergedModuleColors)
colorTable = table(mergedModuleColors)
colorTable = colorTable[colorLabels]
colors = labels2colors(colorLabels)
barplot(-colorTable, col = colors, names.arg = colorLabels, horiz = TRUE, axes = FALSE, border = NA)

```

```{r}
# Extract the unique color labels from the mergedColors_female vector
uniqueMergedColors <- unique(mergedColors_female)

# Determine the new number of modules after merging
numMergedModules <- length(uniqueMergedColors)

# Print the new number of modules
print(paste("Number of modules after merging:", numMergedModules))

```


# Males - Signed

## M - Signed Figures

```{r}
# Module detection for males
net_male = blockwiseModules(t(log2_cpm_male), power = 7,
                       TOMType = "signed", minModuleSize = 30,
                       reassignThreshold = 0, mergeCutHeight = 0.25,
                       numericLabels = TRUE, pamRespectsDendro = FALSE,
                       saveTOMs = TRUE, saveTOMFileBase = "TOM_male",
                       verbose = 3)

# Convert labels to colors for easier plotting for males
mergedColors_male = labels2colors(net_male$colors)

# Plotting for males
plotDendroAndColors(net_male$dendrograms[[1]], mergedColors_male[net_male$blockGenes[[1]]],
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)

# Count the number of genes in each male module
moduleSizes_male <- table(net_male$colors)

# Plot the male module sizes
barplot(moduleSizes_male, main = "Number of genes in each male module",
        xlab = "Module colors", ylab = "Number of genes",
        col = names(moduleSizes_male))

```

```{r}
# Calculate module eigengenes
MEs_male = moduleEigengenes(t(log2_cpm_male), colors = net_male$colors)$eigengenes

# Organize the module eigengenes into a data frame for easier handling later
MEs_male = orderMEs(MEs_male)

```

```{r}
# Hierarchical clustering of module eigengenes
mediss = 1 - cor(MEs_male)
METree = hclust(as.dist(mediss), method = "average")

```

```{r}
# Plot the dendrogram
plot(METree, main = "Clustering of module eigengenes", xlab = "", sub = "")

```

```{r}
# Calculate the dissimilarity threshold from the correlation threshold
dissimilarityThreshold = 1 - 0.85

# Merge module colors based on the cut height (dissimilarity threshold)
# Note: This step actually performs the merging based on the determined cut height
mergedModules_male = mergeCloseModules(t(log2_cpm_male), colors = net_male$colors, cutHeight = dissimilarityThreshold)

# Extract the 'colors' vector from the list returned by mergeCloseModules
mergedColors_male = mergedModules_male$colors

# New module eigengenes after merging
newMEs_male = moduleEigengenes(t(log2_cpm_male), colors = mergedColors_male)$eigengenes
newMEs_male = orderMEs(newMEs_male)

```

```{r}
# Recalculate Module Eigengenes with the new merged colors
MEs_male_merged = moduleEigengenes(t(log2_cpm_male), colors = mergedColors_male)$eigengenes

# Reorder the module eigengenes
MEs_male_merged = orderMEs(MEs_male_merged)

# Recompute the hierarchical clustering dendrogram using the merged module eigengenes
medissMerged = 1 - cor(MEs_male_merged)
METreeMerged = hclust(as.dist(medissMerged), method = "average")

# Plot the dendrogram and add the module colors underneath
plot(METreeMerged, main = "Clustering of merged module eigengenes", xlab = "", sub = "")
abline(h = dissimilarityThreshold, col = "red")

# Add module colors underneath the dendrogram
mergedModuleColors = mergedColors_male[order(MEs_male_merged)]
colorLabels = as.character(unique(mergedModuleColors))
moduleColors = labels2colors(mergedModuleColors)
colorTable = table(mergedModuleColors)
colorTable = colorTable[colorLabels]
colors = labels2colors(colorLabels)
barplot(-colorTable, col = colors, names.arg = colorLabels, horiz = TRUE, axes = FALSE, border = NA)

```

```{r}
# Extract the unique color labels from the mergedColors_female vector
uniqueMergedColors <- unique(mergedColors_male)

# Determine the new number of modules after merging
numMergedModules <- length(uniqueMergedColors)

# Print the new number of modules
print(paste("Number of modules after merging:", numMergedModules))

```















