---
title: "06-WGCNA-TreeCut-distance-matrix"
author: "Zach Bengtsson"
date: "2023-08-22"
output: html_document
---

From Ariana's original code here: <https://github.com/AHuffmyer/EarlyLifeHistory_Energetics/blob/master/Mcap2020/Scripts/TagSeq/Genome_V3/1_WGCNA_Mcap_V3.Rmd>

WGCNA analysis of TagSeq dataset for Montipora capitata developmental time series (2020) using V3 of the M. capitata genome.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
# The following setting is important, do not omit.
options(stringsAsFactors = FALSE) #Set Strings to character
```

Load required libraries.

```{r}
if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse') 
if ("genefilter" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("genefilter") 
if ("DESeq2" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install('DESeq2') 
if ("RColorBrewer" %in% rownames(installed.packages()) == 'FALSE') install.packages('RColorBrewer') 
if ("WGCNA" %in% rownames(installed.packages()) == 'FALSE') install.packages('WGCNA') 
if ("flashClust" %in% rownames(installed.packages()) == 'FALSE') install.packages('flashClust') 
if ("gridExtra" %in% rownames(installed.packages()) == 'FALSE') install.packages('gridExtra') 
if ("ComplexHeatmap" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install('ComplexHeatmap') 
if ("goseq" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install('goseq')
if ("dplyr" %in% rownames(installed.packages()) == 'FALSE') install.packages('dplyr') 
if ("clusterProfiler" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install('clusterProfiler') 
if ("pheatmap" %in% rownames(installed.packages()) == 'FALSE') install.packages('pheatmap') 
if ("magrittr" %in% rownames(installed.packages()) == 'FALSE') install.packages('magrittr') 
if ("vegan" %in% rownames(installed.packages()) == 'FALSE') install.packages('vegan') 
if ("factoextra" %in% rownames(installed.packages()) == 'FALSE') install.packages('factoextra') 

library("tidyverse")
library("genefilter")
library("DESeq2")
library("RColorBrewer")
library("WGCNA")
library("flashClust")
library("gridExtra")
library("ComplexHeatmap")
library("goseq")
library("dplyr")
library("clusterProfiler")
library("pheatmap")
library("magrittr")
library("vegan")
library("factoextra")
library("dplyr")

```

# Data Import

Import all of the data so it can be subset by sex later.

```{r}
#full import before sex filtering
samples <- data.frame(
  sample = c("S12M", "S13M", "S16F", "S19F", "S22F", "S23M", "S29F", "S31M", "S35F", "S36F", "S39F", "S3F", "S41F", "S44F", "S48M", "S50F", "S52F", "S53F", "S54F", "S59M", "S64M", "S6M", "S76F", "S77F", "S7M", "S9M"),
  treatment = c("Exposed", "Control", "Control", "Control", "Exposed", "Exposed", "Exposed", "Exposed", "Exposed", "Exposed", "Control", "Exposed", "Exposed", "Control", "Exposed", "Exposed", "Control", "Control", "Control", "Exposed", "Control", "Control", "Control", "Exposed", "Control", "Exposed"),
  sex = c("male", "male", "female", "female", "female", "male", "female", "male", "female", "female", "female", "female", "female", "female", "male", "female", "female", "female", "female", "male", "male", "male", "female", "female", "male", "male")
)

head(samples)
```

```{r}
# Import count matrix
counts <- read.table("~/github/oyster-lnc/output/02-kallisto-merge/merged_counts.txt", header = TRUE, row.names = 1, check.names = FALSE)

head(counts)
```

```{r}
#round counts to the nearest integer
counts <- round(counts, 0)
str(counts)
```

# Female

## **Data input and filtering**

Subset data by sex, for females in this case.

```{r}
#load metadata but only for female samples
metadataF <- samples[samples$sex == "female", ]
head(metadataF)

#load counts but only for female samples
# Assuming female_counts contains a column named sample_id
#load counts but only for female samples
countsF <- counts[, metadataF$sample]
head(countsF)


# countsF <- subset(counts, sex == "female")
# head(countsF)
```

```{r}
#load gene count matrix for females only
gcountF <- countsF
head(gcountF)

```

Check that there are no genes with 0 counts across all samples.

```{r}
nrow(gcountF)
gcountF<-gcountF %>%
     mutate(Total = rowSums(.[, 1:16]))%>%
    filter(!Total==0)%>%
    dplyr::select(!Total)
nrow(gcountF)
```

We had 4750 lncRNAs, which was filtered down to 4552 by removing genes with row sums of 0 (those not detected in our sequences).

Conduct data filtering, this includes:

*pOverA*: Basically using this to choose lncRNAs that are expressed in 50% of samples (since there are two groups control and treatment 50/50). Setting to 0 seems important since there are 2 treatments, meaning some samples would be at 0 if the lncRNA is only environmentally sensitive.

```{r}
filtF <- filterfun(pOverA(1,5))

#create filter for the counts data
gfiltF <- genefilter(gcountF, filtF)

#identify genes to keep by count filter
gkeepF <- gcountF[gfiltF,]

#identify genes to keep by count filter
gkeepF <- gcountF[gfiltF,]

#identify gene lists
gn.keepF <- rownames(gkeepF)

#gene count data filtered in PoverA, P percent of the samples have counts over A
gcount_filtF <- as.data.frame(gcountF[which(rownames(gcountF) %in% gn.keepF),])

#How many rows do we have before and after filtering?
nrow(gcountF) #Before
nrow(gcount_filtF) #After
```

Filtered from 4552 to 2412 assuming 1 out of all individuals expressing at least 5 counts is an appropriate metric.

```{r}
head(gcount_filtF)
```

Before filtering, we had 4552 genes. After filtering for pOverA, we have approximately 2412 genes.

In order for the DESeq2 algorithms to work, the SampleIDs on the metadta file and count matrices have to match exactly and in the same order. The following R clump will check to make sure that these match. Should return TRUE.

```{r}
#Checking that all row and column names match. Should return "TRUE"
all(rownames(metadataF$sample) %in% colnames(gcount_filtF))
all(rownames(metadataF$sample) == colnames(gcount_filtF)) 
```

Display current order of metadata and gene count matrix.

```{r}
# samples$sample
# colnames(gcount_filtF)
```

Order metadata the same as the column order in the gene matrix.

```{r}
# Filter out male samples using subset
metadata_orderedF <- subset(samples, sex != "male")


head(metadata_orderedF)
```

# **Construct DESeq2 data set**

Create a DESeqDataSet design from gene count matrix and labels. Here we set the design to look at lifestage to test for any differences in gene expression across timepoints.

```{r}
#Set DESeq2 design
gdds <- DESeqDataSetFromMatrix(countData = gcount_filtF,
                              colData = metadata_orderedF,
                              design = ~treatment)
```

First we are going to log-transform the data using a variance stabilizing transforamtion (VST). This is only for visualization purposes. Essentially, this is roughly similar to putting the data on the log2 scale. It will deal with the sampling variability of low counts by calculating within-group variability (if blind=FALSE). Importantly, it does not use the design to remove variation in the data, and so can be used to examine if there may be any variability do to technical factors such as extraction batch effects.

To do this we first need to calculate the size factors of our samples. This is a rough estimate of how many reads each sample contains compared to the others. In order to use VST (the faster log2 transforming process) to log-transform our data, the size factors need to be less than 4.

Chunk should return TRUE if \<4.

```{r}
SF.gdds <- estimateSizeFactors(gdds) #estimate size factors to determine if we can use vst  to transform our data. Size factors should be less than 4 for us to use vst
print(sizeFactors(SF.gdds)) #View size factors

all(sizeFactors(SF.gdds)) < 4
```

All size factors are less than 4, so we can use VST transformation.

```{r}
gvst <- vst(gdds, blind=FALSE) #apply a variance stabilizing transformation to minimize effects of small counts and normalize wrt library size
head(assay(gvst), 3) #view transformed gene count data for the first three genes in the dataset.  
```

# **Examine PCA and sample distances**

Plot a heatmap to sample to sample distances

```{r}
gsampleDists <- dist(t(assay(gvst))) #calculate distance matix
gsampleDistMatrix <- as.matrix(gsampleDists) #distance matrix
rownames(gsampleDistMatrix) <- colnames(gvst) #assign row names
colnames(gsampleDistMatrix) <- NULL #assign col names
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255) #assign colors

save_pheatmap_pdf <- function(x, filename, width=7, height=7) {
   stopifnot(!missing(x))
   stopifnot(!missing(filename))
   pdf(filename, width=width, height=height)
   grid::grid.newpage()
   grid::grid.draw(x$gtable)
   dev.off()
}

pht<-pheatmap(gsampleDistMatrix, #plot matrix
         clustering_distance_rows=gsampleDists, #cluster rows
         clustering_distance_cols=gsampleDists, #cluster columns
         col=colors) #set colors

```

Plot a PCA of samples by lifestages

```{r}
gPCAdata <- plotPCA(gvst, intgroup = c("treatment"), returnData=TRUE, ntop=2412) #use ntop to specify all genes

percentVar <- round(100*attr(gPCAdata, "percentVar")) #plot PCA of samples with all data

allgenesfilt_PCA <- ggplot(gPCAdata, aes(PC1, PC2, shape=treatment)) + 
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) +
  coord_fixed()+
  theme_bw() + #Set background color
  theme(panel.border = element_blank(), # Set border
                     #panel.grid.major = element_blank(), #Set major gridlines 
                     #panel.grid.minor = element_blank(), #Set minor gridlines
                     axis.line = element_line(colour = "black"), #Set axes color
        plot.background=element_blank()) # + #Set the plot background
  #theme(legend.position = ("none")) #set title attributes
allgenesfilt_PCA
#ggsave("Mcap2020/Figures/TagSeq/GenomeV3/allgenesfilt-PCA.pdf", allgenesfilt_PCA, width=11, height=8)
```

# **WGCNA analysis using Dynamic Tree Cut**

Data are analyzed using dynamic tree cut approach, our data set is not large enough to have to use blockwiseModules to break data into "blocks". This code uses step by step network analysis and module detection based on scripts from <https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/index.html>.

Transpose the filtered gene count matrix so that the gene IDs are rows and the sample IDs are columns.

```{r}
datExpr <- as.data.frame(t(assay(gvst))) #transpose to output to a new data frame with the column names as row names. And make all data numeric
```

Check for genes and samples with too many missing values with goodSamplesGenes. There shouldn't be any because we performed pre-filtering

```{r}
gsg = goodSamplesGenes(datExpr, verbose = 3)
gsg$allOK #Should return TRUE if not, the R chunk below will take care of flagged data
```

Remove flagged samples if the allOK is FALSE, not used here.

```{r}
#ncol(datExpr) #number genes before
#if (!gsg$allOK) #If the allOK is FALSE...
#{
# Optionally, print the gene and sample names that are flagged:
#if (sum(!gsg$goodGenes)>0)
#printFlush(paste("Removing genes:", paste(names(datExpr)[!gsg$goodGenes], collapse = ", ")));
#if (sum(!gsg$goodSamples)>0)
#printFlush(paste("Removing samples:", paste(rownames(datExpr)[!gsg$goodSamples], collapse = ", ")));
# Remove the offending genes and samples from the data:
#datExpr = datExpr[gsg$goodSamples, gsg$goodGenes]
#}
#ncol(datExpr) #number genes after
```

Look for outliers by examining tree of samples

```{r}
# sampleTree = hclust(dist(datExpr), method = "average");
# # Plot the sample tree: Open a graphic output window of size 12 by 9 inches
# # The user should change the dimensions if the window is too large or too small.
# plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5, cex.axis = 1.5, cex.main = 2)
# dev.off()
```

There don't look to be any outliers, so we will move on with business as usual.

## Network construction and consensus module detection

### Choosing a soft-thresholding power: Analysis of network topology β

The soft thresholding power (β) is the number to which the co-expression similarity is raised to calculate adjacency. The function pickSoftThreshold performs a network topology analysis. The user chooses a set of candidate powers, however the default parameters are suitable values.

```{r, message=FALSE, warning=FALSE}
allowWGCNAThreads()
# # Choose a set of soft-thresholding powers
powers <- c(seq(from = 1, to=19, by=2), c(21:30)) #Create a string of numbers from 1 through 10, and even numbers from 10 through 20
# 
# # Call the network topology analysis function
sft <-pickSoftThreshold(datExpr, powerVector = powers, verbose = 5)
```

Plot the results.

```{r}
sizeGrWindow(9, 5)
par(mfrow = c(1,2));
cex1 = 0.9;
# # # Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
      xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"));
 text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=cex1,col="red");
# # # this line corresponds to using an R^2 cut-off
 abline(h=0.9,col="red")
# # # Mean connectivity as a function of the soft-thresholding power
 plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity"))
 text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
```

Power of 7

### Co-expression adjacency and topological overlap matrix similarity

```{r}
# Load the data saved
load(file="~/github/oyster-lnc/data/datExpr.RData")
```

```{r}
# Set up workspace and allow multi-threading within WGCNA
options(stringsAsFactors = FALSE)
enableWGCNAThreads() 

# Run analysis
softPower = 7
adjacency = adjacency(datExpr, power = softPower, type = "signed")
TOM = TOMsimilarity(adjacency, TOMType = "signed")
dissTOM = 1 - TOM

# Save the results
save(adjacency, TOM, dissTOM, file = "~/github/oyster-lnc/data/adjTOM.RData")
save(dissTOM, file = "~/github/oyster-lnc/data/dissTOM.RData")
```

### Clustering using TOM

```{r}
# dim(dissTOM)
# str(geneTree)  # Structure of the dendrogram object

```

```{r}
# Load in dissTOM file
tmp <- load(file="~/github/oyster-lnc/data/dissTOM.RData")

# Form distance matrix and plot a dendrogram of genes
geneTree <- flashClust(as.dist(dissTOM), method="average")
plot(geneTree, xlab="", sub="", main= "Gene Clustering on TOM-based dissimilarity",hang=0.04)
```

### Module identification using dynamicTreeCut

```{r}
# Define minimum module size and identify modules
minModuleSize = 5
dynamicMods = cutreeDynamic(dendro = geneTree, distM = dissTOM,
                            deepSplit = 4, pamRespectsDendro = FALSE,
                            minClusterSize = minModuleSize)

# Display and save the results
print(table(dynamicMods))
save(dynamicMods, geneTree, file = "~/github/oyster-lnc/data/dyMod_geneTree.RData")


# Load modules calculated from the adjacency matrix
load(file = "~/github/oyster-lnc/data/dyMod_geneTree.RData")

# Plot the module assignment under the gene dendrogram
dynamicColors = labels2colors(dynamicMods)
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",
                    dendroLabels = FALSE, hang = 0.03, addGuide = TRUE, guideHang = 0.05,
                    main = "Gene dendrogram and module colors")
```

### Merge modules with similar expression profiles

Plot module similarity based on eigengene value

```{r}
#Calculate eigengenes
MEList = moduleEigengenes(datExpr, colors = dynamicColors, softPower = 7)
MEs = MEList$eigengenes

#Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs)

#Cluster again and plot the results
METree = flashClust(as.dist(MEDiss), method = "average")

# pdf(file="Mcap2020/Figures/TagSeq/GenomeV3/eigengeneClustering1.pdf", width = 20)
# plot(METree, main = "Clustering of module eigengenes", xlab = "", sub = "")
```

**Merge modules with \>85% eigengene similarity.** Most studies use somewhere between 80-90% similarity. I will use 85% similarity as my merging threshold.

```{r}
MEDissThres= 0.15 #merge modules that are 85% similar

# plot(METree, main = "Clustering of module eigengenes", xlab = "", sub = "")
# abline(h=MEDissThres, col="red")
# dev.off()

merge= mergeCloseModules(datExpr, dynamicColors, cutHeight= MEDissThres, verbose =3)

mergedColors= merge$colors
mergedMEs= merge$newMEs

plotDendroAndColors(geneTree, cbind(dynamicColors, mergedColors), c("Dynamic Tree Cut", "Merged dynamic"), dendroLabels= FALSE, hang=0.03, addGuide= TRUE, guideHang=0.05)
```

Save new colors

```{r}
moduleLabels=mergedColors
moduleColors = mergedColors # Rename to moduleColors
#colorOrder = c("grey", standardColors(50)); # Construct numerical labels corresponding to the colors
#moduleLabels = match(moduleColors, colorOrder)-1;
MEs = mergedMEs;
ncol(MEs) #How many modules do we have now?
```

Plot new tree with modules.

```{r}
#Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs)
#Cluster again and plot the results
METree = flashClust(as.dist(MEDiss), method = "average")
# MEtreePlot = plot(METree, main = "Clustering of module eigengenes", xlab = "", sub = "")
```

Display table of module gene counts.

```{r}
table(mergedColors)
table<-as.data.frame(table(mergedColors))
write_csv(table, "~/github/oyster-lnc/output/genes_per_module.csv")
```

## Relating modules to developmental stage

### Quantifying module--trait associations

Prepare trait data. Data has to be numeric, so I will substitute time_points and type for numeric values. The "trait" we are considering here is ontogeny.

Make a dataframe that has a column for each lifestage name and a row for samples. Populate a 1 for samples that match each lifestage and a 0 for samples not matching respective lifestages.

This process changes lifestages from a categorical variable into a binary variable. This will allow for correlations between mean eigengenes and lifestage.

```{r}
metadata_orderedF$num <- c("1")
allTraits <- as.data.frame(pivot_wider(metadata_orderedF, names_from = treatment, values_from = num, id_cols = sample))
allTraits[is.na(allTraits)] <- c("0")
rownames(allTraits) <- allTraits$sample
datTraits <- allTraits[,c(-1)]
datTraits
```

Define numbers of genes and samples and print.

```{r}
nGenes = ncol(datExpr)
nSamples = nrow(datExpr)

nGenes
nSamples
```

We have 2412 genes and 16 samples.

Generate labels for module eigengenes as numbers.

```{r}
MEs0 = moduleEigengenes(datExpr, moduleLabels, softPower=7)$eigengenes
MEs = orderMEs(MEs0)
names(MEs)
```

Our module names are:

Like a lot

```{r}
# Check the class of MEs
class(MEs)

# Check the class of datTraits
class(datTraits)

```

```{r}
# Check the first few rows of MEs
head(MEs)

# Check the first few rows of datTraits
head(datTraits)

```

Correlations of traits with eigengenes

```{r}
moduleTraitCor = cor(MEs, datTraits, use = "p");
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples);
Colors=sub("ME","", names(MEs))

# pdf(file="~/github/oyster-lnc/output/hclust.pdf")
moduleTraitTree = hclust(dist(t(moduleTraitCor)), method = "average")
# plot(moduleTraitTree)
```

```{r}
# Check the dimensions of the moduleTraitCor matrix
corDimensions <- dim(moduleTraitCor)

# Check if it's a square matrix
if (corDimensions[1] == corDimensions[2]) {
  cat("moduleTraitCor is a square matrix with dimensions:", corDimensions[1], "x", corDimensions[2], "\n")
} else {
  cat("moduleTraitCor is not a square matrix. Dimensions:", corDimensions[1], "x", corDimensions[2], "\n")
}

```

Correlations of genes with eigengenes. Calculate correlations between ME's and lifestages.

```{r}
moduleGeneCor=cor(MEs,datExpr)
moduleGenePvalue = corPvalueStudent(moduleGeneCor, nSamples);
```

Calculate kME values (module membership).

```{r}
datKME = signedKME(datExpr, MEs, outputColumnName = "kME")
head(datKME)
```

Save module colors and labels for use in subsequent analyses.

```{r}
#save(MEs, moduleLabels, moduleColors, geneTree, file="Mcap2020/Output/TagSeq/NetworkConstruction-stepByStep.RData") 
```

### Plot module-trait associations

Generate a complex heatmap of module-trait relationships.

```{r}
#bold sig p-values
#dendrogram with WGCNA MEtree cut-off
#colored y-axis


#Create list of pvalues for eigengene correlation with specific life stages
heatmappval <- signif(moduleTraitPvalue, 1)

#Make list of heatmap row colors
htmap.colors <- names(MEs)
htmap.colors <- gsub("ME", "", htmap.colors)

# lifestage_order<-c("Egg (1 hpf)", "Embryo (5 hpf)", "Embryo (38 hpf)", "Embryo (65 hpf)", "Larvae (93 hpf)", "Larvae (163 hpf)", "Larvae (231 hpf)", "Metamorphosed Polyp (231 hpf)", "Attached Recruit (183 hpf)", "Attached Recruit (231 hpf)")

library(dendsort)
row_dend = dendsort(hclust(dist(moduleTraitCor)))
col_dend = dendsort(hclust(dist(t(moduleTraitCor))))

pdf(file = "~/github/oyster-lnc/output/heatmap.pdf", height = 25, width = 10)
Heatmap(moduleTraitCor, name = "Eigengene", row_title = "Gene Module", column_title = "Module-Treatment Eigengene Correlation", 
        col = blueWhiteRed(50), 
        row_names_side = "left", 
        #row_dend_side = "left",
        width = unit(5, "in"), 
        height = unit(15, "in"), 
        #column_dend_reorder = TRUE, 
        #cluster_columns = col_dend,
        row_dend_reorder = TRUE,
        #column_split = 6,
        row_split=3,
        #column_dend_height = unit(.5, "in"),
        # column_order = lifestage_order, 
        cluster_rows = row_dend, 
        row_gap = unit(2.5, "mm"), 
        border = TRUE,
        cell_fun = function(j, i, x, y, w, h, col) {
        if(heatmappval[i, j] < 0.05) {
            grid.text(sprintf("%s", heatmappval[i, j]), x, y, gp = gpar(fontsize = 10, fontface = "bold"))
        }
        else {
            grid.text(sprintf("%s", heatmappval[i, j]), x, y, gp = gpar(fontsize = 10, fontface = "plain"))
        }},
        column_names_gp =  gpar(fontsize = 12, border=FALSE),
        column_names_rot = 35,
        row_names_gp = gpar(fontsize = 12, alpha = 0.75, border = FALSE))
#draw(ht)
dev.off()

```

# **Plot mean eigengene over developmental stages**

View module eigengene data and make dataframe for Strader plots.

```{r}
head(MEs)
names(MEs)
Strader_MEs <- MEs
Strader_MEs$sample <- metadata_orderedF$sample
Strader_MEs$sample <- rownames(Strader_MEs)
head(Strader_MEs)

Strader_MEs<-Strader_MEs%>%
  droplevels() #drop unused level

dim(Strader_MEs)
head(Strader_MEs)
```

Plot mean module eigengene for each module.

```{r, echo=true}
#convert wide format to long format for plotting  
plot_MEs<-Strader_MEs%>%
  gather(., key="Module", value="Mean", MEtan:MEyellow3)

write_csv(plot_MEs, "~/github/oyster-lnc/output/plotMEs")

#dev.off() 

#time_point_order = c("Egg (1 hpf)", "Embryo (5 hpf)", "Embryo (38 hpf)", "Embryo (65 hpf)", "Larvae (93 hpf)", "Larvae (163 hpf)", "Larvae (231 hpf)", "Metamorphosed Polyp (231 hpf)", "Attached Recruit (183 hpf)", "Attached Recruit (231 hpf)") #Set time_point order

#add group to dataframe for scale colour manual 
#plot_MEs<-plot_MEs%>%
  # mutate(group=if_else(grepl("Egg", lifestage), "Fertilized Egg", as.character(lifestage)))%>%
  # mutate(group=if_else(grepl("Embryo", lifestage), "Embryo", as.character(group)))%>%
  # mutate(group=if_else(grepl("Larvae", lifestage), "Larvae", as.character(group)))%>%
  # mutate(group=if_else(grepl("Polyp", lifestage), "Metamorphosed Polyp", as.character(group)))%>%
  # mutate(group=if_else(grepl("Recruit", lifestage), "Attached Recruit", as.character(group)))

# list_groups<-c("Fertilized Egg", "Embryo", "Larvae", "Metamorphosed Polyp", "Attached Recruit")
# plot_MEs$group<-factor(as.character(plot_MEs$group), levels=list_groups)

expression_plots<-plot_MEs%>%
  group_by(Module, sample) %>%
  ggplot(aes(x=sample, y=Mean, group=sample, colour=sample)) +
  facet_wrap(~ Module)+
  geom_jitter(alpha = 0.5) +
  geom_boxplot(alpha=0) +
  #scale_x_discrete(name="", limits=time_point_order) +
  #ylim(-0.5,1) +
  ylab("Mean Module Eigenegene") +
  #scale_colour_manual(name="Lifestage", values = c("Fertilized Egg"="#8C510A", "Embryo"="#DFC27D", "Larvae"="#80CDC1", "Metamorphosed Polyp"="#003C30", "Attached Recruit"="#BA55D3")) +
  geom_hline(yintercept = 0, linetype="dashed", color = "grey")+
  theme_bw() + 
  theme(axis.text.x=element_text(angle = 45, hjust=1, size = 12), #set x-axis label size
        axis.title.x=element_text(size = 14), #set x-axis title size
        axis.ticks.x=element_blank(), #No x-label ticks
        #axis.title.y=element_blank(), #No y-axis title
        axis.text.y=element_text(size = 14), #set y-axis label size, 
        panel.border = element_rect(color = "black", fill = NA, size = 1), #set border
        panel.grid.major = element_blank(), #Set major gridlines
        panel.grid.minor = element_blank(), #Set minor gridlines
        axis.line = element_line(colour = "black"), #Set axes color
        plot.background=element_blank(),
        plot.title = element_text(size=22)); expression_plots

ggsave(expression_plots, file="~/github/oyster-lnc/output/expression_eigengene.jpeg", height=8, width=10)

```

```{r, echo = TRUE}
#convert wide format to long format for plotting  
plot_MEs<-Strader_MEs%>%
  gather(., key="Module", value="Mean", MEtan:MEyellow3)

write_csv(plot_MEs, "~/github/oyster-lnc/output/plotMEs-treatment")

#dev.off() 

#time_point_order = c("Egg (1 hpf)", "Embryo (5 hpf)", "Embryo (38 hpf)", "Embryo (65 hpf)", "Larvae (93 hpf)", "Larvae (163 hpf)", "Larvae (231 hpf)", "Metamorphosed Polyp (231 hpf)", "Attached Recruit (183 hpf)", "Attached Recruit (231 hpf)") #Set time_point order

#add group to dataframe for scale colour manual 
#plot_MEs<-plot_MEs%>%
  # mutate(group=if_else(grepl("Egg", lifestage), "Fertilized Egg", as.character(lifestage)))%>%
  # mutate(group=if_else(grepl("Embryo", lifestage), "Embryo", as.character(group)))%>%
  # mutate(group=if_else(grepl("Larvae", lifestage), "Larvae", as.character(group)))%>%
  # mutate(group=if_else(grepl("Polyp", lifestage), "Metamorphosed Polyp", as.character(group)))%>%
  # mutate(group=if_else(grepl("Recruit", lifestage), "Attached Recruit", as.character(group)))

# list_groups<-c("Fertilized Egg", "Embryo", "Larvae", "Metamorphosed Polyp", "Attached Recruit")
# plot_MEs$group<-factor(as.character(plot_MEs$group), levels=list_groups)

expression_plots<-plot_MEs%>%
  group_by(Module, treatment) %>%
  ggplot(aes(x=sample, y=Mean, group=treatment, colour=treatment)) +
  facet_wrap(~ Module)+
  geom_jitter(alpha = 0.5) +
  geom_boxplot(alpha=0) +
  #scale_x_discrete(name="", limits=time_point_order) +
  #ylim(-0.5,1) +
  ylab("Mean Module Eigenegene") +
  #scale_colour_manual(name="Lifestage", values = c("Fertilized Egg"="#8C510A", "Embryo"="#DFC27D", "Larvae"="#80CDC1", "Metamorphosed Polyp"="#003C30", "Attached Recruit"="#BA55D3")) +
  geom_hline(yintercept = 0, linetype="dashed", color = "grey")+
  theme_bw() + 
  theme(axis.text.x=element_text(angle = 45, hjust=1, size = 12), #set x-axis label size
        axis.title.x=element_text(size = 14), #set x-axis title size
        axis.ticks.x=element_blank(), #No x-label ticks
        #axis.title.y=element_blank(), #No y-axis title
        axis.text.y=element_text(size = 14), #set y-axis label size, 
        panel.border = element_rect(color = "black", fill = NA, size = 1), #set border
        panel.grid.major = element_blank(), #Set major gridlines
        panel.grid.minor = element_blank(), #Set minor gridlines
        axis.line = element_line(colour = "black"), #Set axes color
        plot.background=element_blank(),
        plot.title = element_text(size=22)); expression_plots

ggsave(expression_plots, file="~/github/oyster-lnc/output/treatment_expression_eigengene.jpeg", height=8, width=10)
```

# **Calculating gene significance and module membership**

We quantify associations of individual genes with life stage by defining Gene Significance GS as the absolute value of the correlation between the gene and the lifestage timepoint. For each module, we also define a quantitative measure of module membership (MM) as the correlation of the module eigengene and the gene expression profile.

Define lifestage as a numeric value (ranging from 1 (egg, first timepoint)-10 (recruit, last time point)).

```{r}
expressionProfile_data <- Strader_MEs

expressionProfile_data<-expressionProfile_data%>%
  droplevels() #drop unused level

cols.num <- c(1:9)

expressionProfile_data[cols.num] <- sapply(expressionProfile_data[cols.num],as.numeric)
sapply(expressionProfile_data, class)

dim(expressionProfile_data)
head(expressionProfile_data)

#need to get number of order of lifestage for this part 

expressionProfile_data$time_stage<-metadata$time_stage[match(expressionProfile_data$lifestage, metadata$lifestage)]
head(expressionProfile_data)

time_stage <- as.data.frame(expressionProfile_data$time_stage)
names(time_stage) = "lifestage_num"

time_stage$lifestage_num <- sub("_[^_]+$", "", time_stage$lifestage_num) #only extract the number to turn lifestage into numeric values 

time_stage$lifestage_num<-as.numeric(time_stage$lifestage_num)
dim(time_stage)
```

Generate data frame for module membership for each gene (absolute value of correlation between module eigengene and gene).\
- geneModuleMembership = correlation of the module eigengene and the gene - MMPvalue = p-value of correlation between module eigengene and the gene

```{r}
#names(datExpr) = gsub(pattern = "*.t1", replacement = "", x = names(datExpr))
#remove ".ti" from names in datExpr file and assign gene id as probe names

modNames = substring(names(MEs), 3) #pull out module names

geneModuleMembership = as.data.frame(cor(datExpr, MEs, use = "p"));
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples));

names(geneModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");
```

Generate dataframe with Gene Significance GS as the absolute value of the correlation between the gene and the lifestage timepoint - geneTraitSignificance = correlation of the gene and lifestage time point - GSPvalue = p-value of correlation between gene and lifestage time point

```{r}

geneTraitSignificance = as.data.frame(cor(datExpr, time_stage, use = "p"));
GSPvalue = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificance), nSamples));

names(geneTraitSignificance) = paste("GS.", names(time_stage), sep="");
names(GSPvalue) = paste("p.GS.", names(time_stage), sep="");
```

Generate a list of genes associated with each module. Change the module number in this code to look at respective module.

```{r}
names(datExpr)[moduleColors=="1"]
```

# **Generate files for Functional Enrichment**

Functional annotation file obtained from: <http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.EggNog_results.txt.gz> on 18 December 2022.

```{r}
Mcap.annot <- read.table("Mcap2020/Data/TagSeq/Montipora_capitata_HIv3.genes.EggNog_results.txt",  quote="", sep="\t", header=TRUE)
dim(Mcap.annot)

# column names
head(Mcap.annot)
tail(Mcap.annot)
```

This functional annotation as 24,072 genes. This is 44% of total protein predicted genes described in the publication associated with this annotation. <https://academic.oup.com/gigascience/article/doi/10.1093/gigascience/giac098/6815755>

## Make a data frame that connects genes, traits, and modules

Match up genes in datExpr file to annotation file

```{r}
names(Mcap.annot)

probes = names(datExpr)

probes2annot = match(probes, Mcap.annot$query) #match genes in datExpr to genes in annotation file, note I removed the # before query in this file before loading

# The following is the number of probes without annotation... 
sum(is.na(probes2annot))

row_nas<-which(is.na(probes2annot))

#view the genes that do not have a match in the annotation file
missing<-as.data.frame(probes[row_nas])
print(missing)
```

3394 genes without annotation information. Spot check confirms the annotation file does not contain these genes.

Create the starting data frame

```{r}
names(Mcap.annot)

geneInfo0 = data.frame(gene_id = probes, #add gene id
Accession = Mcap.annot$seed_ortholog[probes2annot], #add accession number
Bitscore = Mcap.annot$score[probes2annot], #add bitscore
eValue = Mcap.annot$evalue[probes2annot], #add e value
Description = Mcap.annot$Description[probes2annot], #add description of gene
Annotation.GO.ID = Mcap.annot$GOs[probes2annot], #add GO ID's
moduleColor = moduleColors, #add module number
geneTraitSignificance, #add gene significance 
GSPvalue)

head(geneInfo0)
```

Add KEGG annotation information. Downloaded from: <http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.KEGG_results.txt.gz> on 20 December 2022.

```{r}
kegg<-read.table("Mcap2020/Data/TagSeq/Montipora_capitata_HIv3.genes.KEGG_results.txt", sep="", quote="", na.strings=c("","NA"), blank.lines.skip = FALSE, header=FALSE)

head(kegg)
```

Add KEGG annotations to each gene.

```{r}
geneInfo0$KEGG<-kegg$V2[match(geneInfo0$gene_id, kegg$V1)]
```

Order modules by their significance for time_point

```{r}
modOrder = order(-abs(cor(MEs, time_stage, use = "p")))
```

Add module membership information in the chosen order

```{r}
for (mod in 1:ncol(geneModuleMembership))
{
oldNames = names(geneInfo0)
geneInfo0 = data.frame(geneInfo0, geneModuleMembership[, modOrder[mod]],
MMPvalue[, modOrder[mod]]);
names(geneInfo0) = c(oldNames, paste("MM.", modNames[modOrder[mod]], sep=""),
paste("p.MM.", modNames[modOrder[mod]], sep=""))
}
```

Order the genes in the geneInfo variable first by module color, then by geneTraitSignificance

```{r}
geneOrder = order(geneInfo0$moduleColor, -abs(geneInfo0$GS.lifestage_num));
geneInfo = geneInfo0[geneOrder, ]
head(geneInfo)
```

Save geneInfo as a CSV

```{r}
head(geneInfo)
dim(geneInfo)

geneInfo$Annotation.GO.ID <- gsub(";NA", "", geneInfo$Annotation.GO.ID) #Remove NAs
geneInfo$Annotation.GO.ID <- gsub("NA", "", geneInfo$Annotation.GO.ID) #Remove NAs
dim(geneInfo)

write.csv(geneInfo, file = "Mcap2020/Output/TagSeq/GenomeV3/geneInfo.csv") #gene info for reference/supplement

```

Format our gene vst data frame for plotting.

```{r}
gene_expr<-as.matrix(datExpr)
gene_expr<-t(gene_expr)
gene_expr<-as.data.frame(gene_expr, row.names = rownames(gene_expr), make.names=TRUE)
```

Output for other analysis

```{r}
write.csv(gene_expr, "Mcap2020/Output/TagSeq/GenomeV3/transformed_filtered_gene_counts.csv")
```
