---
title: "11-mRNA-counts-DESeq2"
author: "Zach Bengtsson"
date: "2024-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
DESeq2 run for coding sequences for comparison to lncRNA.

Generate kallisto index from CDS fasta obtained from Steven's previous workflow:
https://github.com/sr320/ceabigr/blob/main/code/14-RNA-kallisto.Rmd
## kallisto index
```{bash}
/home/shared/kallisto/kallisto \
index -i /home/shared/8TB_HDD_02/zbengt/github/oyster-lnc/output/11-mRNA-counts-DESeq2/kallisto-index/GCF_002022765.2_C_virginica-3.0_cds_from_genomic.index \
/home/shared/8TB_HDD_01/sr320/github/ceabigr/data/GCF_002022765.2_C_virginica-3.0_cds_from_genomic.fna
```

## kallisto abundance
Get abundance for fq files using the CDS index...
```{bash}
find /home/shared/8TB_HDD_02/zbengt/data/oyster-lnc/01-lncRNA-kallisto/fq/*_R1.fastp-trim.20bp-5prime.20220224.fq.gz \
| xargs -n 1 basename -s _R1.fastp-trim.20bp-5prime.20220224.fq.gz \
| xargs -I{} /home/shared/kallisto/kallisto quant \
  -i ~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/kallisto-index/GCF_002022765.2_C_virginica-3.0_cds_from_genomic.index \
  -o ~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/kallisto-output/{} \
  -t 20 \
  /home/shared/8TB_HDD_02/zbengt/data/oyster-lnc/01-lncRNA-kallisto/fq/{}_R1.fastp-trim.20bp-5prime.20220224.fq.gz \
  /home/shared/8TB_HDD_02/zbengt/data/oyster-lnc/01-lncRNA-kallisto/fq/{}_R2.fastp-trim.20bp-5prime.20220224.fq.gz \
2>&1 | tee ~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/{}.out
```

## trinity count matrix

Generate a count matrix from Kallisto abundance estimates for all transcripts in RNA-seq data…
```{bash}
perl /home/shared/trinityrnaseq-v2.12.0/util/abundance_estimates_to_matrix.pl \
--est_method kallisto \
--gene_trans_map none \
--out_prefix  ~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/kallisto-count-matrix/Cvirg_mRNA_count_matrix \
--name_sample_by_basedir \
 ~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/kallisto-output/*/abundance.tsv
```

## Sex Specific Matrices
From the combined count matrix, create female and male count matrices...

```{r}
# Load the data
combined_matrix_mRNA <- read.table("~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/kallisto-count-matrix/Cvirg_mRNA_count_matrix.isoform.counts.matrix", row.names = 1)

# Filter for "M" and "F"
matrix_M_mRNA <- combined_matrix_mRNA[, grepl("M", colnames(combined_matrix_mRNA))]
matrix_F_mRNA <- combined_matrix_mRNA[, grepl("F", colnames(combined_matrix_mRNA))]

# Write to files
write.csv(matrix_M_mRNA, "~/github/oyster-lnc/output/10-count-matrices-DESeq2-final/kallisto-count-matrix/matrix_M_mRNA.csv", row.names = TRUE)
write.csv(matrix_F_mRNA, "~/github/oyster-lnc/output/10-count-matrices-DESeq2-final/kallisto-count-matrix/matrix_F_mRNA.csv", row.names = TRUE)

```
Objects combined_matrix_mRNA, matrix_M_mRNA, and matrix_F_mRNA can be used in the subsequent DESeq2 runs.

## Combined male and female

Load all of the packages you'll need...
```{r}
library(DESeq2)
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(data.table)
```

Create data frame for DESeq to reference. This specifies which sample IDs are control and which are treatment (OA)…
```{r}
### data frame of conditions/treatments
metaCombinedData_mRNA <- data.frame(
  sample = c("S12M", "S13M", "S16F", "S19F", "S22F", "S23M", "S29F", "S31M", "S35F", "S36F", "S39F", "S3F", "S41F", "S44F", "S48M", "S50F", "S52F", "S53F", "S54F", "S59M", "S64M", "S6M", "S76F", "S77F", "S7M", "S9M"),
  treatment = c("Exposed", "Control", "Control", "Control", "Exposed", "Exposed", "Exposed", "Exposed", "Exposed", "Exposed", "Control", "Exposed", "Exposed", "Control", "Exposed", "Exposed", "Control", "Control", "Control", "Exposed", "Control", "Control", "Control", "Exposed", "Control", "Exposed"),
  sex = c("M", "M", "F", "F", "F", "M", "F", "M", "F", "F", "F", "F", "F", "F", "M", "F", "F", "F", "F", "M", "M", "M", "F", "F", "M", "M")
)
print(metaCombinedData_mRNA)
```

Count matrix alredy loaded as object "combined_matrix_mRNA"
```{r}
head(combined_matrix_mRNA)
```

Round counts to integers so they can be run through DESeq2 and summarize countmatrix structure to check and see that it worked…
```{r}
combined_matrix_mRNA <- round(combined_matrix_mRNA, 0)
str(combined_matrix_mRNA)
```

Make sure number of count matrix columns matches number of meta dataframe rows...
```{r}
# Check the dimensions of countData matrix
ncol(combined_matrix_mRNA)

# Check the dimensions of colData matrix
nrow(metaCombinedData_mRNA)

```

Use the DESeq2 package to create a DESeqDataSet object from a count matrix and metadata…
```{r}
combined_mRNA.deseq2.dds <- DESeqDataSetFromMatrix(countData = combined_matrix_mRNA,
                                     colData = metaCombinedData_mRNA, 
                                     design = ~ treatment + sex)
```

Performs differential expression analysis using DESeq2 on the combined_mRNA.deseq2.dds object, extract the results, converts them into a data frame, and then orders the rows of the data frame alphabetically based on the row names…
```{r}
combined_mRNA.deseq2.dds <- DESeq(combined_mRNA.deseq2.dds)
combined_mRNA.deseq2.res <- results(combined_mRNA.deseq2.dds)
combined_mRNA.deseq2.res <- as.data.frame(combined_mRNA.deseq2.res)  # Convert to data frame
combined_mRNA.deseq2.res <- combined_mRNA.deseq2.res[order(rownames(combined_mRNA.deseq2.res)), ]
```

Check...
```{r}
str(combined_mRNA.deseq2.res)
```

Count the number of hits with adjusted p-value less then 0.05…
```{r}
dim(combined_mRNA.deseq2.res[!is.na(combined_mRNA.deseq2.res$padj) & combined_mRNA.deseq2.res$padj <= 0.05, ])
```

36,242 DEGs without NA and p-value less that 0.05

```{r}
# Subset the hits with adjusted p-value less than 0.05
combined_mRNA.hits <- combined_mRNA.deseq2.res[!is.na(combined_mRNA.deseq2.res$padj) & combined_mRNA.deseq2.res$padj <= 0.05, ]

# View the transcript IDs of the hits
combined_mRNA_transcript_ids <- rownames(combined_mRNA.hits)
print(combined_mRNA_transcript_ids)

```

```{r}
# Save transcript IDs to a text file
write.table(combined_mRNA_transcript_ids, "~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/combined_mRNA_transcript_ids.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)
```


Create a plot of differentially expressed lncRNAs with those that have a p-value less than 0.05 highlighted in red…
```{r}
tmp <- combined_mRNA.deseq2.res
# The main plot
plot(tmp$baseMean, tmp$log2FoldChange, pch=20, cex=0.45, ylim=c(-3, 3), log="x", col="darkgray",
     main="Combined mRNA - Treatment + Sex  (pval <= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")
# Getting the significant points and plotting them again so they're a different color
tmp.sig <- combined_mRNA.deseq2.res[!is.na(combined_mRNA.deseq2.res$padj) & combined_mRNA.deseq2.res$padj <= 0.05, ]
points(tmp.sig$baseMean, tmp.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
```

Generate a PCA plot to see how all of the differential expression results cluster based on sex and treatment…
```{r}
combined_mRNA.vsd <- vst(combined_mRNA.deseq2.dds, blind = FALSE)
plotPCA(combined_mRNA.vsd, intgroup = "sex")
```

```{r}
combined_mRNA.vsd <- vst(combined_mRNA.deseq2.dds, blind = FALSE)
plotPCA(combined_mRNA.vsd, intgroup = "treatment")
```

```{r}
print(combined_mRNA.deseq2.dds)
```


Create heatmaps to view the top 50 differentially expressed lncRNAs…
```{r}
# Select top 50 differentially expressed genes
res_mRNA <- results(combined_mRNA.deseq2.dds)
res_ordered_mRNA <- res_mRNA[order(res_mRNA$padj), ]
top_genes_mRNA <- row.names(res_ordered_mRNA)[1:50]

# Extract counts and normalize
counts_mRNA <- counts(combined_mRNA.deseq2.dds, normalized = TRUE)
counts_top_mRNA <- counts_mRNA[top_genes_mRNA, ]

# Log-transform counts
log_counts_top_mRNA <- log2(counts_top_mRNA + 1)

# Open a PDF device
pdf("~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/figure-pdfs/combined-mRNA-50-heatmap.pdf")

# Generate heatmap
pheatmap(log_counts_top_mRNA, scale = "row")

# Close the PDF device
dev.off()

```

Label PCA plot for treatment...
```{r}
install.packages("ggrepel")
library("ggrepel")

combined_mRNA.vsd <- vst(combined_mRNA.deseq2.dds, blind = FALSE)

# get PCA data
pcaData_mRNA <- plotPCA(combined_mRNA.vsd, intgroup = "treatment", returnData = TRUE)

# get percent variation
percentVar_mRNA <- round(100 * attr(pcaData_mRNA, "percentVar_mRNA"))

# create a ggplot object
p <- ggplot(pcaData_mRNA, aes(x = PC1, y = PC2, color = treatment)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ",percentVar_mRNA[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar_mRNA[2],"% variance")) +
  geom_text_repel(aes(label = rownames(pcaData_mRNA)), max.overlaps = Inf)

print(p)

```

Label PCA plot for sex...
```{r}
combined_mRNA.vsd <- vst(combined_mRNA.deseq2.dds, blind = FALSE)

# get PCA data
pcaData_mRNA <- plotPCA(combined_mRNA.vsd, intgroup = "sex", returnData = TRUE)

# get percent variation
percentVar_mRNA <- round(100 * attr(pcaData_mRNA, "percentVar_mRNA"))

# create a ggplot object
p2 <- ggplot(pcaData_mRNA, aes(x = PC1, y = PC2, color = sex)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ",percentVar_mRNA[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar_mRNA[2],"% variance")) +
  geom_text_repel(aes(label = rownames(pcaData_mRNA)), max.overlaps = Inf)

print(p2)

```

## Female oyster DESeq2 run

Create data frame for DESeq to reference. This specifies which sample IDs are control and which are treatment (OA)…
```{r}
metaFemaleData_mRNA <- data.frame(
  sample = c("S16F", "S19F", "S22F", "S29F", "S35F", "S36F", "S39F", "S3F", "S41F", "S44F", "S50F", "S52F", "S53F", "S54F", "S76F", "S77F"),
  treatment = c("Control", "Control", "Exposed", "Exposed", "Exposed", "Exposed", "Control", "Exposed", "Exposed", "Control", "Exposed", "Control", "Control", "Control", "Control", "Exposed")
  )
```

Load female mRNA count matrix already loaded as object "matrix_F_mRNA"
```{r}
head(matrix_F_mRNA)
```

Round counts to integers so they can be run through DESeq2 and summarize countmatrix structure to check and see that it worked…
```{r}
matrix_F_mRNA <- round(matrix_F_mRNA, 0)
str(matrix_F_mRNA)
```

Make sure number of count matrix columns matches number of meta dataframe rows...
```{r}
# Check the dimensions of countData matrix
ncol(matrix_F_mRNA)

# Check the dimensions of colData matrix
nrow(metaFemaleData_mRNA)

```


Use the DESeq2 package to create a DESeqDataSet object from a count matrix and metadata…
```{r}
female_mRNA.deseq2.dds <- DESeqDataSetFromMatrix(countData = matrix_F_mRNA,
                                     colData = metaFemaleData_mRNA, 
                                     design = ~ treatment)
```

Performs differential expression analysis using DESeq2 on the f.deseq2.dds object, extract the results, converts them into a data frame, and then orders the rows of the data frame alphabetically based on the row names…
```{r}
female_mRNA.deseq2.dds <- DESeq(female_mRNA.deseq2.dds)
female_mRNA.deseq2.res <- results(female_mRNA.deseq2.dds)
female_mRNA.deseq2.res <- as.data.frame(female_mRNA.deseq2.res)  # Convert to data frame
female_mRNA.deseq2.res <- female_mRNA.deseq2.res[order(rownames(female_mRNA.deseq2.res)), ]
```

Check...
```{r}
str(female_mRNA.deseq2.res)
```

Count the number of hits with adjusted p-value less then 0.05…
```{r}
dim(female_mRNA.deseq2.res[!is.na(female_mRNA.deseq2.res$padj) & female_mRNA.deseq2.res$padj <= 0.05, ])
```

25 significantly differentially expressed DEG

```{r}
# Subset the hits with adjusted p-value less than 0.05
female_mRNA_hits <- female_mRNA.deseq2.res[!is.na(female_mRNA.deseq2.res$padj) & female_mRNA.deseq2.res$padj <= 0.05, ]

# View the transcript IDs of the hits
female_mRNA_transcript_ids <- rownames(female_mRNA_hits)
print(female_mRNA_transcript_ids)
```

```{r}
# Save transcript IDs to a text file
write.table(female_mRNA_transcript_ids, "~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/female_mRNA_transcript_ids.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)
```

Create a plot of differentially expressed lncRNAs with those that have a p-value less than 0.05 highlighted in red…
```{r}
tmp <- female_mRNA.deseq2.res
# The main plot
plot(tmp$baseMean, tmp$log2FoldChange, pch=20, cex=0.45, ylim=c(-3, 3), log="x", col="darkgray",
     main="Female - Treatment  (pval <= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")
# Getting the significant points and plotting them again so they're a different color
tmp.sig <- female_mRNA.deseq2.res[!is.na(female_mRNA.deseq2.res$padj) & female_mRNA.deseq2.res$padj <= 0.05, ]
points(tmp.sig$baseMean, tmp.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
```

Generate a PCA plot to see how all of the differential expression results cluster based on treatment…
```{r}
female_mRNA.vsd <- vst(female_mRNA.deseq2.dds, blind = FALSE)
plotPCA(female_mRNA.vsd, intgroup = "treatment")
```

```{r}
female_mRNA.vsd <- vst(female_mRNA.deseq2.dds, blind = FALSE)

# get PCA data
pcaData_mRNA <- plotPCA(female_mRNA.vsd, intgroup = "treatment", returnData = TRUE)

# get percent variation
percentVar_mRNA <- round(100 * attr(pcaData_mRNA, "percentVar_mRNA"))

# create a ggplot object
p <- ggplot(pcaData_mRNA, aes(x = PC1, y = PC2, color = treatment)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ",percentVar_mRNA[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar_mRNA[2],"% variance")) +
  geom_text_repel(aes(label = rownames(pcaData_mRNA)))

print(p)

```


```{r}
print(female_mRNA.deseq2.dds)
```

Create heatmaps to view the top 50 differentially expressed lncRNAs…
```{r}
library(pheatmap)

# Select top 50 differentially expressed genes
res_mRNA <- results(female_mRNA.deseq2.dds)
res_ordered_mRNA <- res_mRNA[order(res_mRNA$padj), ]
top_genes_mRNA <- row.names(res_ordered_mRNA)[1:50]

# Extract counts and normalize
counts_mRNA <- counts(female_mRNA.deseq2.dds, normalized = TRUE)
counts_top_mRNA <- counts_mRNA[top_genes_mRNA, ]

# Log-transform counts
log_counts_top_mRNA <- log2(counts_top_mRNA + 1)

# Open a PDF device
pdf("~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/figure-pdfs/female-mRNA-50-heatmap.pdf")

# Generate heatmap
pheatmap(log_counts_top_mRNA, scale = "row")

# Close the PDF device
dev.off()

```

Remake heatmap to see whether or not control and exposed cluster...
```{r}
library(pheatmap)

# Assuming you have a column 'treatment' in your metadata
# Extract treatment information
sample_treatment_mRNA <- as.data.frame(colData(female_mRNA.deseq2.dds)$treatment)
rownames(sample_treatment_mRNA) <- colnames(counts_mRNA)

# Select top 50 differentially expressed genes
res_mRNA <- results(female_mRNA.deseq2.dds)
res_ordered_mRNA <- res_mRNA[order(res_mRNA$padj), ]
top_genes_mRNA <- row.names(res_ordered_mRNA)[1:50]

# Extract counts and normalize
counts_mRNA <- counts(female_mRNA.deseq2.dds, normalized = TRUE)
counts_top_mRNA <- counts_mRNA[top_genes_mRNA, ]

# Log-transform counts
log_counts_top_mRNA <- log2(counts_top_mRNA + 1)

# Open a PDF device with specified dimensions
pdf("~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/figure-pdfs/treatment-group-female-mRNA-50-heatmap.pdf", width = 14, height = 10)

# Generate heatmap with adjusted settings
pheatmap(log_counts_top_mRNA, 
         scale = "row", 
         annotation_col = sample_treatment_mRNA,
         fontsize_row = 8, 
         fontsize_col = 8, 
         cellwidth = 20,
         cellheight = 10,
         border_color = NA) # Adjust fontsize, cell size, and border as needed

# Close the PDF device
dev.off()

```


## Male oyster DESeq2 run

Create data frame for DESeq to reference. This specifies which sample IDs are control and which are treatment (OA)…
```{r}
metaMaleData_mRNA <- data.frame(
  sample = c("S12M", "S13M", "S23M", "S31M", "S48M", "S59M", "S64M", "S6M", "S7M", "S9M"),
  treatment = c("Exposed", "Control", "Exposed", "Exposed", "Exposed", "Exposed", "Control", "Control", "Control", "Exposed")
  )
```

Load male lncRNA count matrix already loaded as object "matrix_M_mRNA"
```{r}
head(matrix_M_mRNA)
```

Round counts to integers so they can be run through DESeq2 and summarize countmatrix structure to check and see that it worked…
```{r}
matrix_M_mRNA <- round(matrix_M_mRNA, 0)
str(matrix_M_mRNA)
```

Make sure number of count matrix columns matches number of meta dataframe rows...
```{r}
# Check the dimensions of countData matrix
ncol(matrix_M_mRNA)

# Check the dimensions of colData matrix
nrow(metaMaleData_mRNA)

```


Use the DESeq2 package to create a DESeqDataSet object from a count matrix and metadata…
```{r}
library(DESeq2)

male_mRNA.deseq2.dds <- DESeqDataSetFromMatrix(countData = matrix_M_mRNA,
                                     colData = metaMaleData_mRNA, 
                                     design = ~ treatment)
```

Performs differential expression analysis using DESeq2 on the f.deseq2.dds object, extract the results, converts them into a data frame, and then orders the rows of the data frame alphabetically based on the row names…
```{r}
male_mRNA.deseq2.dds <- DESeq(male_mRNA.deseq2.dds)
male_mRNA.deseq2.res <- results(male_mRNA.deseq2.dds)
male_mRNA.deseq2.res <- as.data.frame(male_mRNA.deseq2.res)  # Convert to data frame
male_mRNA.deseq2.res <- male_mRNA.deseq2.res[order(rownames(male_mRNA.deseq2.res)), ]
```

Check...
```{r}
str(male_mRNA.deseq2.res)
```

Count the number of hits with adjusted p-value less then 0.05…
```{r}
dim(male_mRNA.deseq2.res[!is.na(male_mRNA.deseq2.res$padj) & male_mRNA.deseq2.res$padj <= 0.05, ])
```

107 significantly differentially expressed DEGs

```{r}
# Subset the hits with adjusted p-value less than 0.05
male_mRNA_hits <- male_mRNA.deseq2.res[!is.na(male_mRNA.deseq2.res$padj) & male_mRNA.deseq2.res$padj <= 0.05, ]

# View the transcript IDs of the hits
male_mRNA_transcript_ids <- rownames(male_mRNA_hits)
print(male_mRNA_transcript_ids)
```

```{r}
# Save transcript IDs to a text file
write.table(male_mRNA_transcript_ids, "~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/male_mRNA_transcript_ids.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)
```



Create a plot of differentially expressed lncRNAs with those that have a p-value less than 0.05 highlighted in red…
```{r}
tmp <- male_mRNA.deseq2.res
# The main plot
plot(tmp$baseMean, tmp$log2FoldChange, pch=20, cex=0.45, ylim=c(-3, 3), log="x", col="darkgray",
     main="Male - Treatment  (pval <= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")
# Getting the significant points and plotting them again so they're a different color
tmp.sig <- male_mRNA.deseq2.res[!is.na(male_mRNA.deseq2.res$padj) & male_mRNA.deseq2.res$padj <= 0.05, ]
points(tmp.sig$baseMean, tmp.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
```

Generate a PCA plot to see how all of the differential expression results cluster based on treatment…
```{r}
male_mRNA.vsd <- vst(male_mRNA.deseq2.dds, blind = FALSE)
plotPCA(male_mRNA.vsd, intgroup = "treatment")
```

```{r}
library(ggplot2)
library(ggrepel)

male_mRNA.vsd <- vst(male_mRNA.deseq2.dds, blind = FALSE)

# get PCA data
pcaData_mRNA <- plotPCA(male_mRNA.vsd, intgroup = "treatment", returnData = TRUE)

# get percent variation
percentVar_mRNA <- round(100 * attr(pcaData_mRNA, "percentVar_mRNA"))

# create a ggplot object
p <- ggplot(pcaData_mRNA, aes(x = PC1, y = PC2, color = treatment)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ",percentVar_mRNA[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar_mRNA[2],"% variance")) +
  geom_text_repel(aes(label = rownames(pcaData_mRNA)))

print(p)

```


```{r}
print(male_mRNA.deseq2.dds)
```

Create heatmaps to view the top 50 differentially expressed lncRNAs…
```{r}
library(pheatmap)

# Select top 50 differentially expressed genes
res_mRNA <- results(male_mRNA.deseq2.dds)
res_ordered_mRNA <- res_mRNA[order(res_mRNA$padj), ]
top_genes_mRNA <- row.names(res_ordered_mRNA)[1:50]

# Extract counts and normalize
counts_mRNA <- counts(male_mRNA.deseq2.dds, normalized = TRUE)
counts_top_mRNA <- counts_mRNA[top_genes_mRNA, ]

# Log-transform counts
log_counts_top_mRNA <- log2(counts_top_mRNA + 1)

# Open a PDF device
pdf("~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/figure-pdfs/male-mRNA-50-heatmap.pdf")

# Generate heatmap
pheatmap(log_counts_top_mRNA, scale = "row")

# Close the PDF device
dev.off()

```

Remake heatmap to see whether or not control and exposed cluster...
```{r}
library(pheatmap)

# Assuming you have a column 'treatment' in your metadata
# Extract treatment information
sample_treatment_mRNA <- as.data.frame(colData(male_mRNA.deseq2.dds)$treatment)
rownames(sample_treatment_mRNA) <- colnames(counts_mRNA)

# Select top 50 differentially expressed genes
res_mRNA <- results(male_mRNA.deseq2.dds)
res_ordered_mRNA <- res_mRNA[order(res_mRNA$padj), ]
top_genes_mRNA <- row.names(res_ordered_mRNA)[1:50]

# Extract counts and normalize
counts_mRNA <- counts(male_mRNA.deseq2.dds, normalized = TRUE)
counts_top_mRNA <- counts_mRNA[top_genes_mRNA, ]

# Log-transform counts
log_counts_top_mRNA <- log2(counts_top_mRNA + 1)

# Open a PDF device with specified dimensions
pdf("~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/figure-pdfs/treatment-group-male-mRNA-50-heatmap.pdf", width = 14, height = 10)

# Generate heatmap with adjusted settings
pheatmap(log_counts_top_mRNA, 
         scale = "row", 
         annotation_col = sample_treatment_mRNA,
         fontsize_row = 8, 
         fontsize_col = 8, 
         cellwidth = 20,
         cellheight = 10,
         border_color = NA) # Adjust fontsize, cell size, and border as needed

# Close the PDF device
dev.off()
```