---
title: "15-pearsons"
author: "Zach Bengtsson"
date: "2025-11-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(tibble)

data_dir <- "~/github/oyster-lnc/data/15-pearsons"
lnc_file  <- file.path(data_dir, "lnc_counts.tsv")
mrna_file <- "~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/kallisto_gene_counts.tsv"

## load matrices
lnc_raw <- readr::read_tsv(lnc_file, show_col_types = FALSE)
colnames(lnc_raw)[1] <- "gene_id"

lnc_mat <- lnc_raw %>%
  tibble::column_to_rownames("gene_id") %>%
  as.matrix()

mrna_raw <- readr::read_tsv(mrna_file, show_col_types = FALSE)

mrna_mat <- mrna_raw %>%
  tibble::column_to_rownames("gene_id") %>%
  as.matrix()

## coerce to numeric
lnc_mat <- apply(lnc_mat, 2, as.numeric)
rownames(lnc_mat) <- lnc_raw$gene_id

mrna_mat <- apply(mrna_mat, 2, as.numeric)
rownames(mrna_mat) <- mrna_raw$gene_id

## align samples
common_samples <- intersect(colnames(lnc_mat), colnames(mrna_mat))

lnc_mat  <- lnc_mat[, common_samples, drop = FALSE]
mrna_mat <- mrna_mat[, common_samples, drop = FALSE]

n_samples <- length(common_samples)
print(n_samples)

## optional sanity checks
head(colnames(lnc_mat))
head(colnames(mrna_mat))
dim(lnc_mat)
dim(mrna_mat)
identical(colnames(lnc_mat), colnames(mrna_mat))

## filter low-expression rows
min_count   <- 10
min_samples <- 2

filter_rows <- function(mat, min_count, min_samples) {
  keep <- rowSums(mat >= min_count) >= min_samples
  mat[keep, , drop = FALSE]
}

lnc_mat_f  <- filter_rows(lnc_mat,  min_count, min_samples)
mrna_mat_f <- filter_rows(mrna_mat, min_count, min_samples)

## log transform
lnc_log  <- log2(lnc_mat_f + 1)
mrna_log <- log2(mrna_mat_f + 1)

## pearson correlation
cor_mat <- cor(
  t(lnc_log),
  t(mrna_log),
  method = "pearson",
  use = "pairwise.complete.obs"
)
```

### Compute p-values and FDRs for all pairs

```{r}
# element-wise
# Handle edge case |r| == 1 to avoid division by zero
r_vals <- cor_mat

# Clip slightly if needed
r_vals[r_vals >  0.999999] <-  0.999999
r_vals[r_vals < -0.999999] <- -0.999999

t_vals <- r_vals * sqrt((n_samples - 2) / (1 - r_vals^2))
p_vals <- 2 * pt(-abs(t_vals), df = n_samples - 2)

#Now flatten matrices into a long table and do BH FDR correction
lnc_ids  <- rownames(cor_mat)
mrna_ids <- colnames(cor_mat)

coexp_df <- as.data.frame(as.table(cor_mat)) %>%
  as_tibble()

colnames(coexp_df) <- c("lnc_id", "mrna_id", "r")

# match p-values into the same long format
p_df <- as.data.frame(as.table(p_vals)) %>%
  as_tibble()

colnames(p_df) <- c("lnc_id", "mrna_id", "p_value")

coexp_df <- coexp_df %>%
  left_join(p_df, by = c("lnc_id", "mrna_id"))

# Global FDR across all pairs
coexp_df <- coexp_df %>%
  mutate(FDR = p.adjust(p_value, method = "BH"))


```

### Find significant pairs
```{r}
#Using common thresholds of significance
## |r| ≥ 0.8
## FDR < 0.05

r_threshold   <- 0.8
fdr_threshold <- 0.05

sig_pairs <- coexp_df %>%
  filter(abs(r) >= r_threshold,
         FDR <= fdr_threshold) %>%
  arrange(FDR, desc(abs(r)))

# Inspect top hits
print(sig_pairs %>% head())

```


# Co-expression summary statistics

## Output directory

```{r}
out_dir <- "~/github/oyster-lnc/output/15-pearsons"
dir.create(path.expand(out_dir), showWarnings = FALSE, recursive = TRUE)
```

## Summary statistics table

```{r}
# Total number of tested lncRNA–mRNA pairs
n_total_pairs <- nrow(coexp_df)

# Total significant pairs
n_sig_pairs <- nrow(sig_pairs)

# Unique lncRNAs and mRNAs represented in significant pairs
n_sig_lnc  <- sig_pairs %>% dplyr::distinct(lnc_id)  %>% nrow()
n_sig_mrna <- sig_pairs %>% dplyr::distinct(mrna_id) %>%
  nrow()

summary_stats <- tibble(
  metric = c(
    "Total pairs tested",
    "Significant pairs",
    "Unique lncRNAs in significant pairs",
    "Unique mRNAs in significant pairs"
  ),
  value = c(
    n_total_pairs,
    n_sig_pairs,
    n_sig_lnc,
    n_sig_mrna
  )
)

summary_stats
```

## Positive versus negative significant correlations

```{r}
direction_summary <- sig_pairs %>%
  dplyr::mutate(direction = dplyr::if_else(r > 0, "positive", "negative")) %>%
  dplyr::count(direction, name = "n_pairs") %>%
  dplyr::mutate(prop = n_pairs / sum(n_pairs))

direction_summary
```

## Top lncRNA hubs table

```{r}
lnc_hubs <- sig_pairs %>%
  dplyr::count(lnc_id, name = "n_correlated_mRNAs") %>%
  dplyr::arrange(dplyr::desc(n_correlated_mRNAs))

lnc_hubs %>% dplyr::slice_head(n = 20)
```

## Top mRNA hubs table

```{r}
mrna_hubs <- sig_pairs %>%
  dplyr::count(mrna_id, name = "n_correlated_lncRNAs") %>%
  dplyr::arrange(dplyr::desc(n_correlated_lncRNAs))

mrna_hubs %>% dplyr::slice_head(n = 20)
```

## lncRNA degree summary table


```{r}
lnc_degree_summary <- lnc_hubs %>%
  dplyr::summarise(
    min_degree = min(n_correlated_mRNAs),
    median_degree = median(n_correlated_mRNAs),
    mean_degree = mean(n_correlated_mRNAs),
    max_degree = max(n_correlated_mRNAs)
  ) %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "metric",
    values_to = "value"
  )

lnc_degree_summary
```

## mRNA degree summary table

```{r}
mrna_degree_summary <- mrna_hubs %>%
  dplyr::summarise(
    min_degree = min(n_correlated_lncRNAs),
    median_degree = median(n_correlated_lncRNAs),
    mean_degree = mean(n_correlated_lncRNAs),
    max_degree = max(n_correlated_lncRNAs)
  ) %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "metric",
    values_to = "value"
  )

mrna_degree_summary
```

## Optional compact narrative summary

```{r}
n_pos <- direction_summary %>%
  dplyr::filter(direction == "positive") %>%
  dplyr::pull(n_pairs)

n_neg <- direction_summary %>%
  dplyr::filter(direction == "negative") %>%
  dplyr::pull(n_pairs)

top_lnc <- lnc_hubs %>% dplyr::slice(1)
top_mrna <- mrna_hubs %>% dplyr::slice(1)

cat(
  "A total of", n_total_pairs, "lncRNA-mRNA pairs were tested.",
  n_sig_pairs, "pairs met the significance thresholds of |r| >=", r_threshold,
  "and FDR <=", fdr_threshold, ".",
  "These significant pairs involved", n_sig_lnc, "unique lncRNAs and",
  n_sig_mrna, "unique mRNAs.",
  n_pos, "significant pairs were positive and", n_neg, "were negative.",
  "The most connected lncRNA was", top_lnc$lnc_id,
  "with", top_lnc$n_correlated_mRNAs, "correlated mRNAs.",
  "The most connected mRNA was", top_mrna$mrna_id,
  "with", top_mrna$n_correlated_lncRNAs, "correlated lncRNAs."
)
```

## Export summary tables

```{r}
readr::write_tsv(summary_stats,
                 file.path(path.expand(out_dir), "coexpression_summary_stats.tsv"))

readr::write_tsv(direction_summary,
                 file.path(path.expand(out_dir), "coexpression_direction_summary.tsv"))

readr::write_tsv(lnc_hubs,
                 file.path(path.expand(out_dir), "lnc_hub_summary.tsv"))

readr::write_tsv(mrna_hubs,
                 file.path(path.expand(out_dir), "mrna_hub_summary.tsv"))

readr::write_tsv(lnc_degree_summary,
                 file.path(path.expand(out_dir), "lnc_degree_summary.tsv"))

readr::write_tsv(mrna_degree_summary,
                 file.path(path.expand(out_dir), "mrna_degree_summary.tsv"))
```

## Optional figure: lncRNA hub degree distribution

```{r}
ggplot(lnc_hubs, aes(x = n_correlated_mRNAs)) +
  geom_histogram(bins = 40) +
  theme_bw() +
  labs(
    x = "Number of correlated mRNAs per lncRNA",
    y = "Count of lncRNAs",
    title = "Distribution of lncRNA connectivity in the co-expression network"
  )
```

## Unique significant mRNAs

```{r}
unique_mrnas <- sig_pairs %>%
  dplyr::distinct(mrna_id) %>%
  dplyr::pull(mrna_id)

unique_mrnas

n_unique_mrnas <- sig_pairs %>%
  dplyr::distinct(mrna_id) %>%
  nrow()

n_unique_mrnas

unique_mrna_table <- sig_pairs %>%
  dplyr::distinct(mrna_id)

unique_mrna_table

readr::write_tsv(
  unique_mrna_table,
  "~/github/oyster-lnc/output/15-pearsons/unique_mrnas_in_sig_pairs.tsv"
)

mrna_degree <- sig_pairs %>%
  dplyr::count(mrna_id, name = "n_lncRNAs_correlated") %>%
  dplyr::arrange(desc(n_lncRNAs_correlated))

mrna_degree
```

```{r}
mrna_lnc_summary <- sig_pairs %>%
  dplyr::group_by(mrna_id) %>%
  dplyr::summarise(
    n_lncRNAs_correlated = dplyr::n_distinct(lnc_id),
    .groups = "drop"
  ) %>%
  dplyr::arrange(desc(n_lncRNAs_correlated))

mrna_lnc_summary

mrna_lnc_summary %>%
  dplyr::slice_head(n = 20)

readr::write_tsv(
  mrna_lnc_summary,
  "~/github/oyster-lnc/output/15-pearsons/mrna_lncRNA_degree_table.tsv"
)
```

```{r}
# =========================
# Install packages (run once)
# =========================
if (!requireNamespace("rentrez", quietly = TRUE)) {
  install.packages("rentrez")
}

if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

# =========================
# Load libraries
# =========================
library(rentrez)
library(tidyverse)

# helper for NULL values
`%||%` <- function(a, b) if (!is.null(a)) a else b

# =========================
# Read your gene list
# =========================
ids <- readr::read_tsv(
  "~/github/oyster-lnc/output/15-pearsons/justIDs.tsv",
  col_names = FALSE,
  show_col_types = FALSE
)[[1]]

# =========================
# Function to query NCBI
# =========================
search_one <- function(sym) {
  q <- paste0(sym, "[sym] AND Crassostrea virginica[orgn]")

  s <- rentrez::entrez_search(
    db = "gene",
    term = q,
    retmax = 1
  )

  if (length(s$ids) == 0) {
    return(tibble(
      input_id = sym,
      gene_id = NA_character_,
      symbol = NA_character_,
      description = NA_character_
    ))
  }

  sm <- rentrez::entrez_summary(
    db = "gene",
    id = s$ids[[1]]
  )

  tibble(
    input_id   = sym,
    gene_id    = as.character(sm$uid),
    symbol     = sm$name %||% NA_character_,
    description= sm$description %||% NA_character_
  )
}

# =========================
# Run batch annotation
# =========================
ann <- purrr::map_dfr(ids, search_one)

# =========================
# Save output
# =========================
readr::write_tsv(ann, "gene_annotations.tsv")

# view result
ann
```














