---
title: "15-pearsons"
author: "Zach Bengtsson"
date: "2025-11-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(tidyverse)
library(tibble)
library(igraph)
library(pheatmap)
library(RColorBrewer)
library(patchwork)
library(rtracklayer)
library(GenomicRanges)
```

```{r inputs}
data_dir <- "~/github/oyster-lnc/data/15-pearsons"
lnc_file  <- file.path(data_dir, "lnc_counts.tsv")
mrna_file <- "~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/kallisto_gene_counts.tsv"

# GFF/GFF3 files for in-notebook nearest-gene and cis classification
mrna_gff <- "~/github/oyster-lnc/data/11-mRNA-counts-DESeq2/C_virginica-3.0_Gnomon_mRNA.gff3"
lnc_gff  <- "~/github/oyster-lnc/data/14-comparative-BLASTs/GCF_002022765.2_C_virginica-3.0_lncRNA.gff"

# Output directory
out_dir <- "~/github/oyster-lnc/output/15-pearsons"
dir.create(path.expand(out_dir), showWarnings = FALSE, recursive = TRUE)

# Distance threshold for calling a pair cis
cis_window_bp <- 10000
```

```{r}
library(tidyverse)
library(tibble)

data_dir <- "~/github/oyster-lnc/data/15-pearsons"
lnc_file  <- file.path(data_dir, "lnc_counts.tsv")
mrna_file <- "~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/kallisto_gene_counts.tsv"

# GFF/GFF3 files for in-notebook nearest-gene and cis classification
mrna_gff <- "~/github/oyster-lnc/data/11-mRNA-counts-DESeq2/C_virginica-3.0_Gnomon_mRNA.gff3"
lnc_gff  <- "~/github/oyster-lnc/data/14-comparative-BLASTs/GCF_002022765.2_C_virginica-3.0_lncRNA.gff"

# Output directory
out_dir <- "~/github/oyster-lnc/output/15-pearsons"
dir.create(path.expand(out_dir), showWarnings = FALSE, recursive = TRUE)

# Distance threshold for calling a pair cis
cis_window_bp <- 10000

## load matrices
lnc_raw <- readr::read_tsv(lnc_file, show_col_types = FALSE)
colnames(lnc_raw)[1] <- "gene_id"

lnc_mat <- lnc_raw %>%
  tibble::column_to_rownames("gene_id") %>%
  as.matrix()

mrna_raw <- readr::read_tsv(mrna_file, show_col_types = FALSE)

mrna_mat <- mrna_raw %>%
  tibble::column_to_rownames("gene_id") %>%
  as.matrix()

## coerce to numeric
lnc_mat <- apply(lnc_mat, 2, as.numeric)
rownames(lnc_mat) <- lnc_raw$gene_id

mrna_mat <- apply(mrna_mat, 2, as.numeric)
rownames(mrna_mat) <- mrna_raw$gene_id

## align samples
common_samples <- intersect(colnames(lnc_mat), colnames(mrna_mat))

lnc_mat  <- lnc_mat[, common_samples, drop = FALSE]
mrna_mat <- mrna_mat[, common_samples, drop = FALSE]

n_samples <- length(common_samples)
print(n_samples)

## optional sanity checks
head(colnames(lnc_mat))
head(colnames(mrna_mat))
dim(lnc_mat)
dim(mrna_mat)
identical(colnames(lnc_mat), colnames(mrna_mat))

## filter low-expression rows
min_count   <- 10
min_samples <- 2

filter_rows <- function(mat, min_count, min_samples) {
  keep <- rowSums(mat >= min_count) >= min_samples
  mat[keep, , drop = FALSE]
}

lnc_mat_f  <- filter_rows(lnc_mat,  min_count, min_samples)
mrna_mat_f <- filter_rows(mrna_mat, min_count, min_samples)

## log transform
lnc_log  <- log2(lnc_mat_f + 1)
mrna_log <- log2(mrna_mat_f + 1)

## pearson correlation
cor_mat <- cor(
  t(lnc_log),
  t(mrna_log),
  method = "pearson",
  use = "pairwise.complete.obs"
)
```

### Compute p-values and FDRs for all pairs

```{r}
# element-wise
# Handle edge case |r| == 1 to avoid division by zero
r_vals <- cor_mat

# Clip slightly if needed
r_vals[r_vals >  0.999999] <-  0.999999
r_vals[r_vals < -0.999999] <- -0.999999

t_vals <- r_vals * sqrt((n_samples - 2) / (1 - r_vals^2))
p_vals <- 2 * pt(-abs(t_vals), df = n_samples - 2)

#Now flatten matrices into a long table and do BH FDR correction
lnc_ids  <- rownames(cor_mat)
mrna_ids <- colnames(cor_mat)

coexp_df <- as.data.frame(as.table(cor_mat)) %>%
  as_tibble()

colnames(coexp_df) <- c("lnc_id", "mrna_id", "r")

# match p-values into the same long format
p_df <- as.data.frame(as.table(p_vals)) %>%
  as_tibble()

colnames(p_df) <- c("lnc_id", "mrna_id", "p_value")

coexp_df <- coexp_df %>%
  left_join(p_df, by = c("lnc_id", "mrna_id"))

# Global FDR across all pairs
coexp_df <- coexp_df %>%
  mutate(FDR = p.adjust(p_value, method = "BH"))


```

### Find significant pairs

```{r significant-pairs}
r_threshold   <- 0.8
fdr_threshold <- 0.05

sig_pairs <- coexp_df %>%
  filter(
    abs(r) >= r_threshold,
    FDR <= fdr_threshold
  ) %>%
  arrange(FDR, desc(abs(r))) %>%
  mutate(direction = if_else(r > 0, "positive", "negative"))

print(sig_pairs %>% head())
```


Old version that does not include positive or negative
```{r}
#Using common thresholds of significance
## |r| ≥ 0.8
## FDR < 0.05

r_threshold   <- 0.8
fdr_threshold <- 0.05

sig_pairs <- coexp_df %>%
  filter(abs(r) >= r_threshold,
         FDR <= fdr_threshold) %>%
  arrange(FDR, desc(abs(r)))

# Inspect top hits
print(sig_pairs %>% head())

```

## Export main pair table

```{r export-main-table}
readr::write_tsv(
  sig_pairs,
  file.path(path.expand(out_dir), "significant_lnc_mrna_pairs.tsv")
)
```

# Co-expression summary statistics

## Output directory

```{r}
out_dir <- "~/github/oyster-lnc/output/15-pearsons"
dir.create(path.expand(out_dir), showWarnings = FALSE, recursive = TRUE)
```

## Summary statistics table

```{r}

# Total number of tested lncRNA–mRNA pairs
n_total_pairs <- nrow(coexp_df)

# Total significant pairs
n_sig_pairs <- nrow(sig_pairs)

# Unique lncRNAs and mRNAs represented in significant pairs
n_sig_lnc  <- sig_pairs %>% dplyr::distinct(lnc_id)  %>% nrow()
n_sig_mrna <- sig_pairs %>% dplyr::distinct(mrna_id) %>%
  nrow()

summary_stats <- tibble(
  metric = c(
    "Total pairs tested",
    "Significant pairs",
    "Unique lncRNAs in significant pairs",
    "Unique mRNAs in significant pairs"
  ),
  value = c(
    n_total_pairs,
    n_sig_pairs,
    n_sig_lnc,
    n_sig_mrna
  )
)

summary_stats
```

## Positive versus negative significant correlations

```{r}
direction_summary <- sig_pairs %>%
  dplyr::mutate(direction = dplyr::if_else(r > 0, "positive", "negative")) %>%
  dplyr::count(direction, name = "n_pairs") %>%
  dplyr::mutate(prop = n_pairs / sum(n_pairs))

direction_summary
```

## Top lncRNA hubs table

```{r}
lnc_hubs <- sig_pairs %>%
  dplyr::count(lnc_id, name = "n_correlated_mRNAs") %>%
  dplyr::arrange(dplyr::desc(n_correlated_mRNAs))

lnc_hubs %>% dplyr::slice_head(n = 20)
```

## Top mRNA hubs table

```{r}
mrna_hubs <- sig_pairs %>%
  dplyr::count(mrna_id, name = "n_correlated_lncRNAs") %>%
  dplyr::arrange(dplyr::desc(n_correlated_lncRNAs))

mrna_hubs %>% dplyr::slice_head(n = 20)
```

## lncRNA degree summary table


```{r}
lnc_degree_summary <- lnc_hubs %>%
  dplyr::summarise(
    min_degree = min(n_correlated_mRNAs),
    median_degree = median(n_correlated_mRNAs),
    mean_degree = mean(n_correlated_mRNAs),
    max_degree = max(n_correlated_mRNAs)
  ) %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "metric",
    values_to = "value"
  )

lnc_degree_summary
```

## mRNA degree summary table

```{r}
mrna_degree_summary <- mrna_hubs %>%
  dplyr::summarise(
    min_degree = min(n_correlated_lncRNAs),
    median_degree = median(n_correlated_lncRNAs),
    mean_degree = mean(n_correlated_lncRNAs),
    max_degree = max(n_correlated_lncRNAs)
  ) %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "metric",
    values_to = "value"
  )

mrna_degree_summary
```

## Optional compact narrative summary

```{r}
n_pos <- direction_summary %>%
  dplyr::filter(direction == "positive") %>%
  dplyr::pull(n_pairs)

n_neg <- direction_summary %>%
  dplyr::filter(direction == "negative") %>%
  dplyr::pull(n_pairs)

top_lnc <- lnc_hubs %>% dplyr::slice(1)
top_mrna <- mrna_hubs %>% dplyr::slice(1)

cat(
  "A total of", n_total_pairs, "lncRNA-mRNA pairs were tested.",
  n_sig_pairs, "pairs met the significance thresholds of |r| >=", r_threshold,
  "and FDR <=", fdr_threshold, ".",
  "These significant pairs involved", n_sig_lnc, "unique lncRNAs and",
  n_sig_mrna, "unique mRNAs.",
  n_pos, "significant pairs were positive and", n_neg, "were negative.",
  "The most connected lncRNA was", top_lnc$lnc_id,
  "with", top_lnc$n_correlated_mRNAs, "correlated mRNAs.",
  "The most connected mRNA was", top_mrna$mrna_id,
  "with", top_mrna$n_correlated_lncRNAs, "correlated lncRNAs."
)
```

## Export summary tables

```{r}
readr::write_tsv(summary_stats,
                 file.path(path.expand(out_dir), "coexpression_summary_stats.tsv"))

readr::write_tsv(direction_summary,
                 file.path(path.expand(out_dir), "coexpression_direction_summary.tsv"))

readr::write_tsv(lnc_hubs,
                 file.path(path.expand(out_dir), "lnc_hub_summary.tsv"))

readr::write_tsv(mrna_hubs,
                 file.path(path.expand(out_dir), "mrna_hub_summary.tsv"))

readr::write_tsv(lnc_degree_summary,
                 file.path(path.expand(out_dir), "lnc_degree_summary.tsv"))

readr::write_tsv(mrna_degree_summary,
                 file.path(path.expand(out_dir), "mrna_degree_summary.tsv"))
```

## Optional figure: lncRNA hub degree distribution

```{r}
ggplot(lnc_hubs, aes(x = n_correlated_mRNAs)) +
  geom_histogram(bins = 40) +
  theme_bw() +
  labs(
    x = "Number of correlated mRNAs per lncRNA",
    y = "Count of lncRNAs",
    title = "Distribution of lncRNA connectivity in the co-expression network"
  )
```

## Unique significant mRNAs

```{r}
unique_mrnas <- sig_pairs %>%
  dplyr::distinct(mrna_id) %>%
  dplyr::pull(mrna_id)

unique_mrnas

n_unique_mrnas <- sig_pairs %>%
  dplyr::distinct(mrna_id) %>%
  nrow()

n_unique_mrnas

unique_mrna_table <- sig_pairs %>%
  dplyr::distinct(mrna_id)

unique_mrna_table

readr::write_tsv(
  unique_mrna_table,
  "~/github/oyster-lnc/output/15-pearsons/unique_mrnas_in_sig_pairs.tsv"
)

mrna_degree <- sig_pairs %>%
  dplyr::count(mrna_id, name = "n_lncRNAs_correlated") %>%
  dplyr::arrange(desc(n_lncRNAs_correlated))

mrna_degree
```

```{r}
mrna_lnc_summary <- sig_pairs %>%
  dplyr::group_by(mrna_id) %>%
  dplyr::summarise(
    n_lncRNAs_correlated = dplyr::n_distinct(lnc_id),
    .groups = "drop"
  ) %>%
  dplyr::arrange(desc(n_lncRNAs_correlated))

mrna_lnc_summary

mrna_lnc_summary %>%
  dplyr::slice_head(n = 20)

readr::write_tsv(
  mrna_lnc_summary,
  "~/github/oyster-lnc/output/15-pearsons/mrna_lncRNA_degree_table.tsv"
)
```

# Visualizations

## Visualization 1: Correlation 

```{r}
sig_pairs <- sig_pairs %>%
  mutate(direction = if_else(r > 0, "positive", "negative"))
```


```{r correlation-distribution}
p_hist <- ggplot(sig_pairs, aes(x = r, fill = direction)) +
  geom_histogram(bins = 50, color = "white", linewidth = 0.2) +
  scale_fill_manual(values = c("positive" = "#E63946", "negative" = "#457B9D")) +
  labs(
    title = "Distribution of Significant lncRNA–mRNA Correlations",
    subtitle = paste0("|r| ≥ ", r_threshold, ", FDR ≤ ", fdr_threshold),
    x = "Pearson r",
    y = "Count",
    fill = "Direction"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

print(p_hist)

ggsave(file.path(path.expand(out_dir), "correlation_distribution.png"),
       p_hist, width = 8, height = 5, dpi = 300)
```

## Visualization 2: Top hub lncRNAs

```{r top-hub-lncrnas}
top_hubs <- lnc_hubs %>% slice_head(n = 20)

p_hubs <- ggplot(top_hubs, aes(x = reorder(lnc_id, n_correlated_mRNAs),
                               y = n_correlated_mRNAs)) +
  geom_col(fill = "#E63946", alpha = 0.85) +
  coord_flip() +
  labs(
    title = "Top 20 Hub lncRNAs by Number of Correlated mRNAs",
    x = NULL,
    y = "Number of significantly correlated mRNAs"
  ) +
  theme_minimal(base_size = 12)

print(p_hubs)

ggsave(file.path(path.expand(out_dir), "hub_lncrnas.png"),
       p_hubs, width = 10, height = 7, dpi = 300)
```

## Visualization 3: Degree histogram for lncRNA hubs

```{r lnc-degree-histogram}
p_degree <- ggplot(lnc_hubs, aes(x = n_correlated_mRNAs)) +
  geom_histogram(bins = 40, fill = "#E63946", color = "white") +
  theme_bw() +
  labs(
    x = "Number of correlated mRNAs per lncRNA",
    y = "Count of lncRNAs",
    title = "Distribution of lncRNA connectivity in the co-expression network"
  )

print(p_degree)

ggsave(file.path(path.expand(out_dir), "lncrna_degree_distribution.png"),
       p_degree, width = 8, height = 5, dpi = 300)
```

## Visualization 4: Network of all significant interactions

```{r network-plot, fig.width=14, fig.height=14}
edges <- sig_pairs %>%
  select(lnc_id, mrna_id, r, FDR, direction)

g <- graph_from_data_frame(edges, directed = FALSE)

node_names <- V(g)$name
node_type  <- ifelse(node_names %in% sig_pairs$lnc_id, "lncRNA", "mRNA")
V(g)$type  <- node_type

V(g)$color <- ifelse(node_type == "lncRNA", "#E63946", "#457B9D")
V(g)$frame.color <- ifelse(node_type == "lncRNA", "#9D0208", "#1D3557")
V(g)$size <- ifelse(
  node_type == "lncRNA",
  pmin(degree(g) * 1.2 + 4, 20),
  pmin(degree(g) * 0.8 + 3, 15)
)
V(g)$label.cex <- 0.35
V(g)$label.color <- "gray20"

E(g)$color <- ifelse(
  edges$direction == "positive",
  adjustcolor("#E63946", alpha.f = 0.3),
  adjustcolor("#457B9D", alpha.f = 0.3)
)
E(g)$width <- abs(edges$r) * 2

set.seed(42)
layout <- layout_with_fr(g)

png(file.path(path.expand(out_dir), "coexpression_network.png"),
    width = 2400, height = 2400, res = 200)
par(mar = c(1, 1, 3, 1))
plot(
  g,
  layout = layout,
  vertex.label = NA,
  edge.curved = 0.1,
  main = "lncRNA–mRNA Co-expression Network"
)
legend(
  "bottomleft",
  legend = c("lncRNA", "mRNA", "Positive r", "Negative r"),
  col = c("#E63946", "#457B9D", "#E63946", "#457B9D"),
  pch = c(16, 16, NA, NA),
  lty = c(NA, NA, 1, 1),
  lwd = 2,
  cex = 0.9,
  bty = "n"
)
dev.off()

par(mar = c(1, 1, 3, 1))
plot(
  g,
  layout = layout,
  vertex.label = NA,
  edge.curved = 0.1,
  main = "lncRNA–mRNA Co-expression Network"
)
legend(
  "bottomleft",
  legend = c("lncRNA", "mRNA", "Positive r", "Negative r"),
  col = c("#E63946", "#457B9D", "#E63946", "#457B9D"),
  pch = c(16, 16, NA, NA),
  lty = c(NA, NA, 1, 1),
  lwd = 2,
  cex = 0.9,
  bty = "n"
)
```

## Visualization 5: Heatmap of top correlations

```{r heatmap, fig.width=12, fig.height=10}
top_lnc <- lnc_hubs %>% slice_head(n = min(15, n())) %>% pull(lnc_id)

top_mrna <- sig_pairs %>%
  filter(lnc_id %in% top_lnc) %>%
  count(mrna_id, name = "n_lnc") %>%
  arrange(desc(n_lnc)) %>%
  slice_head(n = min(40, n())) %>%
  pull(mrna_id)

heatmap_data <- cor_mat[as.character(top_lnc), as.character(top_mrna), drop = FALSE]

png(file.path(path.expand(out_dir), "correlation_heatmap.png"),
    width = 2000, height = 1600, res = 200)
pheatmap(
  heatmap_data,
  color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
  breaks = seq(-1, 1, length.out = 101),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  fontsize_row = 7,
  fontsize_col = 6,
  main = "Pearson Correlation: Top Hub lncRNAs vs Associated mRNAs",
  angle_col = 45
)
dev.off()

pheatmap(
  heatmap_data,
  color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
  breaks = seq(-1, 1, length.out = 101),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  fontsize_row = 7,
  fontsize_col = 6,
  main = "Pearson Correlation: Top Hub lncRNAs vs Associated mRNAs",
  angle_col = 45
)
```

## Visualization 6: Individual scatter plots for selected pairs

```{r scatter-function}
plot_pair_raw <- function(
  lnc_id_sel,
  mrna_id_sel,
  lnc_mat_source = lnc_mat_f,
  mrna_mat_source = mrna_mat_f,
  samples = common_samples
) {
  df <- tibble(
    sample = samples,
    lnc_expr  = as.numeric(lnc_mat_source[lnc_id_sel, samples]),
    mrna_expr = as.numeric(mrna_mat_source[mrna_id_sel, samples])
  )

  r_val <- cor(df$lnc_expr, df$mrna_expr, method = "pearson")

  ggplot(df, aes(x = lnc_expr, y = mrna_expr)) +
    geom_point(size = 3, alpha = 0.75, color = "#264653") +
    geom_smooth(method = "lm", se = TRUE, color = "#E76F51",
                fill = "#E76F51", alpha = 0.15) +
    geom_text(aes(label = sample), size = 2.5,
              nudge_y = max(df$mrna_expr) * 0.03,
              color = "gray40") +
    labs(
      title = paste0(lnc_id_sel, " vs ", mrna_id_sel),
      subtitle = sprintf("Pearson r = %.3f (raw counts, n = %d samples)",
                         r_val, nrow(df)),
      x = paste("lncRNA expression:", lnc_id_sel),
      y = paste("mRNA expression:", mrna_id_sel)
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(size = 11))
}
```

## Top 12 most significant pairs

```{r top12-scatter, fig.width=12, fig.height=16}
top12 <- sig_pairs %>% slice_head(n = min(12, n()))

scatter_list <- pmap(top12, function(lnc_id, mrna_id, ...) {
  plot_pair_raw(lnc_id, mrna_id)
})

combined <- wrap_plots(scatter_list, ncol = 3)
print(combined)

ggsave(file.path(path.expand(out_dir), "top12_scatter_raw.png"),
       combined, width = 16, height = 20, dpi = 300)
```

## Save individual scatter plots for up to 100 significant pairs

```{r save-scatter-plots}
scatter_dir <- file.path(path.expand(out_dir), "scatter_plots")
dir.create(scatter_dir, recursive = TRUE, showWarnings = FALSE)

n_to_save <- min(nrow(sig_pairs), 100)
cat("Saving", n_to_save, "individual scatter plots...\n")

for (i in seq_len(n_to_save)) {
  row <- sig_pairs[i, ]
  p <- plot_pair_raw(row$lnc_id, row$mrna_id)

  fname <- paste0(
    gsub("[^A-Za-z0-9_.-]", "_", row$lnc_id),
    "_vs_",
    gsub("[^A-Za-z0-9_.-]", "_", row$mrna_id),
    ".png"
  )

  ggsave(file.path(scatter_dir, fname), p, width = 7, height = 5, dpi = 200)
}

cat("Scatter plots saved to:", scatter_dir, "\n")
```

## Unique significant mRNAs

```{r unique-significant-mrnas}
unique_mrnas <- sig_pairs %>%
  distinct(mrna_id) %>%
  pull(mrna_id)

unique_mrnas

n_unique_mrnas <- sig_pairs %>%
  distinct(mrna_id) %>%
  nrow()

n_unique_mrnas

unique_mrna_table <- sig_pairs %>%
  distinct(mrna_id)

unique_mrna_table

readr::write_tsv(
  unique_mrna_table,
  file.path(path.expand(out_dir), "unique_mrnas_in_sig_pairs.tsv")
)
```

## mRNA degree table

```{r mrna-degree-table}
mrna_lnc_summary <- sig_pairs %>%
  group_by(mrna_id) %>%
  summarise(
    n_lncRNAs_correlated = n_distinct(lnc_id),
    .groups = "drop"
  ) %>%
  arrange(desc(n_lncRNAs_correlated))

mrna_lnc_summary

mrna_lnc_summary %>%
  slice_head(n = 20)

readr::write_tsv(
  mrna_lnc_summary,
  file.path(path.expand(out_dir), "mrna_lncRNA_degree_table.tsv")
)
```

## Session info

```{r session-info}
sessionInfo()
```






```{r}
# =========================
# Install packages (run once)
# =========================
if (!requireNamespace("rentrez", quietly = TRUE)) {
  install.packages("rentrez")
}

if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

# =========================
# Load libraries
# =========================
library(rentrez)
library(tidyverse)

# helper for NULL values
`%||%` <- function(a, b) if (!is.null(a)) a else b

# =========================
# Read your gene list
# =========================
ids <- readr::read_tsv(
  "~/github/oyster-lnc/output/15-pearsons/justIDs.tsv",
  col_names = FALSE,
  show_col_types = FALSE
)[[1]]

# =========================
# Function to query NCBI
# =========================
search_one <- function(sym) {
  q <- paste0(sym, "[sym] AND Crassostrea virginica[orgn]")

  s <- rentrez::entrez_search(
    db = "gene",
    term = q,
    retmax = 1
  )

  if (length(s$ids) == 0) {
    return(tibble(
      input_id = sym,
      gene_id = NA_character_,
      symbol = NA_character_,
      description = NA_character_
    ))
  }

  sm <- rentrez::entrez_summary(
    db = "gene",
    id = s$ids[[1]]
  )

  tibble(
    input_id   = sym,
    gene_id    = as.character(sm$uid),
    symbol     = sm$name %||% NA_character_,
    description= sm$description %||% NA_character_
  )
}

# =========================
# Run batch annotation
# =========================
ann <- purrr::map_dfr(ids, search_one)

# =========================
# Save output
# =========================
readr::write_tsv(ann, "gene_annotations.tsv")

# view result
ann
```














