---
title: "16-BRH-blast-comparison"
author: "Zach Bengtsson"
date: "2026-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Best reciprocal hit (BRH) analysis for *Crassostrea virginica* lncRNAs

This notebook identifies best reciprocal hit (BRH) relationships between *Crassostrea virginica* lncRNAs and lncRNA datasets from other bivalve species using relaxed similarity thresholds appropriate for lncRNAs. It uses explicit BLAST executable paths for reproducibility on HPC systems.

## Libraries and directories

```{r}
library(tidyverse)
library(Biostrings)
library(purrr)

data_dir  <- "~/github/oyster-lnc/data/14-comparative-BLASTs"
blast_dir <- "~/github/oyster-lnc/output/16-BRH-blast-comparison/blast"
out_dir   <- "~/github/oyster-lnc/output/16-BRH-blast-comparison/tables"
fasta_dir <- "~/github/oyster-lnc/output/16-BRH-blast-comparison/fastas"

dir.create(path.expand(blast_dir), showWarnings = FALSE, recursive = TRUE)
dir.create(path.expand(out_dir),   showWarnings = FALSE, recursive = TRUE)
dir.create(path.expand(fasta_dir), showWarnings = FALSE, recursive = TRUE)
```

## Species metadata

```{r}
species_meta <- tibble::tribble(
  ~species,           ~label,                     ~fasta,
  "Cvir",             "Crassostrea virginica",    "~/github/oyster-lnc/data/14-comparative-BLASTs/GCF_002022765.2_C_virginica-3.0_lncRNA.fa",
  "Cgigas_roslin",    "Crassostrea gigas",        "~/github/oyster-lnc/data/14-comparative-BLASTs/Cgigas_lncRNA.fasta",
  "Mgigas_new",       "Magallana gigas",          "~/github/oyster-lnc/data/14-comparative-BLASTs/Magallana_lncRNA.fasta",
  "Myessoensis",      "Mizuhopecten yessoensis",  "~/github/oyster-lnc/data/14-comparative-BLASTs/Myessoensis_lncRNA.fasta",
  "Rphilippinarum",   "Ruditapes philippinarum",  "~/github/oyster-lnc/data/14-comparative-BLASTs/Rphilippanarum_lncRNA.fasta",
  "Mtrossulus",       "Mytilus trossulus",        "~/github/oyster-lnc/data/14-comparative-BLASTs/Mtrossulus_lncRNA.fasta"
) %>%
  dplyr::mutate(fasta = path.expand(fasta))

species_meta
```

## Relaxed BRH thresholds

```{r}
PIDENT_MIN <- 50
QCOV_MIN   <- 0.30
EVALUE_MAX <- 1e-5
```

## Helper functions

### Read FASTA lengths

```{r}
get_lnc_lengths <- function(fasta_path) {
  dna <- Biostrings::readDNAStringSet(fasta_path)

  tibble(
    seq_id    = names(dna),
    length_nt = as.integer(width(dna))
  )
}
```

### Read FASTA with names preserved

```{r}
read_lnc_fasta <- function(fasta_path) {
  Biostrings::readDNAStringSet(fasta_path)
}
```

### Parse BLAST output and retain best hit per query

```{r}
read_blast_best_hits <- function(path,
                                 pident_min = PIDENT_MIN,
                                 qcov_min   = QCOV_MIN,
                                 evalue_max = EVALUE_MAX) {
  empty <- tibble(
    qseqid   = character(),
    sseqid   = character(),
    pident   = numeric(),
    aln_len  = integer(),
    qlen     = integer(),
    slen     = integer(),
    evalue   = numeric(),
    bitscore = numeric(),
    qcov     = numeric()
  )

  if (!file.exists(path)) {
    warning("BLAST file not found: ", path)
    return(empty)
  }

  x <- readr::read_tsv(
    path,
    col_names = c("qseqid", "sseqid", "pident",
                  "aln_len", "qlen", "slen", "evalue", "bitscore"),
    show_col_types = FALSE
  )

  if (nrow(x) == 0) {
    warning("BLAST file is empty: ", path)
    return(empty)
  }

  x %>%
    dplyr::mutate(qcov = aln_len / qlen) %>%
    dplyr::filter(
      pident >= pident_min,
      qcov   >= qcov_min,
      evalue <= evalue_max
    ) %>%
    dplyr::group_by(qseqid) %>%
    dplyr::arrange(
      dplyr::desc(bitscore),
      evalue,
      dplyr::desc(qcov),
      .by_group = TRUE
    ) %>%
    dplyr::slice(1) %>%
    dplyr::ungroup()
}
```

### Identify best reciprocal hit pairs

```{r}
get_brh_pairs <- function(q_to_s, s_to_q, q_species, s_species) {
  dplyr::inner_join(
    q_to_s %>% dplyr::select(qseqid, sseqid),
    s_to_q %>% dplyr::select(qseqid, sseqid),
    by = c("qseqid" = "sseqid", "sseqid" = "qseqid")
  ) %>%
    dplyr::transmute(
      !!q_species := qseqid,
      !!s_species := sseqid
    )
}
```

### Enforce unique 1:1 relationships

```{r}
enforce_one_to_one <- function(df, col1, col2) {
  col1 <- rlang::ensym(col1)
  col2 <- rlang::ensym(col2)

  df %>%
    dplyr::add_count(!!col1, name = "n1") %>%
    dplyr::add_count(!!col2, name = "n2") %>%
    dplyr::filter(n1 == 1, n2 == 1) %>%
    dplyr::select(-n1, -n2)
}
```

### Write FASTA subsets

```{r}
write_subset_fasta <- function(dna, ids, outfile, prefix = NULL) {
  keep <- ids[ids %in% names(dna)]
  x <- dna[keep]

  if (!is.null(prefix)) {
    names(x) <- paste0(prefix, "_", names(x))
  }

  Biostrings::writeXStringSet(x, outfile)
}
```

## Check BLAST installation

```{bash}
BLAST=/home/shared/ncbi-blast-2.11.0+/bin
"$BLAST/blastn" -version
"$BLAST/makeblastdb" -version
```

## Build BLAST databases

```{bash}
BLAST=/home/shared/ncbi-blast-2.11.0+/bin
DATA_DIR=~/github/oyster-lnc/data/14-comparative-BLASTs
BLAST_DIR=~/github/oyster-lnc/output/16-BRH-blast-comparison/blast

mkdir -p "$BLAST_DIR"

"$BLAST/makeblastdb" \
  -in "$DATA_DIR/GCF_002022765.2_C_virginica-3.0_lncRNA.fa" \
  -dbtype nucl \
  -out "$BLAST_DIR/Cvir_db"

"$BLAST/makeblastdb" \
  -in "$DATA_DIR/Cgigas_lncRNA.fasta" \
  -dbtype nucl \
  -out "$BLAST_DIR/Cgigas_roslin_db"

"$BLAST/makeblastdb" \
  -in "$DATA_DIR/Magallana_lncRNA.fasta" \
  -dbtype nucl \
  -out "$BLAST_DIR/Mgigas_new_db"

"$BLAST/makeblastdb" \
  -in "$DATA_DIR/Myessoensis_lncRNA.fasta" \
  -dbtype nucl \
  -out "$BLAST_DIR/Myessoensis_db"

"$BLAST/makeblastdb" \
  -in "$DATA_DIR/Rphilippanarum_lncRNA.fasta" \
  -dbtype nucl \
  -out "$BLAST_DIR/Rphilippinarum_db"

"$BLAST/makeblastdb" \
  -in "$DATA_DIR/Mtrossulus_lncRNA.fasta" \
  -dbtype nucl \
  -out "$BLAST_DIR/Mtrossulus_db"
```

## Run pairwise reciprocal BLASTn comparisons

```{bash}
BLAST=/home/shared/ncbi-blast-2.11.0+/bin
DATA_DIR=~/github/oyster-lnc/data/14-comparative-BLASTs
BLAST_DIR=~/github/oyster-lnc/output/16-BRH-blast-comparison/blast

mkdir -p "$BLAST_DIR"

# C. virginica as query against each comparison species
"$BLAST/blastn" \
  -query "$DATA_DIR/GCF_002022765.2_C_virginica-3.0_lncRNA.fa" \
  -db "$BLAST_DIR/Cgigas_roslin_db" \
  -evalue 1e-5 \
  -max_target_seqs 5 \
  -outfmt "6 qseqid sseqid pident length qlen slen evalue bitscore" \
  > "$BLAST_DIR/Cvir_vs_Cgigas_roslin.blastn.tsv"

"$BLAST/blastn" \
  -query "$DATA_DIR/GCF_002022765.2_C_virginica-3.0_lncRNA.fa" \
  -db "$BLAST_DIR/Mgigas_new_db" \
  -evalue 1e-5 \
  -max_target_seqs 5 \
  -outfmt "6 qseqid sseqid pident length qlen slen evalue bitscore" \
  > "$BLAST_DIR/Cvir_vs_Mgigas_new.blastn.tsv"

"$BLAST/blastn" \
  -query "$DATA_DIR/GCF_002022765.2_C_virginica-3.0_lncRNA.fa" \
  -db "$BLAST_DIR/Myessoensis_db" \
  -evalue 1e-5 \
  -max_target_seqs 5 \
  -outfmt "6 qseqid sseqid pident length qlen slen evalue bitscore" \
  > "$BLAST_DIR/Cvir_vs_Myessoensis.blastn.tsv"

"$BLAST/blastn" \
  -query "$DATA_DIR/GCF_002022765.2_C_virginica-3.0_lncRNA.fa" \
  -db "$BLAST_DIR/Rphilippinarum_db" \
  -evalue 1e-5 \
  -max_target_seqs 5 \
  -outfmt "6 qseqid sseqid pident length qlen slen evalue bitscore" \
  > "$BLAST_DIR/Cvir_vs_Rphilippinarum.blastn.tsv"

"$BLAST/blastn" \
  -query "$DATA_DIR/GCF_002022765.2_C_virginica-3.0_lncRNA.fa" \
  -db "$BLAST_DIR/Mtrossulus_db" \
  -evalue 1e-5 \
  -max_target_seqs 5 \
  -outfmt "6 qseqid sseqid pident length qlen slen evalue bitscore" \
  > "$BLAST_DIR/Cvir_vs_Mtrossulus.blastn.tsv"

# Reciprocal searches back to C. virginica
"$BLAST/blastn" \
  -query "$DATA_DIR/Cgigas_lncRNA.fasta" \
  -db "$BLAST_DIR/Cvir_db" \
  -evalue 1e-5 \
  -max_target_seqs 5 \
  -outfmt "6 qseqid sseqid pident length qlen slen evalue bitscore" \
  > "$BLAST_DIR/Cgigas_roslin_vs_Cvir.blastn.tsv"

"$BLAST/blastn" \
  -query "$DATA_DIR/Magallana_lncRNA.fasta" \
  -db "$BLAST_DIR/Cvir_db" \
  -evalue 1e-5 \
  -max_target_seqs 5 \
  -outfmt "6 qseqid sseqid pident length qlen slen evalue bitscore" \
  > "$BLAST_DIR/Mgigas_new_vs_Cvir.blastn.tsv"

"$BLAST/blastn" \
  -query "$DATA_DIR/Myessoensis_lncRNA.fasta" \
  -db "$BLAST_DIR/Cvir_db" \
  -evalue 1e-5 \
  -max_target_seqs 5 \
  -outfmt "6 qseqid sseqid pident length qlen slen evalue bitscore" \
  > "$BLAST_DIR/Myessoensis_vs_Cvir.blastn.tsv"

"$BLAST/blastn" \
  -query "$DATA_DIR/Rphilippanarum_lncRNA.fasta" \
  -db "$BLAST_DIR/Cvir_db" \
  -evalue 1e-5 \
  -max_target_seqs 5 \
  -outfmt "6 qseqid sseqid pident length qlen slen evalue bitscore" \
  > "$BLAST_DIR/Rphilippinarum_vs_Cvir.blastn.tsv"

"$BLAST/blastn" \
  -query "$DATA_DIR/Mtrossulus_lncRNA.fasta" \
  -db "$BLAST_DIR/Cvir_db" \
  -evalue 1e-5 \
  -max_target_seqs 5 \
  -outfmt "6 qseqid sseqid pident length qlen slen evalue bitscore" \
  > "$BLAST_DIR/Mtrossulus_vs_Cvir.blastn.tsv"
```

## Read BLAST tables and retain best hits

```{r}
cvir_vs_cgigas1 <- read_blast_best_hits(file.path(path.expand(blast_dir), "Cvir_vs_Cgigas_roslin.blastn.tsv"))
cgigas1_vs_cvir <- read_blast_best_hits(file.path(path.expand(blast_dir), "Cgigas_roslin_vs_Cvir.blastn.tsv"))

cvir_vs_cgigas2 <- read_blast_best_hits(file.path(path.expand(blast_dir), "Cvir_vs_Mgigas_new.blastn.tsv"))
cgigas2_vs_cvir <- read_blast_best_hits(file.path(path.expand(blast_dir), "Mgigas_new_vs_Cvir.blastn.tsv"))

cvir_vs_myes <- read_blast_best_hits(file.path(path.expand(blast_dir), "Cvir_vs_Myessoensis.blastn.tsv"))
myes_vs_cvir <- read_blast_best_hits(file.path(path.expand(blast_dir), "Myessoensis_vs_Cvir.blastn.tsv"))

cvir_vs_rphi <- read_blast_best_hits(file.path(path.expand(blast_dir), "Cvir_vs_Rphilippinarum.blastn.tsv"))
rphi_vs_cvir <- read_blast_best_hits(file.path(path.expand(blast_dir), "Rphilippinarum_vs_Cvir.blastn.tsv"))

cvir_vs_mtro <- read_blast_best_hits(file.path(path.expand(blast_dir), "Cvir_vs_Mtrossulus.blastn.tsv"))
mtro_vs_cvir <- read_blast_best_hits(file.path(path.expand(blast_dir), "Mtrossulus_vs_Cvir.blastn.tsv"))
```

## Build BRH tables

```{r}
brh_Cvir_Cgigas_roslin  <- get_brh_pairs(cvir_vs_cgigas1, cgigas1_vs_cvir, "Cvir", "Cgigas_roslin")
brh_Cvir_Mgigas_new     <- get_brh_pairs(cvir_vs_cgigas2, cgigas2_vs_cvir, "Cvir", "Mgigas_new")
brh_Cvir_Myessoensis    <- get_brh_pairs(cvir_vs_myes, myes_vs_cvir, "Cvir", "Myessoensis")
brh_Cvir_Rphilippinarum <- get_brh_pairs(cvir_vs_rphi, rphi_vs_cvir, "Cvir", "Rphilippinarum")
brh_Cvir_Mtrossulus     <- get_brh_pairs(cvir_vs_mtro, mtro_vs_cvir, "Cvir", "Mtrossulus")
```

## Enforce unique 1:1 BRH relationships

```{r}
brh_Cvir_Cgigas_roslin_1to1  <- enforce_one_to_one(brh_Cvir_Cgigas_roslin,  Cvir, Cgigas_roslin)
brh_Cvir_Mgigas_new_1to1     <- enforce_one_to_one(brh_Cvir_Mgigas_new,     Cvir, Mgigas_new)
brh_Cvir_Myessoensis_1to1    <- enforce_one_to_one(brh_Cvir_Myessoensis,    Cvir, Myessoensis)
brh_Cvir_Rphilippinarum_1to1 <- enforce_one_to_one(brh_Cvir_Rphilippinarum, Cvir, Rphilippinarum)
brh_Cvir_Mtrossulus_1to1     <- enforce_one_to_one(brh_Cvir_Mtrossulus,     Cvir, Mtrossulus)
```

## Summary table: number of BRH hits with *C. virginica* by species

```{r}
brh_count_table <- tibble::tribble(
  ~comparison_species,        ~n_brh_pairs,                     ~n_brh_pairs_1to1,                     ~n_unique_Cvir_lncRNAs,
  "Crassostrea gigas",        nrow(brh_Cvir_Cgigas_roslin),     nrow(brh_Cvir_Cgigas_roslin_1to1),     dplyr::n_distinct(brh_Cvir_Cgigas_roslin_1to1$Cvir),
  "Magallana gigas",          nrow(brh_Cvir_Mgigas_new),        nrow(brh_Cvir_Mgigas_new_1to1),        dplyr::n_distinct(brh_Cvir_Mgigas_new_1to1$Cvir),
  "Mizuhopecten yessoensis",  nrow(brh_Cvir_Myessoensis),       nrow(brh_Cvir_Myessoensis_1to1),       dplyr::n_distinct(brh_Cvir_Myessoensis_1to1$Cvir),
  "Ruditapes philippinarum",  nrow(brh_Cvir_Rphilippinarum),    nrow(brh_Cvir_Rphilippinarum_1to1),    dplyr::n_distinct(brh_Cvir_Rphilippinarum_1to1$Cvir),
  "Mytilus trossulus",        nrow(brh_Cvir_Mtrossulus),        nrow(brh_Cvir_Mtrossulus_1to1),        dplyr::n_distinct(brh_Cvir_Mtrossulus_1to1$Cvir)
)

brh_count_table

readr::write_tsv(
  brh_count_table,
  file.path(path.expand(out_dir), "Cvir_BRH_counts_by_species.tsv")
)
```

## Focal *C. virginica* membership across comparison species

```{r}
cvir_all <- get_lnc_lengths(
  species_meta$fasta[species_meta$species == "Cvir"]
) %>%
  dplyr::pull(seq_id) %>%
  unique()

membership_cvir <- tibble::tibble(
  Cvir = cvir_all
) %>%
  dplyr::mutate(
    in_Cgigas_roslin   = Cvir %in% brh_Cvir_Cgigas_roslin_1to1$Cvir,
    in_Mgigas_new      = Cvir %in% brh_Cvir_Mgigas_new_1to1$Cvir,
    in_Myessoensis     = Cvir %in% brh_Cvir_Myessoensis_1to1$Cvir,
    in_Rphilippinarum  = Cvir %in% brh_Cvir_Rphilippinarum_1to1$Cvir,
    in_Mtrossulus      = Cvir %in% brh_Cvir_Mtrossulus_1to1$Cvir,
    n_species_shared   = rowSums(dplyr::across(dplyr::starts_with("in_"))),
    sharing_class      = dplyr::case_when(
      n_species_shared == 0 ~ "Cvir_only",
      n_species_shared == 1 ~ "shared_with_one_species",
      n_species_shared >= 2 ~ "shared_with_multiple_species"
    )
  )

membership_cvir %>%
  dplyr::count(sharing_class)

readr::write_tsv(
  membership_cvir,
  file.path(path.expand(out_dir), "Cvir_membership_across_species.tsv")
)
```

## Write FASTAs of BRH-supported *C. virginica* sequences

```{r}
cvir_dna <- read_lnc_fasta(
  species_meta$fasta[species_meta$species == "Cvir"]
)

write_subset_fasta(
  cvir_dna,
  brh_Cvir_Cgigas_roslin_1to1$Cvir,
  file.path(path.expand(fasta_dir), "Cvir_BRH_with_Cgigas_roslin.fasta"),
  prefix = "Cvir"
)

write_subset_fasta(
  cvir_dna,
  brh_Cvir_Mgigas_new_1to1$Cvir,
  file.path(path.expand(fasta_dir), "Cvir_BRH_with_Mgigas_new.fasta"),
  prefix = "Cvir"
)

write_subset_fasta(
  cvir_dna,
  brh_Cvir_Myessoensis_1to1$Cvir,
  file.path(path.expand(fasta_dir), "Cvir_BRH_with_Myessoensis.fasta"),
  prefix = "Cvir"
)

write_subset_fasta(
  cvir_dna,
  brh_Cvir_Rphilippinarum_1to1$Cvir,
  file.path(path.expand(fasta_dir), "Cvir_BRH_with_Rphilippinarum.fasta"),
  prefix = "Cvir"
)

write_subset_fasta(
  cvir_dna,
  brh_Cvir_Mtrossulus_1to1$Cvir,
  file.path(path.expand(fasta_dir), "Cvir_BRH_with_Mtrossulus.fasta"),
  prefix = "Cvir"
)

cvir_shared_any <- membership_cvir %>%
  dplyr::filter(n_species_shared >= 1) %>%
  dplyr::pull(Cvir)

write_subset_fasta(
  cvir_dna,
  cvir_shared_any,
  file.path(path.expand(fasta_dir), "Cvir_BRH_shared_with_any_species.fasta"),
  prefix = "Cvir"
)

cvir_shared_multi <- membership_cvir %>%
  dplyr::filter(n_species_shared >= 2) %>%
  dplyr::pull(Cvir)

write_subset_fasta(
  cvir_dna,
  cvir_shared_multi,
  file.path(path.expand(fasta_dir), "Cvir_BRH_shared_with_multiple_species.fasta"),
  prefix = "Cvir"
)
```

## Export individual BRH tables

```{r}
readr::write_tsv(brh_Cvir_Cgigas_roslin_1to1,  file.path(path.expand(out_dir), "BRH_Cvir_Cgigas_roslin_1to1.tsv"))
readr::write_tsv(brh_Cvir_Mgigas_new_1to1,     file.path(path.expand(out_dir), "BRH_Cvir_Mgigas_new_1to1.tsv"))
readr::write_tsv(brh_Cvir_Myessoensis_1to1,    file.path(path.expand(out_dir), "BRH_Cvir_Myessoensis_1to1.tsv"))
readr::write_tsv(brh_Cvir_Rphilippinarum_1to1, file.path(path.expand(out_dir), "BRH_Cvir_Rphilippinarum_1to1.tsv"))
readr::write_tsv(brh_Cvir_Mtrossulus_1to1,     file.path(path.expand(out_dir), "BRH_Cvir_Mtrossulus_1to1.tsv"))
```

## Session info

```{r}
sessionInfo()
```

## Basic BLASTs

```{bash}
BLAST=/home/shared/ncbi-blast-2.11.0+/bin
DATA_DIR=~/github/oyster-lnc/data/14-comparative-BLASTs
BLAST_DIR=~/github/oyster-lnc/output/16-BRH-blast-comparison/blast

QUERY="$DATA_DIR/GCF_002022765.2_C_virginica-3.0_lncRNA.fa"

"$BLAST/blastn" \
  -query "$QUERY" \
  -db "$BLAST_DIR/Cgigas_roslin_db" \
  -evalue 1e-40 \
  -max_target_seqs 5 \
  -outfmt "6 qseqid sseqid pident length qlen slen evalue bitscore" \
  > "$BLAST_DIR/Cvir_vs_Cgigas_roslin.strict_e40.tsv"

"$BLAST/blastn" \
  -query "$QUERY" \
  -db "$BLAST_DIR/Mgigas_new_db" \
  -evalue 1e-40 \
  -max_target_seqs 5 \
  -outfmt "6 qseqid sseqid pident length qlen slen evalue bitscore" \
  > "$BLAST_DIR/Cvir_vs_Mgigas_new.strict_e40.tsv"

"$BLAST/blastn" \
  -query "$QUERY" \
  -db "$BLAST_DIR/Myessoensis_db" \
  -evalue 1e-40 \
  -max_target_seqs 5 \
  -outfmt "6 qseqid sseqid pident length qlen slen evalue bitscore" \
  > "$BLAST_DIR/Cvir_vs_Myessoensis.strict_e40.tsv"

"$BLAST/blastn" \
  -query "$QUERY" \
  -db "$BLAST_DIR/Rphilippinarum_db" \
  -evalue 1e-40 \
  -max_target_seqs 5 \
  -outfmt "6 qseqid sseqid pident length qlen slen evalue bitscore" \
  > "$BLAST_DIR/Cvir_vs_Rphilippinarum.strict_e40.tsv"

"$BLAST/blastn" \
  -query "$QUERY" \
  -db "$BLAST_DIR/Mtrossulus_db" \
  -evalue 1e-40 \
  -max_target_seqs 5 \
  -outfmt "6 qseqid sseqid pident length qlen slen evalue bitscore" \
  > "$BLAST_DIR/Cvir_vs_Mtrossulus.strict_e40.tsv"
```

## Helper to read outputs
```{r}
read_blast_table <- function(path, comparison_species) {
  if (!file.exists(path) || file.info(path)$size == 0) {
    return(
      tibble(
        qseqid = character(),
        sseqid = character(),
        pident = numeric(),
        aln_len = integer(),
        qlen = integer(),
        slen = integer(),
        evalue = numeric(),
        bitscore = numeric(),
        comparison_species = character(),
        qcov = numeric()
      )
    )
  }

  readr::read_tsv(
    path,
    col_names = c(
      "qseqid", "sseqid", "pident", "aln_len",
      "qlen", "slen", "evalue", "bitscore"
    ),
    show_col_types = FALSE
  ) %>%
    dplyr::mutate(
      comparison_species = comparison_species,
      qcov = aln_len / qlen
    )
}
```

## Combined table of hits
```{r}
blast_hits_all <- dplyr::bind_rows(
  read_blast_table(
    file.path(path.expand(blast_dir), "Cvir_vs_Cgigas_roslin.strict_e40.tsv"),
    "Crassostrea gigas"
  ),
  read_blast_table(
    file.path(path.expand(blast_dir), "Cvir_vs_Mgigas_new.strict_e40.tsv"),
    "Magallana gigas"
  ),
  read_blast_table(
    file.path(path.expand(blast_dir), "Cvir_vs_Myessoensis.strict_e40.tsv"),
    "Mizuhopecten yessoensis"
  ),
  read_blast_table(
    file.path(path.expand(blast_dir), "Cvir_vs_Rphilippinarum.strict_e40.tsv"),
    "Ruditapes philippinarum"
  ),
  read_blast_table(
    file.path(path.expand(blast_dir), "Cvir_vs_Mtrossulus.strict_e40.tsv"),
    "Mytilus trossulus"
  )
)

blast_hits_all
```

## Save combined hits
```{r}
readr::write_tsv(
  blast_hits_all,
  file.path(path.expand(out_dir), "Cvir_vs_all_species_BLAST_hits_strict_e40.tsv")
)
```

## Summary Table

```{r}
readr::write_tsv(
  blast_hits_all,
  file.path(path.expand(out_dir), "Cvir_vs_all_species_BLAST_hits_strict_e40.tsv")
)

print("Cvir_vs_all_species_BLAST_hits_strict_e40.tsv")
```

```{r}
cvir_hit_summary <- blast_hits_all %>%
  dplyr::count(comparison_species, name = "n_hits") %>%
  dplyr::arrange(desc(n_hits))

cvir_hit_summary
```

