---
title: "18-lncRNA-mRNA-pearson"
author: "Zach Bengtsson"
date: "2026-04-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 8)
```

## Packages

```{r packages}
library(tidyverse)
library(igraph)
library(pheatmap)
library(RColorBrewer)
```

## Output Directory

```{r output-dir}
out_dir <- "../output/18-lncRNA-mRNA-pearson"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load Count Matrices

```{r load-data}
data_dir <- "../data/15-pearsons"

lnc_raw <- read_tsv(file.path(data_dir, "lnc_counts.tsv"), show_col_types = FALSE) %>%
  rename(gene_id = 1)

lnc_mat <- lnc_raw %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

mrna_file <- "~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/kallisto_gene_counts.tsv"

mrna_raw <- read_tsv(mrna_file, show_col_types = FALSE) %>%
  rename(gene_id = 1)

mrna_mat <- mrna_raw %>%
  column_to_rownames("gene_id") %>%
  as.matrix()
```

## Align Samples

```{r align-samples}
common_samples <- intersect(colnames(lnc_mat), colnames(mrna_mat))
lnc_mat  <- lnc_mat[, common_samples, drop = FALSE]
mrna_mat <- mrna_mat[, common_samples, drop = FALSE]
n_samples <- length(common_samples)
cat("Common samples:", n_samples, "\n")
```

## Filter Low-Expression Transcripts

```{r filter-low-expression}
min_count   <- 10
min_samples <- 26

filter_rows <- function(mat, min_count, min_samples) {
  keep <- rowSums(mat >= min_count) >= min_samples
  mat[keep, , drop = FALSE]
}

lnc_mat_f  <- filter_rows(lnc_mat,  min_count, min_samples)
mrna_mat_f <- filter_rows(mrna_mat, min_count, min_samples)

cat("lncRNAs after filtering:", nrow(lnc_mat_f), "\n")
cat("mRNAs after filtering:", nrow(mrna_mat_f), "\n")
```

## Filter Transcripts That Do Not Change in Expression

Remove transcripts with very low coefficient of variation (CV) across samples.
This ensures we exclude pairs where neither the lncRNA nor the mRNA meaningfully
varies, which would produce biologically uninformative correlations.

```{r filter-flat-expression}
cv_threshold <- 0.1

compute_cv <- function(x) sd(x) / (mean(x) + 1e-9)

lnc_cv  <- apply(lnc_mat_f,  1, compute_cv)
mrna_cv <- apply(mrna_mat_f, 1, compute_cv)

lnc_variable  <- lnc_mat_f[lnc_cv > cv_threshold, , drop = FALSE]
mrna_variable <- mrna_mat_f[mrna_cv > cv_threshold, , drop = FALSE]

cat("Variable lncRNAs (CV >", cv_threshold, "):", nrow(lnc_variable), "\n")
cat("Variable mRNAs (CV >", cv_threshold, "):", nrow(mrna_variable), "\n")
```

## Log-Transform

```{r log-transform}
lnc_log  <- log2(lnc_variable + 1)
mrna_log <- log2(mrna_variable + 1)
```

## Pearson Correlation

```{r pearson-correlation}
cor_mat <- cor(t(lnc_log), t(mrna_log),
               method = "pearson", use = "pairwise.complete.obs")

cat("Correlation matrix dimensions:", dim(cor_mat), "\n")
```

## Compute P-values and FDR

```{r p-values}
r_vals <- cor_mat
r_vals[r_vals >  0.999999] <-  0.999999
r_vals[r_vals < -0.999999] <- -0.999999

t_vals <- r_vals * sqrt((n_samples - 2) / (1 - r_vals^2))
p_vals <- 2 * pt(-abs(t_vals), df = n_samples - 2)

coexp_df <- as.data.frame(as.table(cor_mat)) %>%
  as_tibble() %>%
  rename(lnc_id = Var1, mrna_id = Var2, r = Freq)

p_df <- as.data.frame(as.table(p_vals)) %>%
  as_tibble() %>%
  rename(lnc_id = Var1, mrna_id = Var2, p_value = Freq)

coexp_df <- coexp_df %>%
  left_join(p_df, by = c("lnc_id", "mrna_id")) %>%
  mutate(FDR = p.adjust(p_value, method = "BH"))
```

## Identify Significant Pairs

```{r significant-pairs}
r_threshold   <- 0.8
fdr_threshold <- 0.05

sig_pairs <- coexp_df %>%
  filter(abs(r) >= r_threshold, FDR <= fdr_threshold) %>%
  arrange(FDR, desc(abs(r)))

cat("Significant lnc-mRNA pairs:", nrow(sig_pairs), "\n")
cat("Unique lncRNAs in network:", n_distinct(sig_pairs$lnc_id), "\n")
cat("Unique mRNAs in network:", n_distinct(sig_pairs$mrna_id), "\n")
cat("Positive correlations:", sum(sig_pairs$r > 0), "\n")
cat("Negative correlations:", sum(sig_pairs$r < 0), "\n")
```

## Export Significant Pairs Table

```{r export-table}
write_tsv(sig_pairs, file.path(out_dir, "significant_lnc_mrna_pairs.tsv"))
```

## Visualization 1: Correlation Distribution

```{r correlation-histogram}
p_hist <- ggplot(sig_pairs, aes(x = r)) +
  geom_histogram(aes(fill = ifelse(r > 0, "Positive", "Negative")),
                 bins = 50, color = "white", linewidth = 0.2) +
  scale_fill_manual(values = c("Positive" = "#E63946", "Negative" = "#457B9D"),
                    name = "Direction") +
  labs(title = "Distribution of Significant lncRNA–mRNA Correlations",
       subtitle = paste0("|r| ≥ ", r_threshold, ", FDR ≤ ", fdr_threshold),
       x = "Pearson r", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

print(p_hist)
ggsave(file.path(out_dir, "correlation_distribution.png"), p_hist,
       width = 8, height = 5, dpi = 300)
```

## Visualization 2: Network of All Significant Interactions

```{r network-plot, fig.width=14, fig.height=14}
edges <- sig_pairs %>%
  select(lnc_id, mrna_id, r, FDR) %>%
  mutate(direction = ifelse(r > 0, "positive", "negative"))

g <- graph_from_data_frame(edges, directed = FALSE)

node_names <- V(g)$name
node_type  <- ifelse(node_names %in% sig_pairs$lnc_id, "lncRNA", "mRNA")
V(g)$type  <- node_type

V(g)$color <- ifelse(node_type == "lncRNA", "#E63946", "#457B9D")
V(g)$frame.color <- ifelse(node_type == "lncRNA", "#9D0208", "#1D3557")
V(g)$size  <- ifelse(node_type == "lncRNA",
                     pmin(degree(g) * 1.2 + 4, 20),
                     pmin(degree(g) * 0.8 + 3, 15))
V(g)$label.cex <- 0.35
V(g)$label.color <- "gray20"

E(g)$color <- ifelse(edges$direction == "positive",
                     adjustcolor("#E63946", alpha.f = 0.3),
                     adjustcolor("#457B9D", alpha.f = 0.3))
E(g)$width <- abs(edges$r) * 2

set.seed(42)
layout <- layout_with_fr(g)

png(file.path(out_dir, "coexpression_network.png"), width = 2400, height = 2400, res = 200)
par(mar = c(1, 1, 3, 1))
plot(g, layout = layout,
     vertex.label = NA,
     edge.curved = 0.1,
     main = "lncRNA–mRNA Co-expression Network")
legend("bottomleft",
       legend = c("lncRNA", "mRNA", "Positive r", "Negative r"),
       col    = c("#E63946", "#457B9D", "#E63946", "#457B9D"),
       pch    = c(16, 16, NA, NA),
       lty    = c(NA, NA, 1, 1),
       lwd    = 2, cex = 0.9, bty = "n")
dev.off()

par(mar = c(1, 1, 3, 1))
plot(g, layout = layout,
     vertex.label = NA,
     edge.curved = 0.1,
     main = "lncRNA–mRNA Co-expression Network")
legend("bottomleft",
       legend = c("lncRNA", "mRNA", "Positive r", "Negative r"),
       col    = c("#E63946", "#457B9D", "#E63946", "#457B9D"),
       pch    = c(16, 16, NA, NA),
       lty    = c(NA, NA, 1, 1),
       lwd    = 2, cex = 0.9, bty = "n")
```

## Visualization 2.1: Re-hash of network to be more interpretable

```{r}
library(igraph)

sig_pairs <- readr::read_tsv(
  "~/github/oyster-lnc/output/18-lncRNA-mRNA-pearson/significant_lnc_mrna_pairs.tsv",
  show_col_types = FALSE
)

edge_df <- sig_pairs %>%
  dplyr::select(from = lnc_id, to = mrna_id, r, FDR)

g <- graph_from_data_frame(edge_df, directed = FALSE)

V(g)$type <- ifelse(V(g)$name %in% sig_pairs$lnc_id, "lncRNA", "mRNA")

deg <- degree(g)

set.seed(123)
layout <- layout_with_fr(g, niter = 2000)

V(g)$color <- ifelse(V(g)$type == "lncRNA", "#E63946", "#457B9D")

V(g)$size <- ifelse(
  V(g)$type == "lncRNA",
  scales::rescale(log1p(deg), to = c(6, 18)),
  scales::rescale(log1p(deg), to = c(4, 10))
)

E(g)$color <- ifelse(edge_df$r > 0, "#E63946AA", "#457B9DAA")
E(g)$width <- 0.5

png(file.path(out_dir, "coexpression_network_refined.png"),
    width = 2400, height = 2400, res = 300)

plot(
  g,
  layout = layout,
  vertex.label = NA,
  vertex.frame.color = NA,
  edge.curved = 0.1
)

legend(
  "topleft",
  legend = c("lncRNA", "mRNA", "Positive correlation", "Negative correlation"),
  col = c("#E63946", "#457B9D", "#E63946", "#457B9D"),
  pch = c(16, 16, NA, NA),
  lty = c(NA, NA, 1, 1),
  bty = "n"
)

dev.off()
```

## Visualization 2.2: Network further refined to only strongest edges

```{r}
edges_filtered <- edges %>%
  dplyr::filter(abs(r) >= 0.8)   # adjust threshold

g <- graph_from_data_frame(edges_filtered, directed = FALSE)

V(g)$type <- ifelse(V(g)$name %in% sig_pairs$lnc_id, "lncRNA", "mRNA")

deg <- degree(g)

set.seed(123)
layout <- layout_with_fr(g, niter = 2000)

V(g)$color <- ifelse(V(g)$type == "lncRNA", "#E63946", "#457B9D")

V(g)$size <- ifelse(
  V(g)$type == "lncRNA",
  scales::rescale(log1p(deg), to = c(6, 18)),
  scales::rescale(log1p(deg), to = c(4, 10))
)

E(g)$color <- ifelse(edge_df$r > 0, "#E63946AA", "#457B9DAA")
E(g)$width <- 0.5

png(file.path(out_dir, "coexpression_network_refined.png"),
    width = 2400, height = 2400, res = 300)

plot(
  g,
  layout = layout,
  vertex.label = NA,
  vertex.frame.color = NA,
  edge.curved = 0.1
)

legend(
  "topleft",
  legend = c("lncRNA", "mRNA", "Positive correlation", "Negative correlation"),
  col = c("#E63946", "#457B9D", "#E63946", "#457B9D"),
  pch = c(16, 16, NA, NA),
  lty = c(NA, NA, 1, 1),
  bty = "n"
)

dev.off()
```


## Visualization 3: Hub lncRNAs (Most Connections)

```{r hub-lncrnas}
lnc_degree <- sig_pairs %>%
  count(lnc_id, name = "n_targets") %>%
  arrange(desc(n_targets))

top_hubs <- head(lnc_degree, 20)

p_hubs <- ggplot(top_hubs, aes(x = reorder(lnc_id, n_targets), y = n_targets)) +
  geom_col(fill = "#E63946", alpha = 0.85) +
  coord_flip() +
  labs(title = "Top 20 Hub lncRNAs by Number of Correlated mRNAs",
       x = NULL, y = "Number of significantly correlated mRNAs") +
  theme_minimal(base_size = 12)

print(p_hubs)
ggsave(file.path(out_dir, "hub_lncrnas.png"), p_hubs,
       width = 10, height = 7, dpi = 300)

write_tsv(lnc_degree, file.path(out_dir, "lncrna_degree_table.tsv"))
```

## Visualization 4: Heatmap of Top Correlations

```{r heatmap, fig.width=12, fig.height=10}
top_lnc <- lnc_degree %>% head(15) %>% pull(lnc_id)

top_mrna <- sig_pairs %>%
  filter(lnc_id %in% top_lnc) %>%
  count(mrna_id, name = "n_lnc") %>%
  arrange(desc(n_lnc)) %>%
  head(40) %>%
  pull(mrna_id)

heatmap_data <- cor_mat[as.character(top_lnc), as.character(top_mrna), drop = FALSE]

png(file.path(out_dir, "correlation_heatmap.png"), width = 2000, height = 1600, res = 200)
pheatmap(heatmap_data,
         color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
         breaks = seq(-1, 1, length.out = 101),
         cluster_rows = TRUE, cluster_cols = TRUE,
         fontsize_row = 7, fontsize_col = 6,
         main = "Pearson Correlation: Top Hub lncRNAs vs. Associated mRNAs",
         angle_col = 45)
dev.off()

pheatmap(heatmap_data,
         color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
         breaks = seq(-1, 1, length.out = 101),
         cluster_rows = TRUE, cluster_cols = TRUE,
         fontsize_row = 7, fontsize_col = 6,
         main = "Pearson Correlation: Top Hub lncRNAs vs. Associated mRNAs",
         angle_col = 45)
```

## Visualization 5: Individual Scatter Plots (Raw-Scale Expression)

This function generates a scatter plot of raw (untransformed) count data for any
lncRNA–mRNA pair, showing the "close to raw" relationship.

```{r scatter-function}
plot_pair_raw <- function(lnc_id_sel, mrna_id_sel,
                          lnc_mat_source = lnc_variable,
                          mrna_mat_source = mrna_variable,
                          samples = common_samples) {

  df <- tibble(
    sample = samples,
    lnc_expr  = as.numeric(lnc_mat_source[lnc_id_sel, samples]),
    mrna_expr = as.numeric(mrna_mat_source[mrna_id_sel, samples])
  )

  r_val <- cor(df$lnc_expr, df$mrna_expr, method = "pearson")

  ggplot(df, aes(x = lnc_expr, y = mrna_expr)) +
    geom_point(size = 3, alpha = 0.75, color = "#264653") +
    geom_smooth(method = "lm", se = TRUE, color = "#E76F51",
                fill = "#E76F51", alpha = 0.15) +
    geom_text(aes(label = sample), size = 2.5, nudge_y = max(df$mrna_expr) * 0.03,
              color = "gray40") +
    labs(title = paste0(lnc_id_sel, " vs ", mrna_id_sel),
         subtitle = sprintf("Pearson r = %.3f  (raw counts, n = %d samples)",
                            r_val, nrow(df)),
         x = paste("lncRNA expression:", lnc_id_sel),
         y = paste("mRNA expression:", mrna_id_sel)) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(size = 11))
}
```

### Example: Top 12 Most Significant Pairs

```{r scatter-top-pairs, fig.width=12, fig.height=16}
top12 <- sig_pairs %>% head(12)

scatter_list <- pmap(top12, function(lnc_id, mrna_id, ...) {
  plot_pair_raw(lnc_id, mrna_id)
})

library(patchwork)
combined <- wrap_plots(scatter_list, ncol = 3)
print(combined)

ggsave(file.path(out_dir, "top12_scatter_raw.png"), combined,
       width = 16, height = 20, dpi = 300)
```

### Save Individual Scatter Plots for All Significant Pairs

```{r scatter-save-all}
scatter_dir <- file.path(out_dir, "scatter_plots")
dir.create(scatter_dir, recursive = TRUE, showWarnings = FALSE)

n_to_save <- min(nrow(sig_pairs), 100)
cat("Saving", n_to_save, "individual scatter plots...\n")

for (i in seq_len(n_to_save)) {
  row <- sig_pairs[i, ]
  p <- plot_pair_raw(row$lnc_id, row$mrna_id)
  fname <- paste0(gsub("[^A-Za-z0-9_.-]", "_", row$lnc_id),
                  "_vs_",
                  gsub("[^A-Za-z0-9_.-]", "_", row$mrna_id),
                  ".png")
  ggsave(file.path(scatter_dir, fname), p, width = 7, height = 5, dpi = 200)
}

cat("Scatter plots saved to:", scatter_dir, "\n")
```

## Summary Statistics

```{r summary}
cat("=== Analysis Summary ===\n")
cat("Input lncRNAs:", nrow(lnc_mat), "\n")
cat("Input mRNAs:", nrow(mrna_mat), "\n")
cat("After low-expression filter: lncRNAs =", nrow(lnc_mat_f),
    ", mRNAs =", nrow(mrna_mat_f), "\n")
cat("After expression-variability filter: lncRNAs =", nrow(lnc_variable),
    ", mRNAs =", nrow(mrna_variable), "\n")
cat("Total pairs tested:", nrow(coexp_df), "\n")
cat("Significant pairs (|r| >=", r_threshold, ", FDR <=", fdr_threshold, "):",
    nrow(sig_pairs), "\n")
cat("Unique lncRNAs with significant associations:", n_distinct(sig_pairs$lnc_id), "\n")
cat("Unique mRNAs with significant associations:", n_distinct(sig_pairs$mrna_id), "\n")
```

## Session Info

```{r session-info}
sessionInfo()
```
