---
title: "18.2-lncRNA-mRNA-pearson"
author: "Zach Bengtsson"
date: "2026-04-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 8)
```

## Packages

```{r packages}
library(tidyverse)
library(igraph)
library(pheatmap)
library(RColorBrewer)
```

## Output Directory

```{r output-dir}
out_dir <- "../output/18.2-lncRNA-mRNA-pearson"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load Count Matrices

```{r load-data}
data_dir <- "../data/15-pearsons"

lnc_raw <- read_tsv(file.path(data_dir, "lnc_counts.tsv"), show_col_types = FALSE) %>%
  rename(gene_id = 1)

lnc_mat <- lnc_raw %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

mrna_file <- "~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/kallisto_gene_counts.tsv"

mrna_raw <- read_tsv(mrna_file, show_col_types = FALSE) %>%
  rename(gene_id = 1)

mrna_mat <- mrna_raw %>%
  column_to_rownames("gene_id") %>%
  as.matrix()
```

## Align Samples

```{r align-samples}
common_samples <- intersect(colnames(lnc_mat), colnames(mrna_mat))
lnc_mat  <- lnc_mat[, common_samples, drop = FALSE]
mrna_mat <- mrna_mat[, common_samples, drop = FALSE]
n_samples <- length(common_samples)
cat("Common samples:", n_samples, "\n")
```

## Filter Low-Expression Transcripts

```{r filter-low-expression}
min_count   <- 10
min_samples <- 26

filter_rows <- function(mat, min_count, min_samples) {
  keep <- rowSums(mat >= min_count) >= min_samples
  mat[keep, , drop = FALSE]
}

lnc_mat_f  <- filter_rows(lnc_mat,  min_count, min_samples)
mrna_mat_f <- filter_rows(mrna_mat, min_count, min_samples)

cat("lncRNAs after filtering:", nrow(lnc_mat_f), "\n")
cat("mRNAs after filtering:", nrow(mrna_mat_f), "\n")
```

## Filter Transcripts That Do Not Change in Expression

Remove transcripts with very low coefficient of variation (CV) across samples.
This ensures we exclude pairs where neither the lncRNA nor the mRNA meaningfully
varies, which would produce biologically uninformative correlations.

```{r filter-flat-expression}
cv_threshold <- 0.1

compute_cv <- function(x) sd(x) / (mean(x) + 1e-9)

lnc_cv  <- apply(lnc_mat_f,  1, compute_cv)
mrna_cv <- apply(mrna_mat_f, 1, compute_cv)

lnc_variable  <- lnc_mat_f[lnc_cv > cv_threshold, , drop = FALSE]
mrna_variable <- mrna_mat_f[mrna_cv > cv_threshold, , drop = FALSE]

cat("Variable lncRNAs (CV >", cv_threshold, "):", nrow(lnc_variable), "\n")
cat("Variable mRNAs (CV >", cv_threshold, "):", nrow(mrna_variable), "\n")
```

## Log-Transform

```{r log-transform}
lnc_log  <- log2(lnc_variable + 1)
mrna_log <- log2(mrna_variable + 1)
```

## Pearson Correlation

```{r pearson-correlation}
cor_mat <- cor(t(lnc_log), t(mrna_log),
               method = "pearson", use = "pairwise.complete.obs")

cat("Correlation matrix dimensions:", dim(cor_mat), "\n")
```

## Compute P-values and FDR

```{r p-values}
r_vals <- cor_mat
r_vals[r_vals >  0.999999] <-  0.999999
r_vals[r_vals < -0.999999] <- -0.999999

t_vals <- r_vals * sqrt((n_samples - 2) / (1 - r_vals^2))
p_vals <- 2 * pt(-abs(t_vals), df = n_samples - 2)

coexp_df <- as.data.frame(as.table(cor_mat)) %>%
  as_tibble() %>%
  rename(lnc_id = Var1, mrna_id = Var2, r = Freq)

p_df <- as.data.frame(as.table(p_vals)) %>%
  as_tibble() %>%
  rename(lnc_id = Var1, mrna_id = Var2, p_value = Freq)

coexp_df <- coexp_df %>%
  left_join(p_df, by = c("lnc_id", "mrna_id")) %>%
  mutate(FDR = p.adjust(p_value, method = "BH"))
```

## Identify Significant Pairs

```{r significant-pairs}
r_threshold   <- 0.99
fdr_threshold <- 0.05

sig_pairs <- coexp_df %>%
  filter(abs(r) >= r_threshold, FDR <= fdr_threshold) %>%
  arrange(FDR, desc(abs(r)))

cat("Significant lnc-mRNA pairs:", nrow(sig_pairs), "\n")
cat("Unique lncRNAs in network:", n_distinct(sig_pairs$lnc_id), "\n")
cat("Unique mRNAs in network:", n_distinct(sig_pairs$mrna_id), "\n")
cat("Positive correlations:", sum(sig_pairs$r > 0), "\n")
cat("Negative correlations:", sum(sig_pairs$r < 0), "\n")
```

## Export Significant Pairs Table

```{r export-table}
write_tsv(sig_pairs, file.path(out_dir, "significant_lnc_mrna_pairs_99.tsv"))
```

## Visualization 1: Correlation Distribution

```{r correlation-histogram}
p_hist <- ggplot(sig_pairs, aes(x = r)) +
  geom_histogram(aes(fill = ifelse(r > 0, "Positive", "Negative")),
                 bins = 50, color = "white", linewidth = 0.2) +
  scale_fill_manual(values = c("Positive" = "#E63946", "Negative" = "#457B9D"),
                    name = "Direction") +
  labs(title = "Distribution of Significant lncRNA–mRNA Correlations",
       subtitle = paste0("|r| ≥ ", r_threshold, ", FDR ≤ ", fdr_threshold),
       x = "Pearson r", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

print(p_hist)
ggsave(file.path(out_dir, "correlation_distribution_99.png"), p_hist,
       width = 8, height = 5, dpi = 300)
```

## Visualization 2: Network of All Significant Interactions

```{r network-plot, fig.width=14, fig.height=14}
edges <- sig_pairs %>%
  select(lnc_id, mrna_id, r, FDR) %>%
  mutate(direction = ifelse(r > 0, "positive", "negative"))

g <- graph_from_data_frame(edges, directed = FALSE)

node_names <- V(g)$name
node_type  <- ifelse(node_names %in% sig_pairs$lnc_id, "lncRNA", "mRNA")
V(g)$type  <- node_type

V(g)$color <- ifelse(node_type == "lncRNA", "#E63946", "#457B9D")
V(g)$frame.color <- ifelse(node_type == "lncRNA", "#9D0208", "#1D3557")
V(g)$size  <- ifelse(node_type == "lncRNA",
                     pmin(degree(g) * 1.2 + 4, 20),
                     pmin(degree(g) * 0.8 + 3, 15))
V(g)$label.cex <- 0.35
V(g)$label.color <- "gray20"

E(g)$color <- ifelse(edges$direction == "positive",
                     adjustcolor("#E63946", alpha.f = 0.3),
                     adjustcolor("#457B9D", alpha.f = 0.3))
E(g)$width <- abs(edges$r) * 2

set.seed(42)
layout <- layout_with_fr(g)

png(file.path(out_dir, "coexpression_network_99.png"), width = 2400, height = 2400, res = 200)
par(mar = c(1, 1, 3, 1))
plot(g, layout = layout,
     vertex.label = NA,
     edge.curved = 0.1,
     main = "lncRNA–mRNA Co-expression Network")
legend("bottomleft",
       legend = c("lncRNA", "mRNA", "Positive r", "Negative r"),
       col    = c("#E63946", "#457B9D", "#E63946", "#457B9D"),
       pch    = c(16, 16, NA, NA),
       lty    = c(NA, NA, 1, 1),
       lwd    = 2, cex = 0.9, bty = "n")
dev.off()

par(mar = c(1, 1, 3, 1))
plot(g, layout = layout,
     vertex.label = NA,
     edge.curved = 0.1,
     main = "lncRNA–mRNA Co-expression Network")
legend("bottomleft",
       legend = c("lncRNA", "mRNA", "Positive r", "Negative r"),
       col    = c("#E63946", "#457B9D", "#E63946", "#457B9D"),
       pch    = c(16, 16, NA, NA),
       lty    = c(NA, NA, 1, 1),
       lwd    = 2, cex = 0.9, bty = "n")
```

## Visualization 2.1: Re-hash of network to be more interpretable

```{r}
library(igraph)

sig_pairs <- readr::read_tsv(
  "~/github/oyster-lnc/output/18.1-lncRNA-mRNA-pearson/significant_lnc_mrna_pairs.tsv",
  show_col_types = FALSE
)

edge_df <- sig_pairs %>%
  dplyr::select(from = lnc_id, to = mrna_id, r, FDR)

g <- graph_from_data_frame(edge_df, directed = FALSE)

V(g)$type <- ifelse(V(g)$name %in% sig_pairs$lnc_id, "lncRNA", "mRNA")

deg <- degree(g)

set.seed(123)
layout <- layout_with_fr(g, niter = 2000)

V(g)$color <- ifelse(V(g)$type == "lncRNA", "#E63946", "#457B9D")

V(g)$size <- ifelse(
  V(g)$type == "lncRNA",
  scales::rescale(log1p(deg), to = c(6, 18)),
  scales::rescale(log1p(deg), to = c(4, 10))
)

E(g)$color <- ifelse(edge_df$r > 0, "#E63946AA", "#457B9DAA")
E(g)$width <- 0.5

png(file.path(out_dir, "coexpression_network_refined.png"),
    width = 2400, height = 2400, res = 300)

plot(
  g,
  layout = layout,
  vertex.label = NA,
  vertex.frame.color = NA,
  edge.curved = 0.1
)

legend(
  "topleft",
  legend = c("lncRNA", "mRNA", "Positive correlation", "Negative correlation"),
  col = c("#E63946", "#457B9D", "#E63946", "#457B9D"),
  pch = c(16, 16, NA, NA),
  lty = c(NA, NA, 1, 1),
  bty = "n"
)

dev.off()
```

## Visualization 2.2: Network further refined to only strongest edges

```{r}
edges_filtered <- edges %>%
  dplyr::filter(abs(r) >= 0.8)   # adjust threshold

g <- graph_from_data_frame(edges_filtered, directed = FALSE)

V(g)$type <- ifelse(V(g)$name %in% sig_pairs$lnc_id, "lncRNA", "mRNA")

deg <- degree(g)

set.seed(123)
layout <- layout_with_fr(g, niter = 2000)

V(g)$color <- ifelse(V(g)$type == "lncRNA", "#E63946", "#457B9D")

V(g)$size <- ifelse(
  V(g)$type == "lncRNA",
  scales::rescale(log1p(deg), to = c(6, 18)),
  scales::rescale(log1p(deg), to = c(4, 10))
)

E(g)$color <- ifelse(edge_df$r > 0, "#E63946AA", "#457B9DAA")
E(g)$width <- 0.5

png(file.path(out_dir, "coexpression_network_refined.png"),
    width = 2400, height = 2400, res = 300)

plot(
  g,
  layout = layout,
  vertex.label = NA,
  vertex.frame.color = NA,
  edge.curved = 0.1
)

legend(
  "topleft",
  legend = c("lncRNA", "mRNA", "Positive correlation", "Negative correlation"),
  col = c("#E63946", "#457B9D", "#E63946", "#457B9D"),
  pch = c(16, 16, NA, NA),
  lty = c(NA, NA, 1, 1),
  bty = "n"
)

dev.off()
```


## Visualization 3: Hub lncRNAs (Most Connections)

```{r hub-lncrnas}
lnc_degree <- sig_pairs %>%
  count(lnc_id, name = "n_targets") %>%
  arrange(desc(n_targets))

top_hubs <- head(lnc_degree, 20)

p_hubs <- ggplot(top_hubs, aes(x = reorder(lnc_id, n_targets), y = n_targets)) +
  geom_col(fill = "#E63946", alpha = 0.85) +
  coord_flip() +
  labs(title = "Top 20 Hub lncRNAs by Number of Correlated mRNAs",
       x = NULL, y = "Number of significantly correlated mRNAs") +
  theme_minimal(base_size = 12)

print(p_hubs)
ggsave(file.path(out_dir, "hub_lncrnas.png"), p_hubs,
       width = 10, height = 7, dpi = 300)

write_tsv(lnc_degree, file.path(out_dir, "lncrna_degree_table_99.tsv"))
```

## Visualization 4: Heatmap of Top Correlations

```{r heatmap, fig.width=12, fig.height=10}
top_lnc <- lnc_degree %>% head(15) %>% pull(lnc_id)

top_mrna <- sig_pairs %>%
  filter(lnc_id %in% top_lnc) %>%
  count(mrna_id, name = "n_lnc") %>%
  arrange(desc(n_lnc)) %>%
  head(40) %>%
  pull(mrna_id)

heatmap_data <- cor_mat[as.character(top_lnc), as.character(top_mrna), drop = FALSE]

png(file.path(out_dir, "correlation_heatmap.png"), width = 2000, height = 1600, res = 200)
pheatmap(heatmap_data,
         color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
         breaks = seq(-1, 1, length.out = 101),
         cluster_rows = TRUE, cluster_cols = TRUE,
         fontsize_row = 7, fontsize_col = 6,
         main = "Pearson Correlation: Top Hub lncRNAs vs. Associated mRNAs",
         angle_col = 45)
dev.off()

pheatmap(heatmap_data,
         color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
         breaks = seq(-1, 1, length.out = 101),
         cluster_rows = TRUE, cluster_cols = TRUE,
         fontsize_row = 7, fontsize_col = 6,
         main = "Pearson Correlation: Top Hub lncRNAs vs. Associated mRNAs",
         angle_col = 45)
```

## Visualization 5: Individual Scatter Plots

This function generates a scatter plot of log count data for any
lncRNA–mRNA pair.

```{r scatter-function}
plot_pair_log <- function(lnc_id_sel, mrna_id_sel,
                          lnc_mat_source = lnc_log,
                          mrna_mat_source = mrna_log,
                          samples = common_samples) {

  df <- tibble::tibble(
    sample = samples,
    lnc_expr  = as.numeric(lnc_mat_source[lnc_id_sel, samples]),
    mrna_expr = as.numeric(mrna_mat_source[mrna_id_sel, samples])
  )

  r_val <- cor(df$lnc_expr, df$mrna_expr,
               method = "pearson",
               use = "pairwise.complete.obs")

  ggplot2::ggplot(df, ggplot2::aes(x = lnc_expr, y = mrna_expr)) +
    ggplot2::geom_point(size = 3, alpha = 0.75, color = "#264653") +
    ggplot2::geom_smooth(
      method = "lm",
      se = TRUE,
      color = "#E76F51",
      fill = "#E76F51",
      alpha = 0.15
    ) +
    ggplot2::geom_text(
      ggplot2::aes(label = sample),
      size = 2.5,
      nudge_y = max(df$mrna_expr, na.rm = TRUE) * 0.03,
      color = "gray40"
    ) +
    ggplot2::labs(
      title = paste0(lnc_id_sel, " vs ", mrna_id_sel),
      subtitle = sprintf(
        "Pearson r = %.3f  (log2-transformed counts, n = %d samples)",
        r_val, nrow(df)
      ),
      x = paste("lncRNA expression:", lnc_id_sel),
      y = paste("mRNA expression:", mrna_id_sel)
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(plot.title = ggplot2::element_text(size = 11))
}
```

### Save Individual Scatter Plots for All Significant Pairs

```{r scatter-save-all}
scatter_dir <- file.path(out_dir, "scatter_plots")
dir.create(scatter_dir, recursive = TRUE, showWarnings = FALSE)

n_to_save <- min(nrow(sig_pairs), 100)
cat("Saving", n_to_save, "individual scatter plots...\n")

for (i in seq_len(n_to_save)) {
  row <- sig_pairs[i, ]
  p <- plot_pair_log(row$lnc_id, row$mrna_id)
  fname <- paste0(gsub("[^A-Za-z0-9_.-]", "_", row$lnc_id),
                  "_vs_",
                  gsub("[^A-Za-z0-9_.-]", "_", row$mrna_id),
                  ".png")
  ggsave(file.path(scatter_dir, fname), p, width = 7, height = 5, dpi = 200)
}

cat("Scatter plots saved to:", scatter_dir, "\n")
```

## Summary Statistics

```{r summary}
cat("=== Analysis Summary ===\n")
cat("Input lncRNAs:", nrow(lnc_mat), "\n")
cat("Input mRNAs:", nrow(mrna_mat), "\n")
cat("After low-expression filter: lncRNAs =", nrow(lnc_mat_f),
    ", mRNAs =", nrow(mrna_mat_f), "\n")
cat("After expression-variability filter: lncRNAs =", nrow(lnc_variable),
    ", mRNAs =", nrow(mrna_variable), "\n")
cat("Total pairs tested:", nrow(coexp_df), "\n")
cat("Significant pairs (|r| >=", r_threshold, ", FDR <=", fdr_threshold, "):",
    nrow(sig_pairs), "\n")
cat("Unique lncRNAs with significant associations:", n_distinct(sig_pairs$lnc_id), "\n")
cat("Unique mRNAs with significant associations:", n_distinct(sig_pairs$mrna_id), "\n")
```

## Session Info

```{r session-info}
sessionInfo()
```

## Hub tables

```{r}
lnc_hub_table <- sig_pairs %>%
  dplyr::count(lnc_id, name = "n_mRNAs_correlated") %>%
  dplyr::arrange(desc(n_mRNAs_correlated))

lnc_hub_table
```

```{r}
readr::write_tsv(
  lnc_hub_table,
  file.path(out_dir, "lncRNA_hub_table.tsv")
)
```

```{r}
mrna_hub_table <- sig_pairs %>%
  dplyr::count(mrna_id, name = "n_lncRNAs_correlated") %>%
  dplyr::arrange(desc(n_lncRNAs_correlated))

mrna_hub_table
```

```{r}
readr::write_tsv(
  mrna_hub_table,
  file.path(out_dir, "mRNA_hub_table.tsv")
)
```

```{bash}
cd ~/github/oyster-lnc/output/18.1-lncRNA-mRNA-pearson/scatter_plots
convert -delay 50 -loop 0 *.png ../scatter_plots.gif
```

### Get scatter points color coded by sex

```{r}
plot_pair_log_sex <- function(lnc_id_sel, mrna_id_sel,
                              lnc_mat_source = lnc_log,
                              mrna_mat_source = mrna_log,
                              samples = common_samples) {

  df <- tibble::tibble(
    sample = samples,
    lnc_expr  = as.numeric(lnc_mat_source[lnc_id_sel, samples]),
    mrna_expr = as.numeric(mrna_mat_source[mrna_id_sel, samples])
  ) %>%
    dplyr::mutate(
      sex = ifelse(grepl("M$", sample), "Male",
                   ifelse(grepl("F$", sample), "Female", NA))
    )

  r_val <- cor(df$lnc_expr, df$mrna_expr,
               method = "pearson",
               use = "pairwise.complete.obs")

  ggplot2::ggplot(df, ggplot2::aes(x = lnc_expr, y = mrna_expr, color = sex)) +
    ggplot2::geom_point(size = 3, alpha = 0.85) +
    ggplot2::geom_smooth(
      method = "lm",
      se = TRUE,
      color = "black",
      fill = "gray70",
      alpha = 0.2
    ) +
    ggplot2::geom_text(
      ggplot2::aes(label = sample),
      size = 2.5,
      nudge_y = max(df$mrna_expr, na.rm = TRUE) * 0.03,
      color = "gray40"
    ) +
    ggplot2::scale_color_manual(
      values = c("Male" = "#457B9D", "Female" = "#E63946")
    ) +
    ggplot2::labs(
      title = paste0(lnc_id_sel, " vs ", mrna_id_sel),
      subtitle = sprintf(
        "Pearson r = %.3f  (log2-transformed counts, n = %d samples)",
        r_val, nrow(df)
      ),
      x = paste("lncRNA expression:", lnc_id_sel),
      y = paste("mRNA expression:", mrna_id_sel),
      color = "Sex"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(plot.title = ggplot2::element_text(size = 11))
}
```

```{r}
scatter_dir_sex <- file.path(out_dir, "scatter_plots_by_sex")
dir.create(scatter_dir_sex, recursive = TRUE, showWarnings = FALSE)
```

```{r}
n_to_save <- min(nrow(sig_pairs), 100)
cat("Saving", n_to_save, "sex-colored scatter plots...\n")

for (i in seq_len(n_to_save)) {
  row <- sig_pairs[i, ]
  p <- plot_pair_log_sex(row$lnc_id, row$mrna_id)

  fname <- paste0(
    gsub("[^A-Za-z0-9_.-]", "_", row$lnc_id),
    "_vs_",
    gsub("[^A-Za-z0-9_.-]", "_", row$mrna_id),
    "_sex.png"   # <-- helpful suffix
  )

  ggsave(file.path(scatter_dir_sex, fname),
         p, width = 7, height = 5, dpi = 200)
}

cat("Sex-colored scatter plots saved to:", scatter_dir_sex, "\n")
```

```{bash}
cd ~/github/oyster-lnc/output/18.2-lncRNA-mRNA-pearson/scatter_plots_by_sex 
convert -delay 50 -loop 0 *.png ../scatter_plots_by_sex.gif
```

### Examine unusually large lncRNA hub

```{r}
## ---------------------------
## Inspect top lncRNA hub expression
## ---------------------------

# Identify top lncRNA hub
top_lnc_hub <- sig_pairs %>%
  dplyr::count(lnc_id, name = "n_mRNAs_correlated") %>%
  dplyr::arrange(dplyr::desc(n_mRNAs_correlated)) %>%
  dplyr::slice_head(n = 1)

top_lnc_id <- top_lnc_hub$lnc_id[1]
top_lnc_id
top_lnc_hub

# Pull all mRNAs correlated with this lncRNA
top_lnc_edges <- sig_pairs %>%
  dplyr::filter(lnc_id == top_lnc_id) %>%
  dplyr::arrange(dplyr::desc(abs(r)))

top_lnc_mrnas <- unique(top_lnc_edges$mrna_id)

length(top_lnc_mrnas)
```

```{r}
## ---------------------------
## Expression summary for the lncRNA hub
## ---------------------------

lnc_expr_summary <- tibble::tibble(
  transcript_id = top_lnc_id,
  type = "lncRNA",
  mean_log_expr = mean(as.numeric(lnc_log[top_lnc_id, common_samples]), na.rm = TRUE),
  sd_log_expr   = sd(as.numeric(lnc_log[top_lnc_id, common_samples]), na.rm = TRUE),
  cv_log_expr   = sd_log_expr / (mean_log_expr + 1e-9),
  min_log_expr  = min(as.numeric(lnc_log[top_lnc_id, common_samples]), na.rm = TRUE),
  max_log_expr  = max(as.numeric(lnc_log[top_lnc_id, common_samples]), na.rm = TRUE),
  range_log_expr = max_log_expr - min_log_expr
)

lnc_expr_summary
```
```{r}
## ---------------------------
## Expression summary for associated mRNAs
## ---------------------------

mrna_expr_summary <- purrr::map_dfr(top_lnc_mrnas, function(gene) {
  x <- as.numeric(mrna_log[gene, common_samples])

  tibble::tibble(
    transcript_id = gene,
    type = "mRNA",
    mean_log_expr = mean(x, na.rm = TRUE),
    sd_log_expr = sd(x, na.rm = TRUE),
    cv_log_expr = sd_log_expr / (mean_log_expr + 1e-9),
    min_log_expr = min(x, na.rm = TRUE),
    max_log_expr = max(x, na.rm = TRUE),
    range_log_expr = max_log_expr - min_log_expr
  )
})

mrna_expr_summary %>%
  dplyr::arrange(cv_log_expr)
```

```{r}
hub_target_expression_table <- top_lnc_edges %>%
  dplyr::left_join(
    mrna_expr_summary,
    by = c("mrna_id" = "transcript_id")
  )

readr::write_tsv(
  hub_target_expression_table,
  file.path(out_dir, paste0("top_lnc_hub_", gsub("[^A-Za-z0-9_.-]", "_", top_lnc_id), "_target_expression_summary.tsv"))
)

readr::write_tsv(
  lnc_expr_summary,
  file.path(out_dir, paste0("top_lnc_hub_", gsub("[^A-Za-z0-9_.-]", "_", top_lnc_id), "_expression_summary.tsv"))
)
```

```{r}
ggplot2::ggplot(mrna_expr_summary, ggplot2::aes(x = cv_log_expr)) +
  ggplot2::geom_histogram(bins = 30) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = paste("CV distribution of mRNAs correlated with", top_lnc_id),
    x = "CV of log2(count + 1) expression",
    y = "Number of mRNAs"
  )
```

```{r}
hub_expression_matrix <- dplyr::bind_rows(
  tibble::tibble(
    transcript_id = top_lnc_id,
    type = "lncRNA",
    as.list(as.numeric(lnc_log[top_lnc_id, common_samples]))
  ),
  mrna_log[top_lnc_mrnas, common_samples, drop = FALSE] %>%
    as.data.frame() %>%
    tibble::rownames_to_column("transcript_id") %>%
    dplyr::mutate(type = "mRNA")
)

colnames(hub_expression_matrix)[3:ncol(hub_expression_matrix)] <- common_samples

readr::write_tsv(
  hub_expression_matrix,
  file.path(out_dir, paste0("top_lnc_hub_", gsub("[^A-Za-z0-9_.-]", "_", top_lnc_id), "_expression_matrix.tsv"))
)
```

```{r}
df_lnc <- tibble(
  sample = common_samples,
  expr = as.numeric(lnc_log["rna-XR_002638603.1", common_samples])
)

ggplot(df_lnc, aes(x = sample, y = expr)) +
  geom_point() +
  geom_line(group = 1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

```{r}
## ---------------------------
## Histograms of expression for mRNAs correlated with top lncRNA hub
## ---------------------------

# Use the top lncRNA hub you already identified
top_lnc_id <- "rna-XR_002638603.1"

# Get correlated mRNAs
top_lnc_mrnas <- sig_pairs %>%
  dplyr::filter(lnc_id == top_lnc_id) %>%
  dplyr::pull(mrna_id) %>%
  unique()

# Convert expression matrix to long format
mrna_expr_long <- mrna_log[top_lnc_mrnas, common_samples, drop = FALSE] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("mrna_id") %>%
  tidyr::pivot_longer(
    cols = -mrna_id,
    names_to = "sample",
    values_to = "expr"
  )

# Plot histograms for all correlated mRNAs
p_mrna_hist <- ggplot2::ggplot(mrna_expr_long, ggplot2::aes(x = expr)) +
  ggplot2::geom_histogram(bins = 10, fill = "gray50", color = "white") +
  ggplot2::facet_wrap(~ mrna_id, scales = "free_y") +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::labs(
    title = paste("Expression distributions of mRNAs correlated with", top_lnc_id),
    x = "log2(count + 1) expression",
    y = "Number of samples"
  )

p_mrna_hist
```

```{r}
## ---------------------------
## Per-mRNA sample expression bar plots
## ---------------------------

mrna_barplot_dir <- file.path(out_dir, "mrna_sample_expression_barplots")
dir.create(mrna_barplot_dir, recursive = TRUE, showWarnings = FALSE)

top_lnc_id <- "rna-XR_002638603.1"

top_lnc_mrnas <- sig_pairs %>%
  dplyr::filter(lnc_id == top_lnc_id) %>%
  dplyr::arrange(dplyr::desc(abs(r))) %>%
  dplyr::pull(mrna_id) %>%
  unique()

cat("Saving sample-level expression barplots for", length(top_lnc_mrnas), "mRNAs...\n")

for (mrna in top_lnc_mrnas) {

  df <- tibble::tibble(
    sample = factor(common_samples, levels = common_samples),
    expr = as.numeric(mrna_log[mrna, common_samples])
  )

  p <- ggplot2::ggplot(df, ggplot2::aes(x = sample, y = expr)) +
    ggplot2::geom_col(fill = "gray50") +
    ggplot2::labs(
      title = mrna,
      subtitle = paste("Expression across samples for mRNA correlated with", top_lnc_id),
      x = "Sample",
      y = "log2(count + 1) expression"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5),
      plot.title = ggplot2::element_text(size = 11)
    )

  fname <- paste0(gsub("[^A-Za-z0-9_.-]", "_", mrna), ".png")

  ggplot2::ggsave(
    file.path(mrna_barplot_dir, fname),
    p,
    width = 7,
    height = 4,
    dpi = 200
  )
}

cat("Saved to:", mrna_barplot_dir, "\n")
```

```{r}
## ---------------------------
## Per-mRNA sample expression bar plots: raw counts
## ---------------------------

mrna_raw_barplot_dir <- file.path(out_dir, "mrna_sample_expression_barplots_raw_counts")
dir.create(mrna_raw_barplot_dir, recursive = TRUE, showWarnings = FALSE)

top_lnc_id <- "rna-XR_002638603.1"

top_lnc_mrnas <- sig_pairs %>%
  dplyr::filter(lnc_id == top_lnc_id) %>%
  dplyr::arrange(dplyr::desc(abs(r))) %>%
  dplyr::pull(mrna_id) %>%
  unique()

cat("Saving raw-count sample-level expression barplots for", length(top_lnc_mrnas), "mRNAs...\n")

for (mrna in top_lnc_mrnas) {

  df <- tibble::tibble(
    sample = factor(common_samples, levels = common_samples),
    expr = as.numeric(mrna_variable[mrna, common_samples])
  )

  p <- ggplot2::ggplot(df, ggplot2::aes(x = sample, y = expr)) +
    ggplot2::geom_col(fill = "gray50") +
    ggplot2::labs(
      title = mrna,
      subtitle = paste("Raw count expression across samples for mRNA correlated with", top_lnc_id),
      x = "Sample",
      y = "Raw count expression"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5),
      plot.title = ggplot2::element_text(size = 11)
    )

  fname <- paste0(gsub("[^A-Za-z0-9_.-]", "_", mrna), ".png")

  ggplot2::ggsave(
    file.path(mrna_raw_barplot_dir, fname),
    p,
    width = 7,
    height = 4,
    dpi = 200
  )
}

cat("Saved to:", mrna_raw_barplot_dir, "\n")
```

### GIFs of expression barplots for mRNAs

```{bash}
cd ~/github/oyster-lnc/output/18.2-lncRNA-mRNA-pearson/mrna_sample_expression_barplots

convert -delay 50 -loop 0 $(ls *.png | sort) ../mrna_sample_expression_barplots.gif
```
```{bash}
cd ~/github/oyster-lnc/output/18.2-lncRNA-mRNA-pearson/mrna_sample_expression_barplots_raw_counts

convert -delay 50 -loop 0 $(ls *.png | sort) ../mrna_sample_expression_barplots_raw_counts.gif
```

### GO enrichment for 0.99 threshold dataset

```{r}
## ---------------------------
## GO enrichment: Biological Process only
## For r >= 0.99 sig_pairs dataset
## ---------------------------

library(tidyverse)
library(clusterProfiler)
library(enrichplot)

out_dir <- "../output/18.2-lncRNA-mRNA-pearson"
go99_dir <- file.path(out_dir, "GO_BP_r099")
dir.create(go99_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r}
## ---------------------------
## Restrict GO annotation to Biological Process
## ---------------------------

gene2go_bp <- gene2go %>%
  dplyr::filter(Category == "Process") %>%   # NCBI gene2go BP category
  dplyr::mutate(GeneID = as.character(GeneID)) %>%
  dplyr::distinct(GeneID, GO_ID, GO_term, Category)

term2gene_bp <- gene2go_bp %>%
  dplyr::transmute(term = GO_ID, gene = GeneID) %>%
  dplyr::distinct()

term2name_bp <- gene2go_bp %>%
  dplyr::transmute(term = GO_ID, name = GO_term) %>%
  dplyr::distinct()
```

```{r}
### Check
gene2go %>%
  dplyr::count(Category)
```

```{r}
## ---------------------------
## Define mRNA hubs from r >= 0.99 network
## ---------------------------

mrna_degree_099 <- sig_pairs %>%
  dplyr::count(mrna_id, name = "n_lncRNAs_correlated") %>%
  dplyr::arrange(dplyr::desc(n_lncRNAs_correlated))

hub_cutoff_mrna_099 <- quantile(
  mrna_degree_099$n_lncRNAs_correlated,
  probs = 0.9
)

hub_mrnas_099 <- mrna_degree_099 %>%
  dplyr::filter(n_lncRNAs_correlated >= hub_cutoff_mrna_099)

readr::write_tsv(
  mrna_degree_099,
  file.path(go99_dir, "mRNA_degree_r099.tsv")
)

readr::write_tsv(
  hub_mrnas_099,
  file.path(go99_dir, "hub_mRNAs_top10percent_r099.tsv")
)
```

```{r}
## ---------------------------
## Map hub mRNAs to GeneIDs
## ---------------------------

hub_gene_ids_099 <- hub_mrnas_099 %>%
  dplyr::left_join(
    id_map %>% dplyr::select(mrna_id, GeneID),
    by = "mrna_id"
  ) %>%
  dplyr::filter(!is.na(GeneID)) %>%
  dplyr::pull(GeneID) %>%
  unique()
```

```{r}
### Redefine universe
mrna_universe_ids <- rownames(mrna_variable)

gene_universe_bp <- id_map %>%
  dplyr::filter(mrna_id %in% mrna_universe_ids, !is.na(GeneID)) %>%
  dplyr::pull(GeneID) %>%
  unique()
```

```{r}
### Run enrichment

ego_hubs_bp_099 <- clusterProfiler::enricher(
  gene          = hub_gene_ids_099,
  universe      = gene_universe_bp,
  TERM2GENE     = term2gene_bp,
  TERM2NAME     = term2name_bp,
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  minGSSize     = 10,
  maxGSSize     = 500
)

ego_hubs_bp_099_df <- as.data.frame(ego_hubs_bp_099)

readr::write_tsv(
  ego_hubs_bp_099_df,
  file.path(go99_dir, "GO_BP_enrichment_hub_mRNAs_r099.tsv")
)
```

```{r}
if (!is.null(ego_hubs_bp_099) && nrow(as.data.frame(ego_hubs_bp_099)) > 0) {

  p_bp_099 <- enrichplot::dotplot(
    ego_hubs_bp_099,
    showCategory = 20
  ) +
    ggplot2::ggtitle("Biological Process GO enrichment of hub mRNAs, |r| ≥ 0.99")

  ggplot2::ggsave(
    file.path(go99_dir, "GO_BP_dotplot_hub_mRNAs_r099.png"),
    p_bp_099,
    width = 10,
    height = 7,
    dpi = 300
  )

  p_bp_099
}
```

```{r}
top_lnc_099 <- sig_pairs %>%
  dplyr::count(lnc_id, name = "n_targets") %>%
  dplyr::arrange(dplyr::desc(n_targets)) %>%
  dplyr::slice_head(n = 5) %>%
  dplyr::pull(lnc_id)

lnc_bp_results_099 <- list()

for (lnc in top_lnc_099) {

  target_gene_ids <- sig_pairs %>%
    dplyr::filter(lnc_id == lnc) %>%
    dplyr::left_join(
      id_map %>% dplyr::select(mrna_id, GeneID),
      by = "mrna_id"
    ) %>%
    dplyr::filter(!is.na(GeneID)) %>%
    dplyr::pull(GeneID) %>%
    unique()

  if (length(target_gene_ids) < 10) next

  ego_lnc_bp <- clusterProfiler::enricher(
    gene          = target_gene_ids,
    universe      = gene_universe_bp,
    TERM2GENE     = term2gene_bp,
    TERM2NAME     = term2name_bp,
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.05,
    minGSSize     = 10,
    maxGSSize     = 500
  )

  if (!is.null(ego_lnc_bp) && nrow(as.data.frame(ego_lnc_bp)) > 0) {

    ego_lnc_bp_df <- as.data.frame(ego_lnc_bp) %>%
      dplyr::mutate(lnc_id = lnc)

    lnc_bp_results_099[[lnc]] <- ego_lnc_bp_df

    safe_name <- gsub("[^A-Za-z0-9_.-]", "_", lnc)

    p_lnc_bp <- enrichplot::dotplot(
      ego_lnc_bp,
      showCategory = 12
    ) +
      ggplot2::ggtitle(paste("BP GO enrichment for targets of", lnc))

    ggplot2::ggsave(
      file.path(go99_dir, paste0("GO_BP_dotplot_", safe_name, "_r099.png")),
      p_lnc_bp,
      width = 10,
      height = 6,
      dpi = 300
    )
  }
}

if (length(lnc_bp_results_099) > 0) {
  all_lnc_bp_099 <- dplyr::bind_rows(lnc_bp_results_099)

  readr::write_tsv(
    all_lnc_bp_099,
    file.path(go99_dir, "GO_BP_enrichment_top_lnc_modules_r099.tsv")
  )
}
```

### Mapping Check

```{r}
mrna_ids <- rownames(mrna_variable)

mapping_df <- tibble(mrna_id = mrna_ids) %>%
  left_join(id_map, by = "mrna_id")

n_total <- nrow(mapping_df)
n_mapped <- sum(!is.na(mapping_df$GeneID))

percent_mapped <- (n_mapped / n_total) * 100

cat("Total mRNAs:", n_total, "\n")
cat("Mapped to GeneID:", n_mapped, "\n")
cat("Percent mapped:", round(percent_mapped, 2), "%\n")
```

```{r}
sig_mrnas <- sig_pairs %>%
  pull(mrna_id) %>%
  unique()

mapping_sig <- tibble(mrna_id = sig_mrnas) %>%
  left_join(id_map, by = "mrna_id")

n_total_sig <- nrow(mapping_sig)
n_mapped_sig <- sum(!is.na(mapping_sig$GeneID))

percent_mapped_sig <- (n_mapped_sig / n_total_sig) * 100

cat("Significant mRNAs:", n_total_sig, "\n")
cat("Mapped to GeneID:", n_mapped_sig, "\n")
cat("Percent mapped:", round(percent_mapped_sig, 2), "%\n")
```

```{r}
mapped_gene_ids <- mapping_sig %>%
  filter(!is.na(GeneID)) %>%
  pull(GeneID)

genes_with_go <- gene2go %>%
  filter(GeneID %in% mapped_gene_ids) %>%
  pull(GeneID) %>%
  unique()

n_with_go <- length(genes_with_go)

percent_with_go <- (n_with_go / length(mapped_gene_ids)) * 100

cat("Mapped genes:", length(mapped_gene_ids), "\n")
cat("With GO terms:", n_with_go, "\n")
cat("Percent with GO:", round(percent_with_go, 2), "%\n")
```

```{r}
gene2go_bp <- gene2go %>%
  filter(Category == "Process")

genes_with_bp <- gene2go_bp %>%
  filter(GeneID %in% mapped_gene_ids) %>%
  pull(GeneID) %>%
  unique()

percent_with_bp <- (length(genes_with_bp) / length(mapped_gene_ids)) * 100

cat("Percent with BP GO terms:", round(percent_with_bp, 2), "%\n")
```

```{r}
## ---------------------------
## Number of correlated lncRNAs per mRNA
## ---------------------------

mrna_lnc_degree <- sig_pairs %>%
  dplyr::count(mrna_id, name = "n_correlated_lncRNAs") %>%
  dplyr::arrange(dplyr::desc(n_correlated_lncRNAs))

readr::write_tsv(
  mrna_lnc_degree,
  file.path(out_dir, "mRNA_correlated_lncRNA_counts.tsv")
)

p_mrna_lnc_degree <- ggplot2::ggplot(
  mrna_lnc_degree,
  ggplot2::aes(x = n_correlated_lncRNAs)
) +
  ggplot2::geom_histogram(bins = 30, fill = "gray50", color = "white") +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::labs(
    title = "Number of correlated lncRNAs per mRNA",
    subtitle = paste0("Significant pairs: |r| ≥ ", r_threshold, ", FDR ≤ ", fdr_threshold),
    x = "Number of correlated lncRNAs",
    y = "Number of mRNAs"
  )

p_mrna_lnc_degree

ggplot2::ggsave(
  file.path(out_dir, "mRNA_correlated_lncRNA_count_distribution.png"),
  p_mrna_lnc_degree,
  width = 8,
  height = 6,
  dpi = 300
)
```

```{r}
lnc_mrna_degree <- sig_pairs %>%
  dplyr::count(lnc_id, name = "n_correlated_mRNAs") %>%
  dplyr::arrange(dplyr::desc(n_correlated_mRNAs))

readr::write_tsv(
  lnc_mrna_degree,
  file.path(out_dir, "lncRNA_correlated_mRNA_counts.tsv")
)
```

```{r}
p_lnc_degree <- ggplot2::ggplot(
  lnc_mrna_degree,
  ggplot2::aes(x = n_correlated_mRNAs)
) +
  ggplot2::geom_histogram(bins = 30, fill = "#E63946", color = "white") +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::labs(
    title = "Number of correlated mRNAs per lncRNA",
    subtitle = paste0("Significant pairs: |r| ≥ ", r_threshold, ", FDR ≤ ", fdr_threshold),
    x = "Number of correlated mRNAs",
    y = "Number of lncRNAs"
  )

p_lnc_degree

ggplot2::ggsave(
  file.path(out_dir, "lncRNA_correlated_mRNA_count_distribution.png"),
  p_lnc_degree,
  width = 8,
  height = 6,
  dpi = 300
)
```

```{r}
p_lnc_degree_log <- ggplot2::ggplot(
  lnc_mrna_degree,
  ggplot2::aes(x = n_correlated_mRNAs)
) +
  ggplot2::geom_histogram(bins = 30, fill = "#E63946", color = "white") +
  ggplot2::scale_y_log10() +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::labs(
    title = "Number of correlated mRNAs per lncRNA",
    subtitle = paste0("Significant pairs: |r| ≥ ", r_threshold, ", FDR ≤ ", fdr_threshold),
    x = "Number of correlated mRNAs",
    y = "Number of lncRNAs (log10 scale)"
  )

p_lnc_degree_log

ggplot2::ggsave(
  file.path(out_dir, "lncRNA_correlated_mRNA_count_distribution_logY.png"),
  p_lnc_degree_log,
  width = 8,
  height = 6,
  dpi = 300
)
```


```{r}
## ---------------------------
## Expression barplots for focal lncRNA hub
## ---------------------------

focal_lnc <- "rna-XR_002638603.1"

lnc_barplot_dir <- file.path(out_dir, "lncRNA_hub_expression_barplots")
dir.create(lnc_barplot_dir, recursive = TRUE, showWarnings = FALSE)

## Log2 expression plot
df_lnc_log <- tibble::tibble(
  sample = factor(common_samples, levels = common_samples),
  expr = as.numeric(lnc_log[focal_lnc, common_samples])
)

p_lnc_log <- ggplot2::ggplot(df_lnc_log, ggplot2::aes(x = sample, y = expr)) +
  ggplot2::geom_col(fill = "#E63946") +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5),
    plot.title = ggplot2::element_text(size = 11)
  ) +
  ggplot2::labs(
    title = focal_lnc,
    subtitle = "lncRNA hub expression across samples",
    x = "Sample",
    y = "log2(count + 1) expression"
  )

ggplot2::ggsave(
  file.path(lnc_barplot_dir, paste0(focal_lnc, "_log2_expression.png")),
  p_lnc_log,
  width = 7,
  height = 4,
  dpi = 200
)

## Raw count plot
df_lnc_raw <- tibble::tibble(
  sample = factor(common_samples, levels = common_samples),
  expr = as.numeric(lnc_variable[focal_lnc, common_samples])
)

p_lnc_raw <- ggplot2::ggplot(df_lnc_raw, ggplot2::aes(x = sample, y = expr)) +
  ggplot2::geom_col(fill = "#E63946") +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5),
    plot.title = ggplot2::element_text(size = 11)
  ) +
  ggplot2::labs(
    title = focal_lnc,
    subtitle = "lncRNA hub raw count expression across samples",
    x = "Sample",
    y = "Raw count expression"
  )

ggplot2::ggsave(
  file.path(lnc_barplot_dir, paste0(focal_lnc, "_raw_count_expression.png")),
  p_lnc_raw,
  width = 7,
  height = 4,
  dpi = 200
)

p_lnc_log
p_lnc_raw
```


## Visualization 6: Cluster-Labeled Network and Cluster-Specific Scatterplots

The full lncRNA–mRNA co-expression network contains multiple disconnected subnetworks. 
Here, each disconnected subnetwork is treated as a separate network cluster using 
`igraph::components()`. Clusters are labeled alphabetically starting with A. For each 
cluster, this section exports a table of the lncRNA–mRNA pairs in that cluster and saves 
sex-colored scatterplots for only those pairs into a cluster-specific folder.

```{r}
## ---------------------------------
## Identify disconnected network clusters with visible labels
## ---------------------------------

cluster_edges <- sig_pairs %>%
  dplyr::select(from = lnc_id, to = mrna_id, r, FDR)

g_clusters <- igraph::graph_from_data_frame(cluster_edges, directed = FALSE)

comp <- igraph::components(g_clusters)

cluster_membership <- tibble::tibble(
  node = names(comp$membership),
  component_id = as.integer(comp$membership),
  cluster_size = comp$csize[as.integer(comp$membership)]
) %>%
  dplyr::mutate(
    node_type = ifelse(node %in% sig_pairs$lnc_id, "lncRNA", "mRNA")
  )

make_cluster_labels <- function(n) {
  sapply(seq_len(n), function(i) {
    x <- i
    out <- character()

    while (x > 0) {
      rem <- (x - 1) %% 26
      out <- c(LETTERS[rem + 1], out)
      x <- (x - 1) %/% 26
    }

    paste0(out, collapse = "")
  })
}

## Label by size: largest cluster = A
cluster_label_key <- cluster_membership %>%
  dplyr::distinct(component_id, cluster_size) %>%
  dplyr::arrange(dplyr::desc(cluster_size), component_id) %>%
  dplyr::mutate(cluster_label = make_cluster_labels(dplyr::n()))

cluster_membership <- cluster_membership %>%
  dplyr::left_join(cluster_label_key, by = c("component_id", "cluster_size"))

## Make layout, then normalize it so plot() and text() use same coordinates
set.seed(42)

cluster_layout_raw <- igraph::layout_with_fr(g_clusters, niter = 3000)

cluster_layout <- igraph::norm_coords(
  cluster_layout_raw,
  xmin = -1,
  xmax = 1,
  ymin = -1,
  ymax = 1
)

layout_tbl <- tibble::tibble(
  node = igraph::V(g_clusters)$name,
  x = cluster_layout[, 1],
  y = cluster_layout[, 2]
) %>%
  dplyr::left_join(cluster_membership, by = "node")

## Place labels above each cluster, within the visible plotting area
cluster_positions <- layout_tbl %>%
  dplyr::group_by(component_id, cluster_label) %>%
  dplyr::summarise(
    xmin = min(x),
    xmax = max(x),
    ymin = min(y),
    ymax = max(y),
    label_x = mean(c(xmin, xmax)),
    label_y = pmin(ymax + 0.08, 1.12),
    n_nodes = dplyr::n(),
    .groups = "drop"
  ) %>%
  dplyr::arrange(cluster_label)

## Add cluster labels to pairs
pair_clusters <- sig_pairs %>%
  dplyr::left_join(
    cluster_membership %>%
      dplyr::select(node, component_id, cluster_label) %>%
      dplyr::rename(
        lnc_id = node,
        component_id_lnc = component_id
      ),
    by = "lnc_id"
  ) %>%
  dplyr::left_join(
    cluster_membership %>%
      dplyr::select(node, component_id) %>%
      dplyr::rename(
        mrna_id = node,
        component_id_mrna = component_id
      ),
    by = "mrna_id"
  )

stopifnot(all(pair_clusters$component_id_lnc == pair_clusters$component_id_mrna))

pair_clusters <- pair_clusters %>%
  dplyr::mutate(component_id = component_id_lnc) %>%
  dplyr::select(-component_id_lnc, -component_id_mrna) %>%
  dplyr::arrange(cluster_label, dplyr::desc(abs(r)), FDR)

cluster_summary <- pair_clusters %>%
  dplyr::group_by(cluster_label, component_id) %>%
  dplyr::summarise(
    n_pairs = dplyr::n(),
    n_lncRNAs = dplyr::n_distinct(lnc_id),
    n_mRNAs = dplyr::n_distinct(mrna_id),
    .groups = "drop"
  ) %>%
  dplyr::arrange(cluster_label)

cluster_summary
```

```{r}
readr::write_tsv(
  cluster_membership,
  file.path(out_dir, "node_cluster_membership_99.tsv")
)

readr::write_tsv(
  pair_clusters,
  file.path(out_dir, "significant_pairs_with_cluster_labels_99.tsv")
)

readr::write_tsv(
  cluster_summary,
  file.path(out_dir, "cluster_summary_99.tsv")
)
```

```{r}
## ---------------------------------
## Plot network with cluster labels that should actually appear
## ---------------------------------

igraph::V(g_clusters)$type <- ifelse(
  igraph::V(g_clusters)$name %in% sig_pairs$lnc_id,
  "lncRNA",
  "mRNA"
)

deg <- igraph::degree(g_clusters)

igraph::V(g_clusters)$color <- ifelse(
  igraph::V(g_clusters)$type == "lncRNA",
  "#E63946",
  "#457B9D"
)

igraph::V(g_clusters)$frame.color <- ifelse(
  igraph::V(g_clusters)$type == "lncRNA",
  "#9D0208",
  "#1D3557"
)

igraph::V(g_clusters)$size <- ifelse(
  igraph::V(g_clusters)$type == "lncRNA",
  scales::rescale(log1p(deg), to = c(6, 22)),
  scales::rescale(log1p(deg), to = c(4, 14))
)

igraph::E(g_clusters)$color <- ifelse(
  cluster_edges$r > 0,
  adjustcolor("#E63946", alpha.f = 0.35),
  adjustcolor("#457B9D", alpha.f = 0.35)
)

igraph::E(g_clusters)$width <- abs(cluster_edges$r) * 1.5

png(
  file.path(out_dir, "coexpression_network_99_cluster_labeled.png"),
  width = 2600,
  height = 2600,
  res = 300
)

par(mar = c(2, 2, 4, 2), xpd = NA)

plot(
  g_clusters,
  layout = cluster_layout,
  vertex.label = NA,
  edge.curved = 0.1,
  main = "lncRNA–mRNA Co-expression Network with Cluster Labels",
  rescale = FALSE,
  xlim = c(-1.2, 1.2),
  ylim = c(-1.2, 1.2)
)

text(
  x = cluster_positions$label_x,
  y = cluster_positions$label_y,
  labels = cluster_positions$cluster_label,
  cex = 1.6,
  font = 2,
  col = "black",
  xpd = NA
)

legend(
  "bottomleft",
  legend = c("lncRNA", "mRNA", "Positive r", "Negative r"),
  col = c("#E63946", "#457B9D", "#E63946", "#457B9D"),
  pch = c(16, 16, NA, NA),
  lty = c(NA, NA, 1, 1),
  lwd = 2,
  cex = 0.9,
  bty = "n"
)

dev.off()

## Also display in notebook
par(mar = c(2, 2, 4, 2), xpd = NA)

plot(
  g_clusters,
  layout = cluster_layout,
  vertex.label = NA,
  edge.curved = 0.1,
  main = "lncRNA–mRNA Co-expression Network with Cluster Labels",
  rescale = FALSE,
  xlim = c(-1.2, 1.2),
  ylim = c(-1.2, 1.2)
)

text(
  x = cluster_positions$label_x,
  y = cluster_positions$label_y,
  labels = cluster_positions$cluster_label,
  cex = 1.6,
  font = 2,
  col = "black",
  xpd = NA
)

legend(
  "bottomleft",
  legend = c("lncRNA", "mRNA", "Positive r", "Negative r"),
  col = c("#E63946", "#457B9D", "#E63946", "#457B9D"),
  pch = c(16, 16, NA, NA),
  lty = c(NA, NA, 1, 1),
  lwd = 2,
  cex = 0.9,
  bty = "n"
)
```

```{r}
## ---------------------------------
## Export cluster membership tables
## ---------------------------------

readr::write_tsv(
  cluster_membership,
  file.path(out_dir, "node_cluster_membership_99.tsv")
)

readr::write_tsv(
  pair_clusters,
  file.path(out_dir, "significant_pairs_with_cluster_labels_99.tsv")
)

readr::write_tsv(
  cluster_summary,
  file.path(out_dir, "cluster_summary_99.tsv")
)

cluster_summary
```

```{r}
## ---------------------------------
## Save sex-colored scatterplots by cluster
## ---------------------------------

cluster_base_dir <- file.path(out_dir, "cluster_scatter_plots_by_sex")
dir.create(cluster_base_dir, recursive = TRUE, showWarnings = FALSE)

safe_name <- function(x) {
  gsub("[^A-Za-z0-9_.-]", "_", x)
}

for (cl in cluster_summary$cluster_label) {

  cluster_dir <- file.path(cluster_base_dir, paste0("cluster_", cl))
  dir.create(cluster_dir, recursive = TRUE, showWarnings = FALSE)

  cluster_pairs <- pair_clusters %>%
    dplyr::filter(cluster_label == cl) %>%
    dplyr::arrange(dplyr::desc(abs(r)), FDR)

  # Save pair table for this cluster
  readr::write_tsv(
    cluster_pairs,
    file.path(cluster_dir, paste0("cluster_", cl, "_pairs.tsv"))
  )

  cat("Saving", nrow(cluster_pairs), "sex-colored scatterplots for cluster", cl, "...\n")

  for (i in seq_len(nrow(cluster_pairs))) {

    row <- cluster_pairs[i, ]

    p <- plot_pair_log_sex(
      lnc_id_sel = row$lnc_id,
      mrna_id_sel = row$mrna_id
    )

    fname <- paste0(
      "cluster_", cl,
      "__",
      safe_name(row$lnc_id),
      "_vs_",
      safe_name(row$mrna_id),
      "_sex.png"
    )

    ggplot2::ggsave(
      filename = file.path(cluster_dir, fname),
      plot = p,
      width = 7,
      height = 5,
      dpi = 200
    )
  }
}

cat("Cluster-specific sex-colored scatterplots saved to:", cluster_base_dir, "\n")
```

```{r}
## ---------------------------------
## Cluster summary with lncRNA and mRNA IDs
## ---------------------------------

cluster_id_summary <- pair_clusters %>%
  dplyr::group_by(cluster_label) %>%
  dplyr::summarise(
    n_pairs = dplyr::n(),
    n_lncRNAs = dplyr::n_distinct(lnc_id),
    n_mRNAs = dplyr::n_distinct(mrna_id),
    lncRNAs = paste(sort(unique(lnc_id)), collapse = "; "),
    mRNAs = paste(sort(unique(mrna_id)), collapse = "; "),
    .groups = "drop"
  ) %>%
  dplyr::arrange(cluster_label)

cluster_id_summary

readr::write_tsv(
  cluster_id_summary,
  file.path(out_dir, "cluster_summary_with_ids_99.tsv")
)
```

