---
title: "19-GO-from-coexpression"
author: "Zach Bengtsson"
date: "2026-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

GO enrichment for Eastern oyster coexpression results
when mRNA IDs look like LOC111120067

## ---------------------------
## Packages
## ---------------------------

```{r}
packages <- c(
  "tidyverse",
  "clusterProfiler",
  "enrichplot",
  "GO.db"
)

to_install <- packages[!vapply(packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(to_install) > 0) {
  install.packages(setdiff(to_install, c("clusterProfiler", "enrichplot", "GO.db")))
  if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
  bioc_pkgs <- intersect(to_install, c("clusterProfiler", "enrichplot", "GO.db"))
  if (length(bioc_pkgs) > 0) {
    BiocManager::install(bioc_pkgs, ask = FALSE, update = FALSE)
  }
}

library(tidyverse)
library(clusterProfiler)
library(enrichplot)
library(GO.db)
```

## ---------------------------
## Input / output
## ---------------------------
```{r}
coexp_dir <- "../output/18-lncRNA-mRNA-pearson"
out_dir   <- "../output/19-GO-from-coexpression"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

sig_pairs_file <- file.path(coexp_dir, "significant_lnc_mrna_pairs.tsv")
mrna_file      <- "~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/kallisto_gene_counts.tsv"
```
## ---------------------------
## Load data
## ---------------------------

```{r}
sig_pairs <- readr::read_tsv(sig_pairs_file, show_col_types = FALSE)

mrna_raw <- readr::read_tsv(mrna_file, show_col_types = FALSE)
colnames(mrna_raw)[1] <- "mrna_id"

mrna_universe_ids <- unique(mrna_raw$mrna_id)
```

## ---------------------------
## Hub mRNA summary
## ---------------------------

```{r}
mrna_degree <- sig_pairs %>%
  count(mrna_id, name = "n_lncRNAs") %>%
  arrange(desc(n_lncRNAs), mrna_id)

readr::write_tsv(mrna_degree, file.path(out_dir, "mrna_degree_table.tsv"))

n_hub <- max(10, ceiling(0.10 * nrow(mrna_degree)))

hub_mrnas <- mrna_degree %>%
  slice_head(n = n_hub)

readr::write_tsv(hub_mrnas, file.path(out_dir, "hub_mrnas_top10pct.tsv"))
```

## ---------------------------
## Download NCBI files
## ---------------------------

```{r}
tax_id <- 6565

gene2go_url   <- "https://ftp.ncbi.nlm.nih.gov/gene/DATA/gene2go.gz"
gene_info_url <- "https://ftp.ncbi.nlm.nih.gov/gene/DATA/GENE_INFO/All_Data.gene_info.gz"

gene2go_oyster_file   <- file.path(out_dir, "gene2go_oyster.tsv")
gene_info_oyster_file <- file.path(out_dir, "gene_info_oyster.tsv")

if (!file.exists(gene2go_oyster_file)) {
  gene2go <- data.table::fread(
    cmd = paste(
      "curl -L", shQuote(gene2go_url), "| gzip -dc"
    ),
    sep = "\t",
    header = FALSE,
    skip = 1,
    quote = "",
    fill = TRUE,
    showProgress = TRUE,
    col.names = c(
      "tax_id", "GeneID", "GO_ID", "Evidence", "Qualifier",
      "GO_term", "PubMed", "Category"
    )
  ) |>
    tibble::as_tibble() |>
    dplyr::filter(tax_id == !!tax_id) |>
    dplyr::mutate(GeneID = as.character(GeneID)) |>
    dplyr::distinct(GeneID, GO_ID, GO_term, Category, .keep_all = TRUE)

  readr::write_tsv(gene2go, gene2go_oyster_file)
}

if (!file.exists(gene_info_oyster_file)) {
  gene_info <- data.table::fread(
    cmd = paste(
      "curl -L", shQuote(gene_info_url), "| gzip -dc"
    ),
    sep = "\t",
    header = FALSE,
    skip = 1,
    quote = "",
    fill = TRUE,
    showProgress = TRUE,
    col.names = c(
      "tax_id", "GeneID", "Symbol", "LocusTag", "Synonyms",
      "dbXrefs", "chromosome", "map_location", "description",
      "type_of_gene", "Symbol_from_nomenclature_authority",
      "Full_name_from_nomenclature_authority",
      "Nomenclature_status", "Other_designations",
      "Modification_date", "Feature_type"
    )
  ) |>
    tibble::as_tibble() |>
    dplyr::filter(tax_id == !!tax_id) |>
    dplyr::mutate(
      GeneID = as.character(GeneID),
      Symbol = dplyr::na_if(Symbol, "-"),
      LocusTag = dplyr::na_if(LocusTag, "-"),
      Synonyms = dplyr::na_if(Synonyms, "-")
    )

  readr::write_tsv(gene_info, gene_info_oyster_file)
}

cat("Cached oyster annotation files:\n")
cat(" -", gene2go_oyster_file, "\n")
cat(" -", gene_info_oyster_file, "\n")
```
## ---------------------------
## Cache Building
## ---------------------------

```{r}
tax_id <- 6565
gene2go_url <- "https://ftp.ncbi.nlm.nih.gov/gene/DATA/gene2go.gz"
gene_info_url <- "https://ftp.ncbi.nlm.nih.gov/gene/DATA/GENE_INFO/All_Data.gene_info.gz"

gene2go_oyster_file   <- file.path(out_dir, "gene2go_oyster.tsv")
gene_info_oyster_file <- file.path(out_dir, "gene_info_oyster.tsv")
```


## ---------------------------
## Read gene2go
## ---------------------------

```{r}
gene2go <- readr::read_tsv(
  gene2go_oyster_file,
  show_col_types = FALSE,
  col_types = readr::cols(.default = readr::col_character())
) %>%
  mutate(GeneID = as.character(GeneID)) %>%
  distinct(GeneID, GO_ID, GO_term, Category, .keep_all = TRUE)
```

## ---------------------------
## Read gene_info
## ---------------------------

```{r}
gene_info <- readr::read_tsv(
  gene_info_oyster_file,
  show_col_types = FALSE,
  col_types = readr::cols(.default = readr::col_character())
) %>%
  mutate(
    GeneID   = as.character(GeneID),
    Symbol   = na_if(Symbol, "-"),
    LocusTag = na_if(LocusTag, "-"),
    Synonyms = na_if(Synonyms, "-")
  )
```

## ---------------------------
## Clean IDs
## ---------------------------

```{r}
clean_id <- function(x) {
  x %>%
    as.character() %>%
    stringr::str_trim() %>%
    stringr::str_remove("\\.[0-9]+$")
}

all_ids_tbl <- tibble(
  mrna_id = unique(c(mrna_universe_ids, sig_pairs$mrna_id))
) %>%
  mutate(mrna_id_clean = clean_id(mrna_id))
```

## ---------------------------
## Primary mapping:
## LOC identifiers should match NCBI gene_info$Symbol directly
## ---------------------------

```{r}
id_map_symbol <- all_ids_tbl %>%
  left_join(
    gene_info %>%
      transmute(
        GeneID,
        Symbol,
        description,
        type_of_gene,
        match_value = Symbol
      ),
    by = c("mrna_id_clean" = "match_value")
  ) %>%
  mutate(match_type = ifelse(!is.na(GeneID), "Symbol", NA_character_))
```

## ---------------------------
## Secondary fallback:
## if an mrna_id is actually already a GeneID number
## ---------------------------

```{r}
id_map_geneid <- all_ids_tbl %>%
  left_join(
    gene_info %>%
      transmute(
        GeneID,
        Symbol,
        description,
        type_of_gene,
        match_value = GeneID
      ),
    by = c("mrna_id_clean" = "match_value")
  ) %>%
  mutate(match_type = ifelse(!is.na(GeneID), "GeneID", NA_character_))
```

## ---------------------------
## Combine mappings, prefer Symbol match first
## ---------------------------

```{r}
id_map <- id_map_symbol %>%
  bind_rows(id_map_geneid) %>%
  group_by(mrna_id) %>%
  arrange(factor(match_type, levels = c("Symbol", "GeneID"))) %>%
  slice_head(n = 1) %>%
  ungroup()

readr::write_tsv(id_map, file.path(out_dir, "mrna_id_to_ncbi_GeneID_map.tsv"))

mapping_summary <- tibble(
  n_unique_mrna_ids = n_distinct(all_ids_tbl$mrna_id),
  n_mapped          = sum(!is.na(id_map$GeneID)),
  prop_mapped       = mean(!is.na(id_map$GeneID))
)

print(mapping_summary)
readr::write_tsv(mapping_summary, file.path(out_dir, "mapping_summary.tsv"))
```

## ---------------------------
## Annotate network tables
## ---------------------------

```{r}
sig_pairs_annot <- sig_pairs %>%
  left_join(
    id_map %>% dplyr::select(mrna_id, GeneID, Symbol, description, type_of_gene, match_type),
    by = "mrna_id"
  )

mrna_degree_annot <- mrna_degree %>%
  left_join(
    id_map %>% dplyr::select(mrna_id, GeneID, Symbol, description, type_of_gene, match_type),
    by = "mrna_id"
  )

readr::write_tsv(sig_pairs_annot, file.path(out_dir, "significant_pairs_with_geneIDs.tsv"))
readr::write_tsv(mrna_degree_annot, file.path(out_dir, "mrna_degree_with_geneIDs.tsv"))
```

## ---------------------------
## TERM2GENE / TERM2NAME
## ---------------------------

```{r}
term2gene <- gene2go %>%
  transmute(term = GO_ID, gene = GeneID) %>%
  distinct()

term2name <- gene2go %>%
  transmute(term = GO_ID, name = GO_term) %>%
  distinct()
```

## ---------------------------
## Recreate filtered mRNA universe (mrna_variable)
## ---------------------------

```{r}
# Load mRNA matrix
mrna_raw <- readr::read_tsv(mrna_file, show_col_types = FALSE) %>%
  dplyr::rename(mrna_id = gene_id)

mrna_mat <- mrna_raw %>%
  tibble::column_to_rownames("mrna_id") %>%
  as.matrix()
```

## ---------------------------
## Filter low-expression genes
## ---------------------------

```{r}
min_count   <- 10
min_samples <- 26

filter_rows <- function(mat, min_count, min_samples) {
  keep <- rowSums(mat >= min_count) >= min_samples
  mat[keep, , drop = FALSE]
}

mrna_mat_f <- filter_rows(mrna_mat, min_count, min_samples)

cat("After expression filter:", nrow(mrna_mat_f), "genes\n")
```

## ---------------------------
## Filter low-variability genes (CV filter)
## ---------------------------

```{r}
cv_threshold <- 0.1

compute_cv <- function(x) {
  sd(x) / (mean(x) + 1e-9)
}

mrna_cv <- apply(mrna_mat_f, 1, compute_cv)

mrna_variable <- mrna_mat_f[mrna_cv > cv_threshold, , drop = FALSE]

cat("After CV filter:", nrow(mrna_variable), "genes\n")
```

## ---------------------------
## Define correct background universe
## ---------------------------

```{r}
mrna_universe_ids <- rownames(mrna_variable)

cat("Final universe size:", length(mrna_universe_ids), "\n")
```

## ---------------------------
## Background universe
## ---------------------------

```{r}
gene_universe <- id_map %>%
  dplyr::filter(mrna_id %in% mrna_universe_ids, !is.na(GeneID)) %>%
  dplyr::pull(GeneID) %>%
  unique()
```

## ---------------------------
## GO enrichment: hub mRNAs
## ---------------------------

```{r}
hub_gene_ids <- hub_mrnas %>%
  left_join(id_map %>% dplyr::select(mrna_id, GeneID), by = "mrna_id") %>%
  dplyr::filter(!is.na(GeneID)) %>%
  dplyr::pull(GeneID) %>%
  unique()

ego_hubs <- enricher(
  gene          = hub_gene_ids,
  universe      = gene_universe,
  TERM2GENE     = term2gene,
  TERM2NAME     = term2name,
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  minGSSize     = 10,
  maxGSSize     = 500
)

if (!is.null(ego_hubs) && nrow(as.data.frame(ego_hubs)) > 0) {
  ego_hubs_df <- as.data.frame(ego_hubs) %>%
    dplyr::mutate(
      ONTOLOGY = purrr::map_chr(ID, ~{
        x <- gene2go %>%
          dplyr::filter(GO_ID == .x) %>%
          dplyr::slice_head(n = 1) %>%
          dplyr::pull(Category)
        if (length(x) == 0) NA_character_ else x
      })
    )

  readr::write_tsv(ego_hubs_df, file.path(out_dir, "GO_enrichment_hub_mrnas.tsv"))

  p_dot <- dotplot(ego_hubs, showCategory = 20) +
    ggplot2::ggtitle("GO enrichment of hub mRNAs")

  ggplot2::ggsave(
    file.path(out_dir, "GO_dotplot_hub_mrnas.png"),
    p_dot, width = 10, height = 7, dpi = 300
  )
}
```

```{r}
as.data.frame(ego_hubs) %>% head()
```


## ---------------------------
## GO enrichment: top hub lncRNA target sets
## ---------------------------

```{r}
top_lnc <- sig_pairs %>%
  dplyr::count(lnc_id, name = "n_targets") %>%
  dplyr::arrange(desc(n_targets)) %>%
  dplyr::slice_head(n = 5) %>%
  dplyr::pull(lnc_id)

lnc_go_results <- list()

for (lnc in top_lnc) {
  target_gene_ids <- sig_pairs %>%
    dplyr::filter(lnc_id == lnc) %>%
    left_join(id_map %>% dplyr::select(mrna_id, GeneID), by = "mrna_id") %>%
    dplyr::filter(!is.na(GeneID)) %>%
    dplyr::pull(GeneID) %>%
    unique()

  if (length(target_gene_ids) < 10) next

  ego_lnc <- enricher(
    gene          = target_gene_ids,
    universe      = gene_universe,
    TERM2GENE     = term2gene,
    TERM2NAME     = term2name,
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.05,
    minGSSize     = 10,
    maxGSSize     = 500
  )

  if (!is.null(ego_lnc) && nrow(as.data.frame(ego_lnc)) > 0) {
    ego_lnc_df <- as.data.frame(ego_lnc) %>%
      dplyr::mutate(lnc_id = lnc)

    lnc_go_results[[lnc]] <- ego_lnc_df

    safe_name <- gsub("[^A-Za-z0-9_.-]", "_", lnc)
    p <- dotplot(ego_lnc, showCategory = 12) +
      ggplot2::ggtitle(paste("GO enrichment for targets of", lnc))

    ggplot2::ggsave(
      file.path(out_dir, paste0("GO_dotplot_", safe_name, ".png")),
      p, width = 10, height = 6, dpi = 300
    )
  }
}

if (length(lnc_go_results) > 0) {
  all_lnc_go <- dplyr::bind_rows(lnc_go_results)
  readr::write_tsv(all_lnc_go, file.path(out_dir, "GO_enrichment_top_lnc_modules.tsv"))
}
```

## ---------------------------
## Convenient annotation table for manuscript use
## ---------------------------

```{r}
top30_hub_mrna_annotation <- mrna_degree %>%
  dplyr::slice_head(n = 30) %>%
  left_join(
    id_map %>% dplyr::select(mrna_id, GeneID, Symbol, description, type_of_gene),
    by = "mrna_id"
  ) %>%
  left_join(
    gene2go %>%
      dplyr::group_by(GeneID) %>%
      dplyr::summarise(
        n_go_terms = dplyr::n_distinct(GO_ID),
        example_go_terms = paste(head(unique(GO_term), 5), collapse = "; "),
        .groups = "drop"
      ),
    by = "GeneID"
  )

readr::write_tsv(
  top30_hub_mrna_annotation,
  file.path(out_dir, "top30_hub_mrna_annotations.tsv")
)

cat("Done.\n")
cat("Mapped mRNA IDs:", sum(!is.na(id_map$GeneID)), "of", dplyr::n_distinct(id_map$mrna_id), "\n")
cat("Output written to:", out_dir, "\n")
```

## ---------------------------
## GO enrichment: lncRNA hub selection for demonstration
## ---------------------------

# Load top lnc hubs
```{r}
all_lnc_go <- readr::read_tsv(
  "~/github/oyster-lnc/output/19-GO-from-coexpression/GO_enrichment_top_lnc_modules.tsv",
  show_col_types = FALSE
)

```

```{r}
lnc_summary <- all_lnc_go %>%
  dplyr::group_by(lnc_id) %>%
  dplyr::summarise(
    n_terms = dplyr::n(),
    min_padj = min(p.adjust, na.rm = TRUE),
    top_term = Description[which.min(p.adjust)],
    .groups = "drop"
  ) %>%
  dplyr::arrange(min_padj, dplyr::desc(n_terms))

# write to TSV
readr::write_tsv(
  lnc_summary,
  file.path("~/github/oyster-lnc/output/19-GO-from-coexpression/lnc_summary.tsv")
)

lnc_summary
```

### Inspect terms by lncRNA

```{r}
all_lnc_go %>%
  dplyr::select(lnc_id, Description, p.adjust, Count) %>%
  dplyr::arrange(lnc_id, p.adjust) %>%
  print(n = 100)
```

```{r}
split(all_lnc_go %>% dplyr::arrange(p.adjust), all_lnc_go$lnc_id)
```

### Figures for 3

```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)

all_lnc_go <- readr::read_tsv(
  "~/github/oyster-lnc/output/19-GO-from-coexpression/GO_enrichment_top_lnc_modules.tsv",
  show_col_types = FALSE
)

focus_lnc <- c(
  "rna-XR_002634986.1",
  "rna-XR_002638098.1",
  "rna-XR_002638049.1"
)

plot_lnc_go <- function(df, lnc_name, panel_title, n_terms = 10) {
  df_sub <- df %>%
    dplyr::filter(lnc_id == lnc_name) %>%
    dplyr::arrange(p.adjust) %>%
    dplyr::slice_head(n = n_terms) %>%
    dplyr::mutate(
      GeneRatio_num = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1]) / as.numeric(x[2])),
      Description = factor(Description, levels = rev(unique(Description)))
    )

  ggplot(df_sub, aes(x = GeneRatio_num, y = Description)) +
    geom_point(aes(size = Count, color = p.adjust)) +
    scale_color_viridis_c(direction = -1) +
    labs(
      title = panel_title,
      x = "GeneRatio",
      y = NULL,
      color = "Adj. p",
      size = "Count"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 11),
      axis.text.y = element_text(size = 9)
    )
}

p1 <- plot_lnc_go(all_lnc_go, "rna-XR_002634986.1", "A. rna-XR_002634986.1")
p2 <- plot_lnc_go(all_lnc_go, "rna-XR_002638098.1", "B. rna-XR_002638098.1")
p3 <- plot_lnc_go(all_lnc_go, "rna-XR_002638049.1", "C. rna-XR_002638049.1")

combined_lnc_go <- (p1 / p2 / p3) +
  plot_annotation(
    title = "GO enrichment of mRNAs associated with representative hub lncRNAs"
  )

ggsave(
  "~/github/oyster-lnc/output/19-GO-from-coexpression/lncRNA_module_GO_panels.png",
  combined_lnc_go,
  width = 10,
  height = 14,
  dpi = 300
)

combined_lnc_go
```

### G protein receptor and calcium

```{r}
library(tidyverse)

gene2go <- readr::read_tsv(
  "~/github/oyster-lnc/output/19-GO-from-coexpression/gene2go_oyster.tsv",
  show_col_types = FALSE
)

sig_pairs <- readr::read_tsv(
  "~/github/oyster-lnc/output/18-lncRNA-mRNA-pearson/significant_lnc_mrna_pairs.tsv",
  show_col_types = FALSE
)

id_map <- readr::read_tsv(
  "~/github/oyster-lnc/output/19-GO-from-coexpression/mrna_id_to_ncbi_GeneID_map.tsv",
  show_col_types = FALSE
)

# calcium and GPCR GO-linked genes from your oyster annotation
calcium_genes <- gene2go %>%
  dplyr::filter(grepl("calcium", GO_term, ignore.case = TRUE)) %>%
  dplyr::pull(GeneID) %>%
  unique()

gpcr_genes <- gene2go %>%
  dplyr::filter(grepl("G protein|receptor", GO_term, ignore.case = TRUE)) %>%
  dplyr::pull(GeneID) %>%
  unique()

# map sig_pairs mRNAs to GeneID
sig_pairs_annot <- sig_pairs %>%
  dplyr::left_join(
    id_map %>% dplyr::select(mrna_id, GeneID, Symbol, description),
    by = "mrna_id"
  )

# score each lncRNA by how many calcium / GPCR genes it connects to
lnc_target_theme_summary <- sig_pairs_annot %>%
  dplyr::group_by(lnc_id) %>%
  dplyr::summarise(
    n_targets = dplyr::n_distinct(mrna_id),
    n_mapped_targets = dplyr::n_distinct(mrna_id[!is.na(GeneID)]),
    n_calcium_targets = dplyr::n_distinct(mrna_id[GeneID %in% calcium_genes]),
    n_gpcr_targets = dplyr::n_distinct(mrna_id[GeneID %in% gpcr_genes]),
    prop_calcium = n_calcium_targets / pmax(n_mapped_targets, 1),
    prop_gpcr = n_gpcr_targets / pmax(n_mapped_targets, 1),
    .groups = "drop"
  ) %>%
  dplyr::arrange(dplyr::desc(n_calcium_targets), dplyr::desc(n_gpcr_targets))

readr::write_tsv(
  lnc_target_theme_summary,
  file.path("~/github/oyster-lnc/output/19-GO-from-coexpression/lnc_target_theme_summary.tsv")
)

lnc_target_theme_summary
```

### Plot

```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)

calcium_lnc <- "rna-XR_002638067.1"
gpcr_lnc    <- "rna-XR_002638049.1"

# calcium / GPCR gene sets
calcium_genes <- gene2go %>%
  dplyr::filter(grepl("calcium", GO_term, ignore.case = TRUE)) %>%
  dplyr::pull(GeneID) %>%
  unique()

gpcr_genes <- gene2go %>%
  dplyr::filter(grepl("G protein|receptor", GO_term, ignore.case = TRUE)) %>%
  dplyr::pull(GeneID) %>%
  unique()

# annotated pairs
sig_pairs_annot <- sig_pairs %>%
  dplyr::left_join(
    id_map %>% dplyr::select(mrna_id, GeneID, Symbol),
    by = "mrna_id"
  )

# extract targets
calcium_targets <- sig_pairs_annot %>%
  dplyr::filter(lnc_id == calcium_lnc, GeneID %in% calcium_genes)

gpcr_targets <- sig_pairs_annot %>%
  dplyr::filter(lnc_id == gpcr_lnc, GeneID %in% gpcr_genes)

plot_target_bar <- function(df, title) {
  df_plot <- df %>%
    dplyr::mutate(
      label = dplyr::if_else(!is.na(Symbol) & Symbol != "", Symbol, mrna_id)
    ) %>%
    dplyr::arrange(dplyr::desc(abs(r))) %>%
    dplyr::slice_head(n = 12)

  ggplot(df_plot, aes(x = reorder(label, abs(r)), y = abs(r), fill = r > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#E63946", "FALSE" = "#457B9D"), guide = "none") +
    labs(
      title = title,
      x = NULL,
      y = "|Pearson r|"
    ) +
    theme_minimal(base_size = 11)
}

p_calcium <- plot_target_bar(
  calcium_targets,
  paste("A.", calcium_lnc, "calcium-related targets")
)

p_gpcr <- plot_target_bar(
  gpcr_targets,
  paste("B.", gpcr_lnc, "GPCR-related targets")
)

final_panel <- p_calcium / p_gpcr +
  plot_annotation(
    title = "Representative lncRNAs associated with calcium- and GPCR-related gene targets"
  )

ggsave(
  "~/github/oyster-lnc/output/19-GO-from-coexpression/lnc_calcium_gpcr_panels.png",
  final_panel,
  width = 10,
  height = 12,
  dpi = 300
)

final_panel
```

### Network plots

```{r}
library(tidyverse)
library(igraph)
library(patchwork)

out_dir <- "~/github/oyster-lnc/output/19-GO-from-coexpression"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

calcium_lnc <- "rna-XR_002638067.1"
gpcr_lnc    <- "rna-XR_002638049.1"

# rebuild annotated pairs if needed
sig_pairs_annot <- sig_pairs %>%
  dplyr::left_join(
    id_map %>% dplyr::select(mrna_id, GeneID, Symbol, description),
    by = "mrna_id"
  )

# define theme gene sets from GO terms
calcium_genes <- gene2go %>%
  dplyr::filter(grepl("calcium", GO_term, ignore.case = TRUE)) %>%
  dplyr::pull(GeneID) %>%
  unique()

gpcr_genes <- gene2go %>%
  dplyr::filter(grepl("G protein|receptor", GO_term, ignore.case = TRUE)) %>%
  dplyr::pull(GeneID) %>%
  unique()

plot_lnc_subnetwork <- function(sig_df, lnc_name, theme_gene_ids, title_text) {
  sub_edges <- sig_df %>%
    dplyr::filter(lnc_id == lnc_name, GeneID %in% theme_gene_ids) %>%
    dplyr::mutate(
      from = lnc_id,
      to = mrna_id,
      plot_label = dplyr::if_else(
        !is.na(Symbol) & Symbol != "",
        Symbol,
        mrna_id
      )
    ) %>%
    dplyr::arrange(dplyr::desc(abs(r)), FDR) %>%
    dplyr::slice_head(n = 15)

  if (nrow(sub_edges) == 0) {
    stop(paste("No edges found for", lnc_name))
  }

  g_sub <- igraph::graph_from_data_frame(
    sub_edges %>% dplyr::select(from, to, r, FDR),
    directed = FALSE
  )

  V(g_sub)$type <- ifelse(V(g_sub)$name == lnc_name, "lncRNA", "mRNA")

  V(g_sub)$label <- ifelse(V(g_sub)$name == lnc_name, lnc_name, NA)

  V(g_sub)$color <- ifelse(V(g_sub)$type == "lncRNA", "#E63946", "#457B9D")
  V(g_sub)$size  <- ifelse(V(g_sub)$type == "lncRNA", 16, 8)
  V(g_sub)$label.cex <- 0.9
  V(g_sub)$label.color <- "gray20"
  V(g_sub)$frame.color <- NA

  E(g_sub)$color <- ifelse(sub_edges$r > 0, "#E63946AA", "#457B9DAA")
  E(g_sub)$width <- scales::rescale(abs(sub_edges$r), to = c(1, 3))

  set.seed(123)
  lay <- igraph::layout_with_fr(g_sub, niter = 1500)

  plot(
    g_sub,
    layout = lay,
    main = title_text,
    vertex.label.family = "sans",
    edge.curved = 0.1,
    margin = 0.1
  )
}

png(
  file.path(out_dir, "lnc_calcium_gpcr_subnetworks.png"),
  width = 2400, height = 3200, res = 300
)

par(mfrow = c(2, 1), mar = c(1, 1, 3, 1))

plot_lnc_subnetwork(
  sig_df = sig_pairs_annot,
  lnc_name = calcium_lnc,
  theme_gene_ids = calcium_genes,
  title_text = "A. Calcium-related subnetwork"
)

plot_lnc_subnetwork(
  sig_df = sig_pairs_annot,
  lnc_name = gpcr_lnc,
  theme_gene_ids = gpcr_genes,
  title_text = "B. GPCR/signaling-related subnetwork"
)

dev.off()
```

### Dot plots for specific GPCR/calcium correlated lncRNAs

```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)

out_dir <- "~/github/oyster-lnc/output/19-GO-from-coexpression"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

calcium_lnc <- "rna-XR_002638067.1"
gpcr_lnc    <- "rna-XR_002638049.1"

# rebuild annotated pairs if needed
sig_pairs_annot <- sig_pairs %>%
  dplyr::left_join(
    id_map %>% dplyr::select(mrna_id, GeneID, Symbol, description),
    by = "mrna_id"
  )

plot_lnc_theme_dotplot <- function(sig_df, gene2go_df, lnc_name, pattern, panel_title, n_terms = 10) {
  
  # all mapped target genes for this lncRNA
  lnc_targets <- sig_df %>%
    dplyr::filter(lnc_id == lnc_name, !is.na(GeneID)) %>%
    dplyr::distinct(mrna_id, GeneID)

  n_total_targets <- nrow(lnc_targets)

  if (n_total_targets == 0) {
    stop(paste("No mapped targets found for", lnc_name))
  }

  # GO terms matching desired theme
  theme_go <- gene2go_df %>%
    dplyr::filter(grepl(pattern, GO_term, ignore.case = TRUE)) %>%
    dplyr::distinct(GeneID, GO_ID, GO_term, Category)

  # join targets to theme GO terms
  plot_df <- lnc_targets %>%
    dplyr::inner_join(theme_go, by = "GeneID") %>%
    dplyr::group_by(GO_ID, GO_term, Category) %>%
    dplyr::summarise(
      Count = dplyr::n_distinct(mrna_id),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      GeneRatio = Count / n_total_targets
    ) %>%
    dplyr::arrange(dplyr::desc(Count), dplyr::desc(GeneRatio)) %>%
    dplyr::slice_head(n = n_terms)

  if (nrow(plot_df) == 0) {
    stop(paste("No GO terms matched pattern for", lnc_name))
  }

  ggplot(plot_df, aes(x = GeneRatio, y = reorder(GO_term, GeneRatio))) +
    geom_point(aes(size = Count, color = Count)) +
    scale_color_viridis_c() +
    labs(
      title = panel_title,
      subtitle = paste0(lnc_name, " (n = ", n_total_targets, " mapped correlated mRNAs)"),
      x = "GeneRatio among correlated mRNAs",
      y = NULL,
      size = "Count",
      color = "Count"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold"),
      axis.text.y = element_text(size = 9)
    )
}

p_calcium <- plot_lnc_theme_dotplot(
  sig_df = sig_pairs_annot,
  gene2go_df = gene2go,
  lnc_name = calcium_lnc,
  pattern = "calcium",
  panel_title = "A. Calcium-related GO terms"
)

p_gpcr <- plot_lnc_theme_dotplot(
  sig_df = sig_pairs_annot,
  gene2go_df = gene2go,
  lnc_name = gpcr_lnc,
  pattern = "G protein|receptor",
  panel_title = "B. GPCR/signaling-related GO terms"
)

combined_theme_plot <- p_calcium / p_gpcr +
  plot_annotation(
    title = "Representative lncRNAs and the GO categories represented among their correlated mRNAs"
  )

ggsave(
  file.path(out_dir, "lnc_calcium_gpcr_theme_dotplots.png"),
  combined_theme_plot,
  width = 10,
  height = 12,
  dpi = 300
)

combined_theme_plot
```



