---
title: "19-GO-from-coexpression"
author: "Zach Bengtsson"
date: "2026-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

GO enrichment for Eastern oyster coexpression results
when mRNA IDs look like LOC111120067

## ---------------------------
## Packages
## ---------------------------

```{r}
packages <- c(
  "tidyverse",
  "clusterProfiler",
  "enrichplot",
  "GO.db"
)

to_install <- packages[!vapply(packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(to_install) > 0) {
  install.packages(setdiff(to_install, c("clusterProfiler", "enrichplot", "GO.db")))
  if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
  bioc_pkgs <- intersect(to_install, c("clusterProfiler", "enrichplot", "GO.db"))
  if (length(bioc_pkgs) > 0) {
    BiocManager::install(bioc_pkgs, ask = FALSE, update = FALSE)
  }
}

library(tidyverse)
library(clusterProfiler)
library(enrichplot)
library(GO.db)
```

## ---------------------------
## Input / output
## ---------------------------
```{r}
coexp_dir <- "../output/18-lncRNA-mRNA-pearson"
out_dir   <- "../output/19-GO-from-coexpression"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

sig_pairs_file <- file.path(coexp_dir, "significant_lnc_mrna_pairs.tsv")
mrna_file      <- "~/github/oyster-lnc/output/11-mRNA-counts-DESeq2/kallisto_gene_counts.tsv"
```
## ---------------------------
## Load data
## ---------------------------

```{r}
sig_pairs <- readr::read_tsv(sig_pairs_file, show_col_types = FALSE)

mrna_raw <- readr::read_tsv(mrna_file, show_col_types = FALSE)
colnames(mrna_raw)[1] <- "mrna_id"

mrna_universe_ids <- unique(mrna_raw$mrna_id)
```

## ---------------------------
## Hub mRNA summary
## ---------------------------

```{r}
mrna_degree <- sig_pairs %>%
  count(mrna_id, name = "n_lncRNAs") %>%
  arrange(desc(n_lncRNAs), mrna_id)

readr::write_tsv(mrna_degree, file.path(out_dir, "mrna_degree_table.tsv"))

n_hub <- max(10, ceiling(0.10 * nrow(mrna_degree)))

hub_mrnas <- mrna_degree %>%
  slice_head(n = n_hub)

readr::write_tsv(hub_mrnas, file.path(out_dir, "hub_mrnas_top10pct.tsv"))
```

## ---------------------------
## Download NCBI files
## ---------------------------

```{r}
tax_id <- 6565

gene2go_url   <- "https://ftp.ncbi.nlm.nih.gov/gene/DATA/gene2go.gz"
gene_info_url <- "https://ftp.ncbi.nlm.nih.gov/gene/DATA/GENE_INFO/All_Data.gene_info.gz"

gene2go_oyster_file   <- file.path(out_dir, "gene2go_oyster.tsv")
gene_info_oyster_file <- file.path(out_dir, "gene_info_oyster.tsv")

if (!file.exists(gene2go_oyster_file)) {
  gene2go <- data.table::fread(
    cmd = paste(
      "curl -L", shQuote(gene2go_url), "| gzip -dc"
    ),
    sep = "\t",
    header = FALSE,
    skip = 1,
    quote = "",
    fill = TRUE,
    showProgress = TRUE,
    col.names = c(
      "tax_id", "GeneID", "GO_ID", "Evidence", "Qualifier",
      "GO_term", "PubMed", "Category"
    )
  ) |>
    tibble::as_tibble() |>
    dplyr::filter(tax_id == !!tax_id) |>
    dplyr::mutate(GeneID = as.character(GeneID)) |>
    dplyr::distinct(GeneID, GO_ID, GO_term, Category, .keep_all = TRUE)

  readr::write_tsv(gene2go, gene2go_oyster_file)
}

if (!file.exists(gene_info_oyster_file)) {
  gene_info <- data.table::fread(
    cmd = paste(
      "curl -L", shQuote(gene_info_url), "| gzip -dc"
    ),
    sep = "\t",
    header = FALSE,
    skip = 1,
    quote = "",
    fill = TRUE,
    showProgress = TRUE,
    col.names = c(
      "tax_id", "GeneID", "Symbol", "LocusTag", "Synonyms",
      "dbXrefs", "chromosome", "map_location", "description",
      "type_of_gene", "Symbol_from_nomenclature_authority",
      "Full_name_from_nomenclature_authority",
      "Nomenclature_status", "Other_designations",
      "Modification_date", "Feature_type"
    )
  ) |>
    tibble::as_tibble() |>
    dplyr::filter(tax_id == !!tax_id) |>
    dplyr::mutate(
      GeneID = as.character(GeneID),
      Symbol = dplyr::na_if(Symbol, "-"),
      LocusTag = dplyr::na_if(LocusTag, "-"),
      Synonyms = dplyr::na_if(Synonyms, "-")
    )

  readr::write_tsv(gene_info, gene_info_oyster_file)
}

cat("Cached oyster annotation files:\n")
cat(" -", gene2go_oyster_file, "\n")
cat(" -", gene_info_oyster_file, "\n")
```
## ---------------------------
## Cache Building
## ---------------------------

```{r}
tax_id <- 6565
gene2go_url <- "https://ftp.ncbi.nlm.nih.gov/gene/DATA/gene2go.gz"
gene_info_url <- "https://ftp.ncbi.nlm.nih.gov/gene/DATA/GENE_INFO/All_Data.gene_info.gz"

gene2go_oyster_file   <- file.path(out_dir, "gene2go_oyster.tsv")
gene_info_oyster_file <- file.path(out_dir, "gene_info_oyster.tsv")
```


## ---------------------------
## Read gene2go
## ---------------------------

```{r}
gene2go <- readr::read_tsv(
  gene2go_oyster_file,
  show_col_types = FALSE,
  col_types = readr::cols(.default = readr::col_character())
) %>%
  mutate(GeneID = as.character(GeneID)) %>%
  distinct(GeneID, GO_ID, GO_term, Category, .keep_all = TRUE)
```

## ---------------------------
## Read gene_info
## ---------------------------

```{r}
gene_info <- readr::read_tsv(
  gene_info_oyster_file,
  show_col_types = FALSE,
  col_types = readr::cols(.default = readr::col_character())
) %>%
  mutate(
    GeneID   = as.character(GeneID),
    Symbol   = na_if(Symbol, "-"),
    LocusTag = na_if(LocusTag, "-"),
    Synonyms = na_if(Synonyms, "-")
  )
```

## ---------------------------
## Clean IDs
## ---------------------------

```{r}
clean_id <- function(x) {
  x %>%
    as.character() %>%
    stringr::str_trim() %>%
    stringr::str_remove("\\.[0-9]+$")
}

all_ids_tbl <- tibble(
  mrna_id = unique(c(mrna_universe_ids, sig_pairs$mrna_id))
) %>%
  mutate(mrna_id_clean = clean_id(mrna_id))
```

## ---------------------------
## Primary mapping:
## LOC identifiers should match NCBI gene_info$Symbol directly
## ---------------------------

```{r}
id_map_symbol <- all_ids_tbl %>%
  left_join(
    gene_info %>%
      transmute(
        GeneID,
        Symbol,
        description,
        type_of_gene,
        match_value = Symbol
      ),
    by = c("mrna_id_clean" = "match_value")
  ) %>%
  mutate(match_type = ifelse(!is.na(GeneID), "Symbol", NA_character_))
```

## ---------------------------
## Secondary fallback:
## if an mrna_id is actually already a GeneID number
## ---------------------------

```{r}
id_map_geneid <- all_ids_tbl %>%
  left_join(
    gene_info %>%
      transmute(
        GeneID,
        Symbol,
        description,
        type_of_gene,
        match_value = GeneID
      ),
    by = c("mrna_id_clean" = "match_value")
  ) %>%
  mutate(match_type = ifelse(!is.na(GeneID), "GeneID", NA_character_))
```

## ---------------------------
## Combine mappings, prefer Symbol match first
## ---------------------------

```{r}
id_map <- id_map_symbol %>%
  bind_rows(id_map_geneid) %>%
  group_by(mrna_id) %>%
  arrange(factor(match_type, levels = c("Symbol", "GeneID"))) %>%
  slice_head(n = 1) %>%
  ungroup()

readr::write_tsv(id_map, file.path(out_dir, "mrna_id_to_ncbi_GeneID_map.tsv"))

mapping_summary <- tibble(
  n_unique_mrna_ids = n_distinct(all_ids_tbl$mrna_id),
  n_mapped          = sum(!is.na(id_map$GeneID)),
  prop_mapped       = mean(!is.na(id_map$GeneID))
)

print(mapping_summary)
readr::write_tsv(mapping_summary, file.path(out_dir, "mapping_summary.tsv"))
```

## ---------------------------
## Annotate network tables
## ---------------------------

```{r}
sig_pairs_annot <- sig_pairs %>%
  left_join(
    id_map %>% dplyr::select(mrna_id, GeneID, Symbol, description, type_of_gene, match_type),
    by = "mrna_id"
  )

mrna_degree_annot <- mrna_degree %>%
  left_join(
    id_map %>% dplyr::select(mrna_id, GeneID, Symbol, description, type_of_gene, match_type),
    by = "mrna_id"
  )

readr::write_tsv(sig_pairs_annot, file.path(out_dir, "significant_pairs_with_geneIDs.tsv"))
readr::write_tsv(mrna_degree_annot, file.path(out_dir, "mrna_degree_with_geneIDs.tsv"))
```

## ---------------------------
## TERM2GENE / TERM2NAME
## ---------------------------

```{r}
term2gene <- gene2go %>%
  transmute(term = GO_ID, gene = GeneID) %>%
  distinct()

term2name <- gene2go %>%
  transmute(term = GO_ID, name = GO_term) %>%
  distinct()
```

## ---------------------------
## Recreate filtered mRNA universe (mrna_variable)
## ---------------------------

```{r}
# Load mRNA matrix
mrna_raw <- readr::read_tsv(mrna_file, show_col_types = FALSE) %>%
  dplyr::rename(mrna_id = gene_id)

mrna_mat <- mrna_raw %>%
  tibble::column_to_rownames("mrna_id") %>%
  as.matrix()
```

## ---------------------------
## Filter low-expression genes
## ---------------------------

```{r}
min_count   <- 10
min_samples <- 26

filter_rows <- function(mat, min_count, min_samples) {
  keep <- rowSums(mat >= min_count) >= min_samples
  mat[keep, , drop = FALSE]
}

mrna_mat_f <- filter_rows(mrna_mat, min_count, min_samples)

cat("After expression filter:", nrow(mrna_mat_f), "genes\n")
```

## ---------------------------
## Filter low-variability genes (CV filter)
## ---------------------------

```{r}
cv_threshold <- 0.1

compute_cv <- function(x) {
  sd(x) / (mean(x) + 1e-9)
}

mrna_cv <- apply(mrna_mat_f, 1, compute_cv)

mrna_variable <- mrna_mat_f[mrna_cv > cv_threshold, , drop = FALSE]

cat("After CV filter:", nrow(mrna_variable), "genes\n")
```

## ---------------------------
## Define correct background universe
## ---------------------------

```{r}
mrna_universe_ids <- rownames(mrna_variable)

cat("Final universe size:", length(mrna_universe_ids), "\n")
```

## ---------------------------
## Background universe
## ---------------------------

```{r}
gene_universe <- id_map %>%
  dplyr::filter(mrna_id %in% mrna_universe_ids, !is.na(GeneID)) %>%
  dplyr::pull(GeneID) %>%
  unique()
```

## ---------------------------
## GO enrichment: hub mRNAs
## ---------------------------

```{r}
hub_gene_ids <- hub_mrnas %>%
  left_join(id_map %>% dplyr::select(mrna_id, GeneID), by = "mrna_id") %>%
  dplyr::filter(!is.na(GeneID)) %>%
  dplyr::pull(GeneID) %>%
  unique()

ego_hubs <- enricher(
  gene          = hub_gene_ids,
  universe      = gene_universe,
  TERM2GENE     = term2gene,
  TERM2NAME     = term2name,
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  minGSSize     = 10,
  maxGSSize     = 500
)

if (!is.null(ego_hubs) && nrow(as.data.frame(ego_hubs)) > 0) {
  ego_hubs_df <- as.data.frame(ego_hubs) %>%
    dplyr::mutate(
      ONTOLOGY = purrr::map_chr(ID, ~{
        x <- gene2go %>%
          dplyr::filter(GO_ID == .x) %>%
          dplyr::slice_head(n = 1) %>%
          dplyr::pull(Category)
        if (length(x) == 0) NA_character_ else x
      })
    )

  readr::write_tsv(ego_hubs_df, file.path(out_dir, "GO_enrichment_hub_mrnas.tsv"))

  p_dot <- dotplot(ego_hubs, showCategory = 20) +
    ggplot2::ggtitle("GO enrichment of hub mRNAs")

  ggplot2::ggsave(
    file.path(out_dir, "GO_dotplot_hub_mrnas.png"),
    p_dot, width = 10, height = 7, dpi = 300
  )
}
```

```{r}
as.data.frame(ego_hubs) %>% head()
```


## ---------------------------
## GO enrichment: top hub lncRNA target sets
## ---------------------------

```{r}
top_lnc <- sig_pairs %>%
  dplyr::count(lnc_id, name = "n_targets") %>%
  dplyr::arrange(desc(n_targets)) %>%
  dplyr::slice_head(n = 5) %>%
  dplyr::pull(lnc_id)

lnc_go_results <- list()

for (lnc in top_lnc) {
  target_gene_ids <- sig_pairs %>%
    dplyr::filter(lnc_id == lnc) %>%
    left_join(id_map %>% dplyr::select(mrna_id, GeneID), by = "mrna_id") %>%
    dplyr::filter(!is.na(GeneID)) %>%
    dplyr::pull(GeneID) %>%
    unique()

  if (length(target_gene_ids) < 10) next

  ego_lnc <- enricher(
    gene          = target_gene_ids,
    universe      = gene_universe,
    TERM2GENE     = term2gene,
    TERM2NAME     = term2name,
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.05,
    minGSSize     = 10,
    maxGSSize     = 500
  )

  if (!is.null(ego_lnc) && nrow(as.data.frame(ego_lnc)) > 0) {
    ego_lnc_df <- as.data.frame(ego_lnc) %>%
      dplyr::mutate(lnc_id = lnc)

    lnc_go_results[[lnc]] <- ego_lnc_df

    safe_name <- gsub("[^A-Za-z0-9_.-]", "_", lnc)
    p <- dotplot(ego_lnc, showCategory = 12) +
      ggplot2::ggtitle(paste("GO enrichment for targets of", lnc))

    ggplot2::ggsave(
      file.path(out_dir, paste0("GO_dotplot_", safe_name, ".png")),
      p, width = 10, height = 6, dpi = 300
    )
  }
}

if (length(lnc_go_results) > 0) {
  all_lnc_go <- dplyr::bind_rows(lnc_go_results)
  readr::write_tsv(all_lnc_go, file.path(out_dir, "GO_enrichment_top_lnc_modules.tsv"))
}
```

## ---------------------------
## Convenient annotation table for manuscript use
## ---------------------------

```{r}
top30_hub_mrna_annotation <- mrna_degree %>%
  dplyr::slice_head(n = 30) %>%
  left_join(
    id_map %>% dplyr::select(mrna_id, GeneID, Symbol, description, type_of_gene),
    by = "mrna_id"
  ) %>%
  left_join(
    gene2go %>%
      dplyr::group_by(GeneID) %>%
      dplyr::summarise(
        n_go_terms = dplyr::n_distinct(GO_ID),
        example_go_terms = paste(head(unique(GO_term), 5), collapse = "; "),
        .groups = "drop"
      ),
    by = "GeneID"
  )

readr::write_tsv(
  top30_hub_mrna_annotation,
  file.path(out_dir, "top30_hub_mrna_annotations.tsv")
)

cat("Done.\n")
cat("Mapped mRNA IDs:", sum(!is.na(id_map$GeneID)), "of", dplyr::n_distinct(id_map$mrna_id), "\n")
cat("Output written to:", out_dir, "\n")
```